---
title: "Group-class and cross-level models: classes of groups from raw ratings"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Group-class and cross-level models: classes of groups from raw ratings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5,
                      fig.height = 4.4, fig.align = "center", dpi = 96)
options(digits = 3, width = 90)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

Members of the same team, pupils of the same class or sessions of the same
student give ratings that share something: the unit they belong to. The
*additive* model of Houle, Morin and Harvey (2026) asks whether those units
fall into a small number of latent **group classes** that differ in where
their members' ratings sit. It has no individual profiles: every member of
a group belongs to the group's class, and varies around the group's own level.

The same machinery fits the other families of that framework: the
*dispersion* family, whose group classes differ in how much members vary
around their group, the *additive-dispersion* family, which allows both, and
two *cross-level* families that estimate individual profiles and group classes
together.

> These families are **experimental**. Estimation, standard errors, tables,
> plots, enumeration and the bootstrap likelihood-ratio test are available;
> missing data and covariates are not yet.

## The model

For group $j$ with $n_j$ members rating $d$ indicators, and group class $h$:

$$
C_j \sim \text{Categorical}(\omega_1, \dots, \omega_H), \qquad
B_{jr} \mid C_j = h \sim N(\mu_{hr}, T_{hr}), \qquad
Y_{ijr} \mid B_{jr} \sim N(B_{jr}, W_r).
$$

Each group has its own intercept $B_{jr}$ per indicator, drawn from its
class's distribution. The group classes differ in their means $\mu_h$ and,
optionally, in the between-group spread $T_h$; the **within-group** variance
$W_r$ is the same in every class. That shared within-group variance is what
makes the model additive. The likelihood is exact: it is computed from each
group's size, mean and within-group scatter, with no numerical integration.

## Data

The ratings below are simulated so that the truth is known: 120 teams of 4 to
12 members rate the team's climate and the support they receive. Teams are
of two types, supportive and strained, which differ in their average ratings
and, for climate, in how much teams of the type vary.

```{r data}
library(latents)
set.seed(2026)
n_teams <- 120
team_type <- rep(c("supportive", "strained"), times = c(66, 54))
team_size <- rep(c(4, 6, 8, 12), length.out = n_teams)
team_climate <- rnorm(n_teams, ifelse(team_type == "supportive", 0.8, -0.8),
                      ifelse(team_type == "supportive", 0.35, 0.6))
team_support <- rnorm(n_teams, ifelse(team_type == "supportive", 0.6, -0.5), 0.45)
ratings <- data.frame(
  team = rep(seq_len(n_teams), team_size),
  type = rep(team_type, team_size),
  climate = rep(team_climate, team_size) + rnorm(sum(team_size), 0, 0.9),
  support = rep(team_support, team_size) + rnorm(sum(team_size), 0, 1))
str(ratings)
```

The model never sees `type`; it is kept to check the classification later.

## Fitting

The additive model is a `family` of `multilpa()`. It takes the ratings, the
group column and the number of group classes; there is no `n_profiles`,
because there are no individual profiles.

```{r fit}
fit <- multilpa(ratings, c("climate", "support"), "team",
                n_group_classes = 2, family = "additive", seed = 1)
fit
```

## How many group classes?

Fit one, two and three classes and compare them by BIC, which for this family
is penalized by the number of groups (`nobs(fit)` is the number of teams).

```{r classes}
fit_one <- multilpa(ratings, c("climate", "support"), "team",
                    n_group_classes = 1, family = "additive", seed = 1)
fit_three <- multilpa(ratings, c("climate", "support"), "team",
                      n_group_classes = 3, family = "additive", seed = 1)
BIC(fit_one, fit, fit_three)
```

Two classes have the lowest BIC. The three-class fit also warns that a
between-group variance went to zero: with a third class there is no
between-team spread left for some class to explain. A likelihood-ratio test of
two against three classes would need a bootstrap reference distribution,
not the chi-square, and that is not yet available for this family.

## Do the classes differ in spread?

`between_variance = "equal"` constrains the between-group variances to be
equal across classes, which leaves the classes differing in location only.

```{r restriction}
fit_equal <- multilpa(ratings, c("climate", "support"), "team",
                      n_group_classes = 2, family = "additive",
                      between_variance = "equal", seed = 1)
BIC(fit, fit_equal)
AIC(fit, fit_equal)
```

Both criteria prefer the equal-spread model, although the data were simulated
with more spread in climate among strained teams. With 120 teams that
difference is too small to be told apart from sampling noise, which is what
the criteria report. The rest of this vignette uses the varying model to show
every parameter.

## Comparing the group-class families

`enumerate_classes()` fits every combination of family, between-variance
restriction and number of group classes in one call. The `model` column names
each candidate for `candidate_fit()`.

```{r enumerate}
grid <- enumerate_classes(ratings, c("climate", "support"), "team",
                          family = c("additive", "dispersion",
                                     "additive_dispersion"),
                          n_group_classes = 1:3, seed = 1)
get_results(grid, "best")
```

The lowest BIC belongs to two additive classes with equal between-group
spread, the model chosen above. Adding class-specific within-group variances
(additive-dispersion) does not improve the fit enough to pay for them, and the
dispersion family, whose classes differ only in spread, fits much worse: these
teams differ in level, not in how much their members disagree.

```{r plot-enumerate, eval = has_ggplot}
plot(grid)
```

Whether two classes are needed at all is a test of one class against two. The
likelihood-ratio statistic has no chi-square reference distribution there, so
`bootstrap_lrt()` simulates it from the one-class fit:

```{r bootstrap}
one_class <- candidate_fit(grid, n_group_classes = 1, model = "additive_equal")
two_classes <- candidate_fit(grid, n_group_classes = 2, model = "additive_equal")
bootstrap_lrt(one_class, two_classes, iter = 19, n_starts = 3, seed = 1)
```

With 19 replicates the smallest attainable p-value is 0.05; use several
hundred replicates for a report.

## Parameters

Every parameter, with Wald standard errors from the observed information:

```{r parameters}
get_results(fit)
```

The warning comes from a check on each class's *effective groups*: the
number of perfectly classified groups that would pin down the class weight
as precisely as this fit does. It is shown in the `group_classes` table below.
Here the strained class has about 38 effective groups of the 49 it is expected
to hold, so most of the information survives the classification; the class is
simply not large. With fewer than 50 effective groups, the simulation study
found interval coverage for class parameters starting to fall below 95%, and
far below it when a class is also poorly separated.

`level` separates the three kinds of parameter: `"between"` rows are the
class means and between-group variances, `"within"` rows the within-group
variances shared by the classes (`group_class = "shared"`), and
`"group_class"` rows the class weights. Intervals for variances are formed on
the log scale and for weights on the logit scale, so they stay in range.
Cluster-robust (sandwich) standard errors over teams take one argument:

```{r robust}
get_results(fit, "parameters", vcov_type = "robust")
```

## Groups and their intercepts

Each team's class probabilities, and its modal class:

```{r groups}
head(get_results(fit, "groups"))
get_results(fit, "group_classes")
```

The posterior mean of each team's intercept shrinks its observed average
towards its class mean, more so for small teams; its variance includes the
uncertainty about which class the team belongs to.

```{r intercepts}
head(get_results(fit, "intercepts"))
```

Because the teams were simulated, the classification can be checked against
the true type:

```{r recovery}
get_results(fit, "recovery", data = ratings, truth = "type")
```

Every supportive team is in one class; most strained teams are in the other,
and the few placed with the supportive teams are those whose ratings happened
to fall close to that class.

```{r check, include = FALSE}
# Guards on the prose above: it must stay true of the rendered numbers.
bic_classes <- BIC(fit_one, fit, fit_three)
stopifnot(
  which.min(bic_classes$BIC) == 2L,
  BIC(fit_equal) < BIC(fit), AIC(fit_equal) < AIC(fit),
  any(fit_three$between_zero)
)
classes <- get_results(fit, "group_classes")
stopifnot(min(classes$effective_groups) < 50,
          min(classes$effective_groups) > 30,
          min(classes$effective_groups / classes$count) > 0.7)
recovery <- get_results(fit, "recovery", data = ratings, truth = "type")
agreement <- with(recovery, sum(tapply(n, assigned, max)) / sum(n))
stopifnot(agreement >= 0.9,
          with(subset(recovery, truth == "supportive"), max(share) > 0.9))
```

## Plots

```{r plot-means, eval = has_ggplot}
plot(fit)
```

```{r plot-variances, eval = has_ggplot}
plot(fit, what = "variances")
```

```{r plot-intercepts, eval = has_ggplot}
plot(fit, what = "intercepts")
```

## Cross-level families

When the same ratings define both individual profiles and group classes, the
cross-level families follow the manifest specification of Houle et al.
(2026): each group's means of the indicators are the group-level indicators
of its class. In the *full* family, the group classes also differ in how
their members are distributed over the individual profiles; in the
*restricted* family they do not, and the two levels are estimated separately.
Here, sessions of `course_engagement` form engagement profiles and students
form group classes.

```{r cross-level}
activity <- c("browse", "lectures", "forum_read", "forum_post", "attendance")
full <- multilpa(course_engagement, activity, "student", n_profiles = 2,
                 n_group_classes = 2, family = "full_cross_level", seed = 1)
restricted <- multilpa(course_engagement, activity, "student", n_profiles = 2,
                       n_group_classes = 2, family = "restricted_cross_level",
                       seed = 1)
get_results(full, "composition")
rbind(get_results(full, "fit"), get_results(restricted, "fit"))
```

In the full model one class of students has most of its sessions in one
profile and the other class mostly in the other, and letting the composition
differ by class raises the likelihood far more than its one extra parameter
costs. Because the group means are computed from the same ratings, these
likelihoods compare cross-level fits of the same data only, not the
group-class families above, and standard errors are not reported.

```{r plot-composition, eval = has_ggplot}
plot(full, what = "composition")
```

```{r check-families, include = FALSE}
best <- get_results(grid, "best")
stopifnot(best$model == "additive_equal", best$n_group_classes == 2,
          with(subset(get_results(grid), n_group_classes == 2),
               bic[model == "dispersion"] > bic[model == "additive_equal"] + 50))
comparison <- rbind(get_results(full, "fit"), get_results(restricted, "fit"))
composition <- get_results(full, "composition")
stopifnot(comparison$log_likelihood[1] > comparison$log_likelihood[2] + 100,
          comparison$bic[1] < comparison$bic[2],
          max(composition$share) > 0.75)
```

## When the between-group variance is zero

In `course_engagement` the student types are discrete by construction, so
students of the same class do not differ in their average activity: the
between-group variance is zero. The fit says so, and the tables report no
standard errors, because a variance on the edge of its space breaks the
approximation Wald inference relies on.

```{r boundary}
activity <- c("browse", "lectures", "forum_read", "forum_post", "attendance")
fit_boundary <- multilpa(course_engagement, activity, "student",
                         n_group_classes = 3, family = "additive",
                         between_variance = "equal", seed = 1)
get_results(fit_boundary, "fit")
```

## Known limitations

A predeclared simulation study (1000 datasets per condition; the report is in
the package repository under `validation/ADDITIVE_SIMULATION.md`) found:

* With a few hundred groups and well-separated classes, means, weights and
  variances are recovered and 95% intervals cover close to 95%.
* Between-group variances are estimated slightly low, the usual small-sample
  bias of maximum likelihood for a variance component; it shrinks as the
  number of groups grows.
* With heavy-tailed ratings, the observed-information standard errors of the
  within-group variances are too small; `vcov_type = "robust"` corrects them.
  With few groups (about 50), prefer the observed-information standard errors.
* When one class is small and poorly separated from another, intervals can
  be badly miscalibrated. The `latents_weak_class` warning (fewer than 50
  effective groups in a class) flagged every such simulated dataset and under
  1% of the well-separated ones with about 100 groups per class; it also
  flags designs with fewer than about 50 groups per class, where coverage
  was mildly low.
* When a between-group variance may truly be zero, its Wald interval is not
  meaningful: either the fit is on the boundary and standard errors are
  withheld, or the interval excludes zero by construction.

## References

Houle, S. A., Morin, A. J. S., & Harvey, J.-F. (2026). Multilevel latent
profile analyses: A comprehensive guide. *Organizational Research Methods*.
doi:10.1177/10944281261469432.
