---
title: "Covariates, outcomes and staged estimation"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariates, outcomes and staged estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", dpi = 96)
options(digits = 3, width = 100)
```

Variables outside the measurement model can be related to the latent classes
in two ways. The three-step approach fits the profiles first and then relates
them to an external variable, correcting for classification error, so the
profiles stay fixed. The one-step approach adds the variable to the model, so
it can also change the profiles. Staged estimation applies the same idea of
fixed profiles to the group classes.

## Data

`course_engagement` has one row per course enrolment: 1,422 enrolments from
106 students. `student` identifies the group, and five activity measures
(`browse`, `lectures`, `forum_read`, `forum_post`, `attendance`) are the
indicators. `previous_grade` is the standardized grade from the course before
each enrolment; it is not an indicator, so it can serve as an external
variable.

```{r setup}
library(latents)
```

## The measurement model

To fit the profiles without the external variable, we call `multilpa()` with
`vars` for the indicators, `id` for the group, `n_profiles`,
`n_group_classes`, `n_starts` and `seed`.

```{r fit}
fit <- multilpa(course_engagement,
                vars = c("browse", "lectures", "forum_read", "forum_post",
                         "attendance"),
                id = "student", n_profiles = 2, n_group_classes = 2,
                n_starts = 10, seed = 1)
fit
```

## Classification error

To obtain the classification-error matrix, we call `get_results()` with
`"classification_errors"` and `level` for the observations or the groups.
Each row gives the probability of each assigned class for cases in a true
class; the three-step methods use it to correct for misclassification.

```{r errors}
get_results(fit, "classification_errors", level = "individuals")
```

## Class means of an external variable

To estimate the mean of an external variable in each profile, we call
`three_step()` with `data` for the data frame and `outcome` for the variable.
The default method, BCH, corrects for classification error; standard errors
are clustered on groups.

```{r three-step}
three_step(fit, data = course_engagement, outcome = "previous_grade")
```

To compare the profiles, we call `three_step()` with `contrast = "pairs"`,
which returns each difference between profile means with a Benjamini-Hochberg
adjusted p-value.

```{r contrasts}
three_step(fit, data = course_engagement, outcome = "previous_grade",
           contrast = "pairs")
```

To see the effect of ignoring classification error, we call `three_step()`
with `method = "modal"`, which treats each case's most probable profile as
known.

```{r modal}
three_step(fit, data = course_engagement, outcome = "previous_grade",
           method = "modal")
```

## Predicting class membership

To regress profile membership on a covariate, we call `r3step()` with `data`
for the data frame and `covariates` for the predictors. It returns
multinomial logit coefficients, corrected for classification error.
`vcov_type = "robust"` clusters the standard errors on groups.

```{r r3step}
r3step(fit, data = course_engagement, covariates = "previous_grade",
       vcov_type = "robust")
```

## One-step estimation

To estimate the covariate together with the profiles, we call `multilpa()`
with `profile_covariates` for predictors of profile membership.
`group_covariates` does the same for group-class membership and needs a
variable that is constant within groups. Covariate models require complete
data.

```{r one-step}
one_step <- multilpa(course_engagement,
                     vars = c("browse", "lectures", "forum_read",
                              "forum_post", "attendance"),
                     id = "student", n_profiles = 2, n_group_classes = 2,
                     profile_covariates = "previous_grade", n_starts = 10,
                     seed = 1)
one_step
```

To obtain the membership coefficients, we call `get_results()` with
`"coefficients"`, and for their standard errors and intervals,
`parameter_inference()`.

```{r coefficients}
get_results(one_step, "coefficients")
parameter_inference(one_step)
```

## Staged estimation

To hold the profiles fixed while the group classes are estimated, we call
`fit_staged()` with the same arguments as `multilpa()`. It first fits the
profiles with one group class, then estimates the group classes with the
profile means and variances held fixed.

```{r staged}
staged <- fit_staged(course_engagement,
                     vars = c("browse", "lectures", "forum_read",
                              "forum_post", "attendance"),
                     id = "student", n_profiles = 2, n_group_classes = 2,
                     n_starts = 10, seed = 1)
staged
```

To see what each stage estimated, we call `get_results()` with `"stages"`.
`parameters` counts the parameters estimated at a stage, and
`parameters_with_measurement` adds those held fixed.

```{r stages}
get_results(staged, "stages")
```

## Reference

The calls below list other functions for external variables and fixed
parameters. Each comment states what the call returns.

```{r reference, eval = FALSE}
get_results(fit, "classification_errors", level = "groups")  # group-level errors
get_results(fit, "bch_weights", level = "individuals")       # BCH weights per case
three_step(fit, data = course_engagement, outcome = "previous_grade",
           method = "proportional")        # weighting by posterior probabilities
three_step(fit, data = course_engagement, outcome = "previous_grade",
           vcov_type = "independent")      # standard errors without clustering
r3step(fit, data = course_engagement,
       covariates = "previous_grade")      # observed-information standard errors
measurement <- multilpa(course_engagement,
                        vars = c("browse", "lectures", "forum_read",
                                 "forum_post", "attendance"),
                        id = "student", n_profiles = 2,
                        n_group_classes = 1, seed = 1)  # a one-class measurement fit
fit_staged(course_engagement,
           vars = c("browse", "lectures", "forum_read", "forum_post",
                    "attendance"),
           id = "student", n_profiles = 2, n_group_classes = 2,
           measurement = measurement)      # stage two from that fit
multilpa(course_engagement,
         vars = c("browse", "lectures", "forum_read", "forum_post",
                  "attendance"),
         id = "student", n_profiles = 2, n_group_classes = 2,
         start = starting_values(fit, what = "measurement"),
         fixed = "variances")              # hold chosen parameters fixed
parameter_inference(staged)                # standard errors of the estimated stage
```

## Assumptions

- Three-step methods assume classification errors are independent of the
  external variable given the true class, and treat the error matrix as known.
- One-step coefficients are conditional on the group class and can change the
  profiles, so they need not equal the three-step coefficients.
- Staged estimation does not carry first-stage uncertainty into the second
  stage, and its likelihood is not the joint maximum.
- All associations are observational.
