---
title: "Evaluating a multilevel latent profile model"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluating a multilevel latent profile model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", dpi = 96)
options(digits = 3, width = 100)
```

A fitted mixture model is evaluated on four questions: whether the estimation
found a stable solution, how clearly observations and groups are classified,
whether the indicators are independent within profiles as the model assumes,
and whether a different number of profiles or group classes fits better.

## Data

`course_engagement` has one row per course enrolment: 1,422 enrolments from
106 students. `student` identifies the group, and five activity measures
(`browse`, `lectures`, `forum_read`, `forum_post`, `attendance`) are the
indicators, each standardized within course.

```{r setup}
library(latents)
```

## The model to evaluate

To fit the model, we call `multilpa()` with `vars` for the indicators, `id`
for the group, `n_profiles` for the number of profiles, `n_group_classes` for
the number of group classes, `n_starts` for the number of random EM starts, and
`seed` for reproducible starts.

```{r fit}
fit <- multilpa(course_engagement,
                vars = c("browse", "lectures", "forum_read", "forum_post",
                         "attendance"),
                id = "student", n_profiles = 2, n_group_classes = 2,
                n_starts = 3, seed = 1)
```

## Estimation

To check convergence, we call `get_results()` with `"model"`. `converged`
reports convergence of the best start, `boundary` whether a variance reached
its lower bound, and `n_best_replicated` how many starts reached the best log
likelihood.

```{r model}
get_results(fit, "model")
```

To check the individual starts, we call `get_results()` with `"starts"`.

```{r starts}
get_results(fit, "starts")
```

To check whether other random seeds reach the same solution, we call
`sensitivity()` with `seeds` for the seeds to try. `optimum` numbers the
distinct maxima found, and `agreement` is the share of observations assigned
as in the original fit, after matching the arbitrary profile labels.

```{r sensitivity}
sensitivity(fit, seeds = 1:3)
```

## Classification

To obtain relative entropy, we call `get_results()` with `"entropy"`. It is
reported for observations and for groups: 1 means every case is assigned with
certainty, 0 means no separation.

```{r entropy}
get_results(fit, "entropy")
```

To obtain classification quality by class, we call `get_results()` with
`"classification"`. `n_modal` is the number of cases assigned to the class,
`average_posterior` the mean posterior probability of those cases, and
`odds_correct_classification` compares that certainty with the class's size.

```{r classification}
get_results(fit, "classification")
```

To see the uncertainty, we call `plot()` with `what = "entropy"` for the
per-case uncertainty within each profile, and with `what = "avepp"` for the
average posterior of each assigned class across all classes.

```{r entropy-plot}
plot(fit, what = "entropy")
```

```{r avepp-plot}
plot(fit, what = "avepp")
```

## Local independence

The default model assumes the indicators are independent within a profile. To
test this, we call `get_results()` with `"residuals"`. `residual` is the
correlation between two indicators that the profiles leave unexplained;
`by = "overall"` pools the profiles. The p-values are approximate.

```{r residuals}
get_results(fit, "residuals", by = "overall")
```

To relax the assumption, we call `multilpa()` with `covariance_model = "full"`,
which estimates the covariances between indicators within each profile. To
compare the two models, we call `get_results()` with
`"information_criteria"` on each; lower values are better.

```{r full-covariance}
full <- multilpa(course_engagement,
                 vars = c("browse", "lectures", "forum_read", "forum_post",
                          "attendance"),
                 id = "student", n_profiles = 2, n_group_classes = 2,
                 covariance_model = "full", n_starts = 3, seed = 1)
get_results(fit, "information_criteria")
get_results(full, "information_criteria")
get_results(full, "residuals", by = "overall")
```

## Number of profiles and group classes

To compare numbers of profiles and group classes, we call
`enumerate_classes()` with `n_profiles` and `n_group_classes` for the counts to
cross. It fits every combination and does not choose a model.

```{r enumerate}
candidates <- enumerate_classes(course_engagement,
                                vars = c("browse", "lectures", "forum_read",
                                         "forum_post", "attendance"),
                                id = "student", n_profiles = 2:4,
                                n_group_classes = 1:3, n_starts = 3, seed = 1)
as.data.frame(candidates)
```

To find which candidate minimizes each criterion, we call `get_results()` with
`"criteria"`. Criteria that depend on a sample size are reported with the
number of groups and with the number of observations.

```{r criteria}
get_results(candidates, "criteria")
```

To draw one criterion across the grid, we call `plot()` with `criterion` for
its name.

```{r enumeration-plot}
plot(candidates, criterion = "bic_groups")
```

To take one candidate out of the grid as a fitted model, we call
`candidate_fit()` with its `n_profiles` and `n_group_classes`.

```{r candidate}
candidate_fit(candidates, n_profiles = 3, n_group_classes = 2)
```

## Parameter uncertainty

To obtain standard errors and 95% Wald intervals, we call
`parameter_inference()` on the fit. With `vcov_type = "robust"`, the standard
errors are clustered on groups.

```{r inference}
parameter_inference(fit)
parameter_inference(fit, vcov_type = "robust")
```

## Reference

The calls below list other evaluation functions and views. Each comment states
what the call returns.

```{r reference, eval = FALSE}
get_results(fit, "information_criteria", format = "long")  # one row per criterion
get_results(fit, "average_posteriors")          # mean posterior by assigned class
get_results(fit, "classification_errors")       # P(assigned class | true class)
get_results(fit, "residuals")                   # residuals within each profile
get_results(candidates, "candidates")           # the full enumeration grid
summary(candidates)                             # the best candidate on each criterion
bootstrap_lrt(fit, candidate_fit(candidates, n_profiles = 3, n_group_classes = 2),
              iter = 199, seed = 42)            # bootstrap test of one more class (slow)
enumerate_classes(course_engagement, vars = c("browse", "lectures"),
                  id = "student", n_profiles = 2:3,
                  model = c("VVI", "VVV"))      # crossing covariance models (slow)
confint(fit)                                    # confidence intervals
vcov(fit, vcov_type = "robust")                 # robust covariance matrix
plot(fit, what = "posteriors")                  # posterior of each assigned profile
plot(fit, what = "bars")                        # profile means with 95% intervals
diagnostics(fit)                                # all classification diagnostics
```

## Limits

- Information criteria compare the candidates fitted, not all possible models.
- Classification measures describe the fitted model; a misspecified model can
  still classify sharply.
- Wald intervals assume a converged fit with no variance at its bound, and are
  conditional on the chosen numbers of profiles and classes.
- Unmodelled dependence between indicators can make extra classes look better.
