---
title: "Growth mixture models: classes of trajectories"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Growth mixture models: classes of trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5,
                      fig.height = 4.6, fig.align = "center", dpi = 96)
options(width = 90)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

When the same people are measured repeatedly, a single average trajectory
often hides several kinds of change: some improve, some hold steady, some
decline. A **growth model with classes** finds those kinds from the data. This
vignette shows the two models `mixture_regression()` offers for it, how to
choose between them, and how to read the results.

## Two models

- **Latent class growth analysis** (Nagin 2005) gives every class its own
  trajectory. Everyone in a class follows that trajectory exactly; anything
  else is noise.
- **The growth mixture model** (Verbeke and Lesaffre 1996; Muthén and Shedden
  1999) also lets people vary *around* their class trajectory: each has their
  own start and rate of change (random intercepts and slopes).

The difference matters. When people vary within their kind, the first model
can only express that variation by adding classes, so it tends to find more
classes than there are (Bauer and Curran 2003). The second model separates
*kinds* of change from *individual* variation.

Both are one call. Persons are the classified units, so `id` names them and
`class_level = "group"` keeps all of a person's measurements in one class;
`random` adds the random effects.

| Role | Argument | Here |
|---|---|---|
| Outcome, measured repeatedly | left of `~` | `score` |
| Time | right of `~` | `wave` |
| Person | `id`, with `class_level = "group"` | `student` |
| Variation around the class trajectory | `random` | `"wave"`: own start and own slope |
| What predicts the class | `membership` | `"motivation"` |

## The data

`growth_scores` is simulated so the answer is known: 300 students tested on up
to seven waves, of three kinds (improving, stable, declining), each student
varying around their kind's trajectory, with motivation predicting the kind.
About 12% of tests were missed, which these models handle without imputation.

```{r data}
library(latents)
head(growth_scores)
```

## How many classes, and which model?

Fit both models for a range of classes and compare them in one table.
`random_covariance = "equal"` gives every class the same random-effect spread,
the usual first choice.

```{r fits, warning = FALSE}
trajectories <- function(k) {
  mixture_regression(score ~ wave, growth_scores, n_classes = k,
                     id = "student", class_level = "group",
                     n_starts = 5, seed = 1)
}
growth <- function(k) {
  mixture_regression(score ~ wave, growth_scores, n_classes = k,
                     id = "student", class_level = "group", random = "wave",
                     random_covariance = "equal", n_starts = 5, seed = 1)
}
comparison <- compare_models(
  trajectories_3 = trajectories(3), trajectories_4 = trajectories(4),
  trajectories_5 = trajectories(5),
  growth_2 = growth(2), growth_3 = growth(3), growth_4 = growth(4))
comparison
```

The growth mixture model with three classes is preferred, with nearly all
the BIC weight. The trajectory model without random effects keeps improving
as classes are added: five classes are still better than four, though only
three kinds of student exist. That is the over-extraction described above.

```{r comparison-plot, eval = has_ggplot, fig.height = 4}
plot(comparison)
```

For a test rather than a criterion, `enumerate_regressions()` fits a range of
classes and, with `bootstrap`, adds the bootstrap likelihood-ratio test of
each number of classes against one fewer (McLachlan 1987; Nylund, Asparouhov
and Muthén 2007). It refits both models on data simulated from the smaller
one, so it takes minutes; on these data it rejects two classes in favour of
three and keeps three against four.

```{r blrt, eval = FALSE}
enumerate_regressions(score ~ wave, growth_scores, n_classes = 1:4,
                      id = "student", class_level = "group", random = "wave",
                      random_covariance = "equal", bootstrap = 99, seed = 1)
```

## The chosen model

Refit it with motivation predicting the class:

```{r chosen}
fit <- mixture_regression(score ~ wave, growth_scores, n_classes = 3,
                          id = "student", class_level = "group",
                          random = "wave", random_covariance = "equal",
                          membership = "motivation", n_starts = 5, seed = 1)
fit
```

Every table is a data frame from `get_results()`; printing shows the columns
people report, while the data frame keeps every column (standard errors,
statistics, exact p-values and bounds).

```{r tables}
get_results(fit, "random")
get_results(fit, "membership")
```

Students start about 4 points apart and change at rates about 0.7 points per
wave apart within each class; those who start higher tend to grow a little
more slowly (a correlation of about -0.25). More motivated students are less
likely to be in the declining class (class 2) or the stable class (class 3)
than in the improving class.

Because the data were simulated, the classes can be checked against the truth:

```{r recovery}
get_results(fit, "recovery", data = growth_scores, truth = "trajectory")
```

## Distal outcomes

Do the classes differ on something measured once per student? `three_step()`
compares class means with the bias-corrected three-step method (Vermunt 2010;
Bakk and Vermunt 2016), which accounts for students being classified with
error. The trajectory model must be fitted without `membership` for this
correction, so motivation is treated here as a distal variable of the
classes from the model without it:

```{r distal}
trajectory_classes <- mixture_regression(score ~ wave, growth_scores,
                                         n_classes = 3, id = "student",
                                         class_level = "group", random = "wave",
                                         random_covariance = "equal",
                                         n_starts = 5, seed = 1)
three_step(trajectory_classes, data = growth_scores, outcome = "motivation",
           contrast = "pairs")
```

## Seeing the classes

The trajectory view shows each class's mean trajectory with its confidence
band, a lighter band where 80% of that class's students' own trajectories lie,
and the observed class means at every wave, which should sit on the curves.

```{r trajectories, eval = has_ggplot}
plot(fit)
```

```{r facet, eval = has_ggplot}
plot(fit, facet = TRUE)
```

Individual students, chosen across each class's range of classification
certainty: their observations, their class trajectory (dashed) and their own
predicted trajectory (solid).

```{r individuals, eval = has_ggplot, fig.height = 6.5}
plot(fit, what = "individuals")
```

The coefficients with their intervals and p-values:

```{r coefficients, eval = has_ggplot}
plot(fit, what = "coefficients")
```

Each student's predicted deviation from their class trajectory, with the
model's 95% ellipse:

```{r random, eval = has_ggplot}
plot(fit, what = "random")
```

How confidently each student is classified, one bar per student:

```{r classification, eval = has_ggplot, fig.height = 4.2}
plot(fit, what = "classification")
```

## Choices and warnings

- **Random effects.** `random = "intercept"` lets students differ in level
  only; `random = "wave"` in level and rate of change. A formula such as
  `~ 0 + wave` gives a random slope without a random intercept.
- **The random-effect spread.** `"equal"` shares it across classes;
  `"varying"` estimates one per class; `"proportional"` shares its shape and
  lets its size differ. `random_diagonal = TRUE` makes the random effects
  uncorrelated.
- **Curved trajectories.** Use `score ~ wave + I(wave^2)` or
  `score ~ splines::ns(wave, df = 3)`.
- **A degenerate covariance.** When the data cannot support a random effect in
  some class (a variance at zero, or two random effects perfectly
  correlated), the fit warns (`latents_random_boundary`), reports no interval
  for that covariance, and marks it as a boundary in the tables and plots.
  Simplify the random effects, or drop them.

## How this was checked

The model's likelihood equals an independently written dense-matrix
likelihood, and the analytic scores equal numerical derivatives. Against
`lcmm::hlme()` (Proust-Lima, Philipps and Liquet 2017), on configurations with
one to three classes, shared, proportional and diagonal covariances and
membership covariates, the likelihoods agree at `hlme`'s own estimates and
`mixture_regression()` reaches at least `hlme`'s maximum.

## References

Bakk, Z., & Vermunt, J. K. (2016). Robustness of stepwise latent class
modeling with continuous distal outcomes. *Structural Equation Modeling*, 23,
20--31.

Bauer, D. J., & Curran, P. J. (2003). Distributional assumptions of growth
mixture models: Implications for overextraction of latent trajectory classes.
*Psychological Methods*, 8, 338--363.

McLachlan, G. J. (1987). On bootstrapping the likelihood ratio test statistic
for the number of components in a normal mixture. *Applied Statistics*, 36,
318--324.

Muthén, B., & Shedden, K. (1999). Finite mixture modeling with mixture
outcomes using the EM algorithm. *Biometrics*, 55, 463--469.

Nagin, D. S. (2005). *Group-based modeling of development*. Harvard University
Press.

Nylund, K. L., Asparouhov, T., & Muthén, B. O. (2007). Deciding on the number
of classes in latent class analysis and growth mixture modeling: A Monte Carlo
simulation study. *Structural Equation Modeling*, 14, 535--569.

Proust-Lima, C., Philipps, V., & Liquet, B. (2017). Estimation of extended
mixed models using latent classes and latent processes: The R package lcmm.
*Journal of Statistical Software*, 78(2), 1--56.

Vermunt, J. K. (2010). Latent class modeling with covariates: Two improved
three-step approaches. *Political Analysis*, 18, 450--469.

Verbeke, G., & Lesaffre, E. (1996). A linear mixed-effects model with
heterogeneity in the random-effects population. *Journal of the American
Statistical Association*, 91, 217--221.
