---
title: "Two-level latent class analysis"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-level latent class analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", dpi = 96)
options(digits = 3, width = 100)
```

Two-level latent class analysis (LCA) is used when the indicators are
categorical: yes/no answers, ratings on a few ordered options, or unordered
choices. Like multilevel latent profile analysis, it estimates latent classes
of observations, called profiles in the output, and latent group classes that
differ in how probable each profile is for their observations. The difference
is in how a profile is described: by the probability of each answer to each
item, instead of a mean and a variance. `multilca()` fits LCA, treating every
indicator as categorical; `multilpa()` with some indicators named in
`categorical` fits a mixed model with both kinds of indicator.

## What the package supports

- **Item types.** Binary, ordinal and nominal items share one
  parameterization: a free probability for every category in every profile.
  Numeric, integer, logical, character and factor columns are accepted; factor
  levels keep their order, other types are sorted.
- **Mixed measurement.** Continuous and categorical indicators in one model.
- **Missing answers.** `missing = "fiml"` uses the observed answers of each
  row.
- **Standard errors.** Observed-information and robust
  (`vcov_type = "robust"`) standard errors through `parameter_inference()`.
- **Model comparison.** `enumerate_classes()` and, for complete data,
  `bootstrap_lrt()`.
- **External variables.** `three_step()` and `r3step()` work on a categorical
  fit.
- **Membership covariates.** `profile_covariates` and `group_covariates` work
  with categorical indicators; standard errors are not available for these
  fits.
- **Transitions.** `lta()` accepts categorical indicators.
- **Bounds.** `min_probability` keeps every response probability above a
  lower bound.

## Data

`student_esm` holds experience-sampling data: 2,582 prompts from 100 university
students, answered at times when the student had not been studying. `student`
identifies the group. Eight yes/no items record the leisure activities done
since the previous prompt, and four items rate affect from 1 to 7. `day` is the
study day, from 0 to 13. `?student_esm` gives the source.

```{r setup}
library(latents)
```

To keep the example short, we call `subset()` with `day <= 6`, which keeps the
first week and all 100 students.

```{r first-week}
first_week <- subset(student_esm, day <= 6)
```

The eight activity items are used in several fits, so we store their names
once.

```{r activities}
activities <- c("time_with_friends", "on_social_media", "tv_video_games",
                "listened_music", "sports", "walking", "reading",
                "part_time_job")
```

## Fitting the model

To fit the model, we call `multilca()` with `vars` for the categorical
indicators, `id` for the group, `n_profiles`, `n_group_classes`, `n_starts`,
`seed`, and `tol` for a tighter convergence tolerance, which standard errors
need.

```{r lca}
lca <- multilca(first_week, vars = activities, id = "student",
                n_profiles = 2, n_group_classes = 2, n_starts = 10,
                tol = 1e-10, seed = 1)
lca
```

To obtain the response probabilities, we call `get_results()` with
`"responses"`. There is one row per profile, item and category;
`probability` is the chance of that answer in that profile.

```{r responses}
get_results(lca, "responses")
```

To see them, we call `plot()` with `what = "responses"`: one line per
profile across the items.

```{r responses-plot}
plot(lca, what = "responses")
```

To obtain the profile probabilities of each group class, we call
`get_results()` with `"profile_probabilities"`, and to compare the classes,
`plot()` with `what = "probabilities"`.

```{r profile-probabilities}
get_results(lca, "profile_probabilities")
```

```{r probability-plot}
plot(lca, what = "probabilities")
```

To obtain classification certainty, we call `get_results()` with
`"entropy"`.

```{r entropy}
get_results(lca, "entropy")
```

## Mixed indicators

To combine both kinds, we call `multilpa()` with all indicators in `vars` and
only the categorical ones in `categorical`. Here three affect ratings are
continuous and the activities categorical. `worried` is left out: about half
its ratings are 1, which can push a profile's variance to its lower bound.

```{r mixed}
mixed <- multilpa(first_week,
                  vars = c("happy", "relaxed", "exhausted", activities),
                  categorical = activities, id = "student", n_profiles = 2,
                  n_group_classes = 2, n_starts = 10, seed = 1)
```

To obtain the means and variances of the continuous indicators, we call
`get_results()` on the fit; for the answer probabilities of the categorical
ones, `get_results()` with `"responses"`.

```{r mixed-results}
get_results(mixed)
get_results(mixed, "responses")
```

## Reference

The calls below list other functions that work with categorical and mixed
models. Each comment states what the call returns.

```{r reference, eval = FALSE}
get_results(lca, "classification")               # class sizes and average posteriors
get_results(lca, "residuals")                    # residual association within profiles
parameter_inference(lca)                         # standard errors and intervals
enumerate_classes(first_week, vars = activities, categorical = activities,
                  id = "student", n_profiles = 2:3,
                  n_group_classes = 1:2, seed = 1)  # compare numbers of classes (slow)
plot(mixed)                                      # means of the continuous indicators
plot(mixed, what = "responses")                  # answers to the categorical ones
multilca(first_week, vars = activities, id = "student", n_profiles = 2,
         n_group_classes = 2, missing = "fiml")  # incomplete answers, if any
```

## Assumptions

- Items are independent within a profile.
- Every category of every item has its own probability in each profile, so
  rare answers can push probabilities to their lower bound (`min_probability`).
- Categories are taken from the observed values; factor levels keep their
  order.
