---
title: "Multilevel latent profile analysis"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multilevel latent profile analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", dpi = 96)
options(digits = 3, width = 100)
```

Multilevel latent profile analysis (MLPA) estimates latent profiles at the
observation level and, at a higher level, identifies latent group classes that
differ in how probable each profile is for their observations. Each profile is characterized by a mean and variance for
each indicator and is defined identically across all group classes. The group
classes therefore differ only in the probabilities associated with the
profiles.

## Data

`course_engagement` has one row per course enrolment: 1,422 enrolments from
106 students. `multilpa()` needs two kinds of variables.

- **Group:** `student` identifies which enrolments belong to the same
  student.
- **Indicators:** five activity measures (`browse`, `lectures`, `forum_read`,
  `forum_post`, `attendance`), each standardized within course, so 0 is the
  course average.

The data are simulated, and two further variables record the structure that
generated them: `engagement` for each enrolment and `student_type` for each
student.

```{r setup}
library(latents)
```

## Clustering within groups

To check how much each indicator varies between groups, we call
`descriptives()` with `vars` for the indicators and `id` for the group. `icc`
is the share of an indicator's variance that lies between groups.

```{r descriptives}
descriptives(course_engagement,
             vars = c("browse", "lectures", "forum_read", "forum_post",
                      "attendance"),
             id = "student")
```

## Fit

To fit the model, we call `multilpa()` with `vars` for the indicators, `id`
for the group, `n_profiles` for the number of profiles, `n_group_classes` for
the number of group classes, `n_starts` for the number of random EM starts, and
`seed` for reproducible starts. Printing the fit gives the profile means.

```{r fit}
fit <- multilpa(course_engagement,
                vars = c("browse", "lectures", "forum_read", "forum_post",
                         "attendance"),
                id = "student", n_profiles = 2, n_group_classes = 2,
                n_starts = 10, seed = 1)
fit
```

To check whether the starts agree, we call `get_results()` with `"starts"`.
Starts that reach the same log likelihood support the solution.

```{r starts}
get_results(fit, "starts")
```

## Profiles

To obtain the profile means and variances, we call `get_results()` on the fit.
There is one row per profile and indicator.

```{r profiles}
get_results(fit)
```

To see the profiles, we call `plot()` on the fit. Each line is a profile's
mean on each indicator, and point size shows how common the profile is.

```{r profile-plot}
plot(fit)
```

## Group classes

To obtain the profile probabilities of each group class, we call
`get_results()` with `"profile_probabilities"`. `probability` is the chance
that an observation of a group in that class falls in each profile, and
`group_class_probability` is the share of groups in the class.

```{r profile-probabilities}
get_results(fit, "profile_probabilities")
```

To compare the group classes, we call `plot()` with `what = "probabilities"`:
one line per group class across the profiles.

```{r probability-plot}
plot(fit, what = "probabilities")
```

## Class sizes

To obtain class sizes, we call `get_results()` with `"counts"`.
`effective_count` sums the posterior probabilities, so it need not be a whole
number.

```{r counts}
get_results(fit, "counts")
```

## Classification

To obtain classification certainty, we call `get_results()` with `"entropy"`.
Relative entropy is reported for observations and for groups: 1 means every
case is assigned with certainty, 0 means no separation.

```{r entropy}
get_results(fit, "entropy")
```

To add each observation's assigned profile and group class to the data, we
call `get_results()` with `"assignments"` and `data` for the data frame. It
returns the data in its original row order with five added columns: `profile`
and `group_class` hold the most probable classes, `uncertainty` is one minus
the posterior probability of the assigned profile, and the `posterior_profile_`
columns hold the posterior probability of each profile. Assignment takes the
most probable class, so `profile` and `group_class` alone ignore
classification uncertainty.

```{r assignments}
get_results(fit, "assignments", data = course_engagement) |> head()
```

## Recovery of the known structure

To compare the fitted classes with known labels, we call `get_results()` with
`"assignments"`, `data` for the data frame, and `truth` for the columns that
hold the labels. `engagement` is compared with the profiles and
`student_type` with the group classes. Class numbers are arbitrary, so the
table is read by its largest cells.

```{r recovery}
get_results(fit, "assignments", data = course_engagement,
            truth = c("engagement", "student_type"))
```

## Reference

The calls below list the other tables, functions and views available for a
fitted model. Each comment states what the call returns.

```{r reference, eval = FALSE}
# Tables: get_results(fit, what)
get_results(fit, "all")                          # every table, as a named list
get_results(fit, "covariances")                  # within-profile residual covariances
get_results(fit, "posteriors")                   # one row per observation and profile
get_results(fit, "posteriors", format = "wide")  # one row per observation
get_results(fit, "group_posteriors")             # one row per group and group class
get_results(fit, "classification")               # class sizes and average posteriors
get_results(fit, "average_posteriors")           # mean posterior by assigned class
get_results(fit, "classification_errors")        # P(assigned class | true class)
get_results(fit, "bch_weights")                  # weights for three-step analysis
get_results(fit, "residuals")                    # residual correlations within profiles
get_results(fit, "information_criteria")         # AIC, BIC and the other criteria
get_results(fit, "model")                        # one row describing the fit
get_results(fit, "data")                         # the variables used in the fit
as.data.frame(fit)                               # the profile table

# Summaries and diagnostics
descriptives(fit)                    # indicator summaries with the ICC
descriptives(fit, by = "profile")    # the same, by assigned profile
diagnostics(fit)                     # every classification diagnostic
report(fit)                          # summary, diagnostics and plots together

# Inference and likelihood
parameter_inference(fit)             # standard errors and Wald intervals
confint(fit)                         # confidence intervals
vcov(fit)                            # covariance matrix of the estimates
logLik(fit)                          # log likelihood
AIC(fit)                             # Akaike information criterion
BIC(fit)                             # BIC with the number of groups
nobs(fit)                            # number of groups

# Plots: plot(fit, what)
plot_views()                # every view and what it shows
plot(fit, what = "bars")             # profile means with 95% intervals
plot(fit, what = "heatmap")          # means in standard deviations
plot(fit, what = "sizes")            # effective size of each profile
plot(fit, what = "entropy")          # per-case classification uncertainty
plot(fit, what = "posteriors")       # posterior of each assigned profile
plot(fit, what = "avepp")            # average posteriors, assigned by posterior class
plot(fit, what = "all")              # every view this fit supports

# Other functions
sensitivity(fit, seeds = 1:10)       # refits under several seeds
starting_values(fit)                 # parameters to start another fit
```

The evaluation vignette covers `enumerate_classes()`, `candidate_fit()` and
`bootstrap_lrt()`; the covariates vignette covers `three_step()`, `r3step()`
and `fit_staged()`; the transitions vignette covers `lta()`, `get_tna()` and
`get_group_tna()`.

## All tables

To see every table at once, we call `summary()` on the fit, with `rows` for
the number of rows shown per table.

```{r summary}
print(summary(fit), rows = 3)
```

## Assumptions

- Profiles have the same means and variances in every group class.
- Observations are independent given their group's class, and groups are
  independent of one another.
- Indicators are independent within a profile (the default diagonal
  covariance; `covariance_model = "full"` relaxes it).
- The numbers of profiles and group classes are chosen by the analyst; see
  the evaluation vignette for comparing them.
