---
title: "Mixture regression"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixture regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", dpi = 96)
old_options <- options(digits = 3, width = 100)
```

A single regression assumes that one line describes everyone. Mixture
regression drops that assumption: the data are taken to come from a small
number of latent classes, each with its own regression, and the model
estimates the regressions, how common each class is, and how probably each
observation belongs to each class — all at once. The method goes by several
names: clusterwise regression, latent class regression, regression mixture
(DeSarbo and Cron 1988; Wedel and DeSarbo 1995).

`mixture_regression()` fits it for continuous (`"gaussian"`), binary or binomial
(`"binomial"`), count (`"poisson"`, and `"negative_binomial"` for overdispersed
counts) and ordered (`"ordinal"`, cumulative logit) outcomes, at one level or
two:

| Question | Call |
|---|---|
| Do observations follow different regressions? | `mixture_regression(y ~ x, data, n_classes)` |
| Do *groups* (persons, schools) follow different regressions? | `mixture_regression(..., id = "person", class_level = "group")` |
| Do groups differ in how often their observations follow each regression? | `mixture_regression(..., id = "person", n_group_classes = 2)` |
| What predicts which class an observation (or group) is in? | `membership = ~ z`, `group_membership = ~ v` |
| Is some effect the same in every class? | `common = ~ x2` |
| How many classes? | `enumerate_regressions(..., n_classes = 1:4)` |

## The data

`study_hours` is simulated so that the truth is known. 150 students report
six weeks each. In each week a student studies with a *deep* or a *surface*
strategy, and the strategy decides how hours turn into a quiz score: deep
weeks gain about 4.5 points per hour from a base of 35, surface weeks under a
point per hour from a base of 55. Steady students study deeply in most weeks,
erratic students in few. The generating strategy and student type are kept in
the data so recovery can be checked; the models never see them.

```{r data}
library(latents)
head(study_hours)
```

## One class per observation

```{r single}
fit <- mixture_regression(score ~ hours, data = study_hours, n_classes = 2,
                          n_starts = 5, seed = 1)
fit
```

The two classes are the two strategies: a steep regression from a low base and
a flat one from a high base. Every table comes from `get_results()`; the
class table gives each class's share, its residual standard deviation and how
certainly its members are assigned.

```{r classes}
get_results(fit, "classes")
```

`get_results(fit, "assignments")` gives one row per observation. Because
these data carry the generating strategy, the recovery table cross-tabulates
the assigned classes against it:

```{r recovery}
get_results(fit, "recovery", data = study_hours, truth = "strategy")
```

```{r plot-fitted}
plot(fit)
```

## How many classes?

`enumerate_regressions()` fits a range of class counts and returns their fit
statistics in one table, marking the minimum-BIC model.

```{r enumerate}
classes <- enumerate_regressions(score ~ hours, data = study_hours,
                                 n_classes = 1:4, n_starts = 3, seed = 1)
classes
```

With `bootstrap = B`, each row also gets a parametric bootstrap
likelihood-ratio test of one class fewer (McLachlan 1987): outcomes are
simulated from the smaller model, both models are refitted, and the p-value
is the share of simulated statistics at least as large as the observed one.
It is slow — `2 B` refits per row — so it is not run here.

## What predicts the class?

`membership` adds a multinomial logit for class membership. Here sleep before
the quiz makes a deep week more likely, so it should lower the odds of the
surface class:

```{r membership}
with_sleep <- mixture_regression(score ~ hours, data = study_hours, n_classes = 2,
                                 membership = ~ sleep, n_starts = 5, seed = 1)
get_results(with_sleep, "membership")
```

The coefficients are estimated jointly with the regressions — a one-step
model, not a regression on assigned classes — so classification error does
not attenuate them.

## Groups that share a class

When every row of a group belongs to the same class, pass `id` and
`class_level = "group"`. A group's evidence for a class is the product of its
rows' likelihoods, so classes of *students* are separated far more sharply
than classes of weeks could be. The binary `passed` outcome in these data
depends on the student's type, not on the week:

```{r group}
students <- mixture_regression(passed ~ hours, data = study_hours, n_classes = 2,
                               family = "binomial", id = "student",
                               class_level = "group", membership = ~ motivation,
                               n_starts = 5, seed = 1)
students
get_results(students, "membership")
```

A mixture of logistic regressions with a single binary outcome per class
assignment is not identified — any mixture of Bernoulli distributions is a
Bernoulli distribution (Follmann and Lambert 1991) — so `mixture_regression()` refuses
`family = "binomial"` with one trial per row unless the class is shared by a
group:

```{r not-identified, error = TRUE}
mixture_regression(passed ~ hours, data = study_hours, n_classes = 2,
                   family = "binomial")
```

## Two levels: groups that differ in their mix

The deep/surface strategy changes from week to week, but steady and erratic
students use the strategies in different proportions. That is a two-level
mixture (Vermunt 2003): rows keep their own regression class, and a latent
*group class* shifts the class probabilities of all rows in its groups.

```{r two-level}
two_level <- mixture_regression(score ~ hours, data = study_hours, n_classes = 2,
                                id = "student", n_group_classes = 2,
                                membership = ~ sleep, group_membership = ~ motivation,
                                n_starts = 5, seed = 1)
summary(two_level)
```

The `group_classes` table is the heart of it: within each group class, the
probability of each regression class. The group-class membership model shows
motivation separating the two kinds of student, and the class model shows the
week-level effect of sleep.

```{r two-level-recovery}
get_results(two_level, "recovery", data = study_hours,
            truth = "student_type", by = "group_class")
```

## Counts, shared effects and equal variances

The Poisson family takes counts, and `offset()` terms for exposure. `common`
names predictors whose coefficient is the same in every class; for the
Gaussian family `variance = "equal"` shares one residual standard deviation.

```{r counts}
asked <- mixture_regression(questions ~ hours + sleep, data = study_hours, n_classes = 2,
                            family = "poisson", common = ~ sleep, n_starts = 5, seed = 1)
get_results(asked, "coefficients")
```

`exp_estimate` is the rate ratio of each coefficient.

## Uncertainty

Standard errors come from the observed information by default. Scores are
analytic (the posterior expectation of the complete-data score), and the
information is their numerical derivative. `vcov_type = "robust"` gives the
sandwich estimator, clustered on the independent unit — the row, or the group
when `id` defines one, including a single-level fit given `id`.

```{r robust}
get_results(fit, "coefficients", vcov_type = "robust")
plot(fit, what = "coefficients")
```

`p_adjusted` applies the Benjamini-Hochberg correction across the slopes.

## Prediction and simulation

`predict()` gives the class-averaged mean (`"response"`), every class's mean
(`"class_response"`), or — when the outcome is present — the posterior class
probabilities of new rows (`"posterior"`). `simulate()` draws new outcomes
from the fitted model, which is what the bootstrap test uses.

```{r predict}
new_weeks <- data.frame(hours = c(1, 6, 11))
predict(fit, new_weeks, type = "class_response")
```

## Checking a solution

Mixture likelihoods have local maxima. `mixture_regression()` starts EM from the
residuals of a pooled regression and from `n_starts` random partitions, and
the `starts` table shows how many starts reached the best likelihood. A
solution reached by one start only deserves more starts.

```{r starts}
get_results(fit, "starts")
plot(fit, what = "posteriors")
```

## How this compares with other software

The single-level and group-level models match `flexmix` (Leisch 2004): the
same likelihood at the same parameters, and the same binomial and Poisson
estimates. `flexmix` divides the Gaussian residual sum of squares by
`n - p`, so its Gaussian estimates are not quite the maximum-likelihood ones
and `mixture_regression()` reaches a slightly higher likelihood. `flexmix` has no
two-level model, no analytic standard errors (it refits with `optim()`), and
no bootstrap likelihood-ratio test.

## References

DeSarbo, W. S., & Cron, W. L. (1988). A maximum likelihood methodology for
clusterwise linear regression. *Journal of Classification*, 5, 249–282.

Follmann, D. A., & Lambert, D. (1991). Identifiability of finite mixtures of
logistic regression models. *Journal of Statistical Planning and Inference*,
27, 375–381.

Leisch, F. (2004). FlexMix: A general framework for finite mixture models and
latent class regression in R. *Journal of Statistical Software*, 11(8).

McLachlan, G. J. (1987). On bootstrapping the likelihood ratio test statistic
for the number of components in a normal mixture. *Applied Statistics*, 36,
318–324.

Vermunt, J. K. (2003). Multilevel latent class models. *Sociological
Methodology*, 33, 213–239.

Wedel, M., & DeSarbo, W. S. (1995). A mixture likelihood approach for
generalized linear models. *Journal of Classification*, 12, 21–55.

```{r restore-options, include = FALSE}
options(old_options)
```
