---
title: "Do schools differ in the students they serve?"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Do schools differ in the students they serve?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>",
                      fig.width = 7, fig.height = 4.2, fig.align = "center",
                      dpi = 96)
options(digits = 4, width = 92)
```

```{r library}
library(multilpa)
```

## The question is about schools, not students

Latent profile analysis asks what kinds of *people* there are. This package
exists for the question one level up: given that there are kinds of people, do
the **groups** they sit in differ in the mix of people they contain?

That is not a rephrasing. It is a different parameter, and a single-level model
does not have it.

Sixty secondary schools, twelve students each. For every student we have weekly
homework hours, a teacher's participation rating, and a self-reported interest
score. Half the schools draw a mostly engaged intake and half do not — but we
will fit the model as though we did not know that, and see whether it finds out.

```{r simulate}
set.seed(2024)

n_schools <- 60
per_school <- 12

engaged_share <- ifelse(rep(1:2, length.out = n_schools) == 1L, 0.75, 0.25)

students <- data.frame(
  school = rep(seq_len(n_schools), each = per_school),
  term   = rep(seq_len(per_school), times = n_schools)
)
students$engaged <- runif(nrow(students)) < rep(engaged_share, each = per_school)

engaged <- students$engaged
students$homework_hours <- round(rnorm(nrow(students), ifelse(engaged, 8.0, 4.2), 1.2), 1)
students$participation  <- round(rnorm(nrow(students), ifelse(engaged, 7.2, 4.0), 1.1), 1)
students$interest       <- round(rnorm(nrow(students), ifelse(engaged, 7.4, 4.3), 1.2), 1)

vars <- c("homework_hours", "participation", "interest")
```

Here is the reason the second level is not a technicality. Across the whole
sample, **`r round(100 * mean(students$engaged))`%** of students are the engaged
kind — a tidy headline number. Now look at the schools one at a time:

```{r the-hook}
per_school_share <- tapply(students$engaged, students$school, mean)
round(range(per_school_share), 2)
```

The share runs from **`r round(min(per_school_share), 2)`** to
**`r round(max(per_school_share), 2)`**. The sample average describes no school
in the data. Any analysis that pools the students and reports one number has
averaged away the thing worth knowing.

## Look at the profiles before reading anything about them

```{r fit}
fit <- multilpa(students, vars, id = "school",
                n_profiles = 2, n_group_classes = 2,
                n_starts = 8, seed = 1)
```

```{r plot-profiles}
plot(fit)
```

Two clearly separated profiles: one doing around four hours of homework with
low participation and interest, one doing around eight with both high. The
separation is what makes the rest of the analysis worth doing; if the lines
crossed or sat on top of each other, no amount of fit statistics would rescue
it.

The numbers behind that picture, one row per profile and indicator:

```{r measurement}
as.data.frame(fit)
```

The indicators are on different scales, so the raw plot partly shows that
homework hours are simply bigger numbers. Standardizing puts them on a common
footing and shows the *shape* of each profile:

```{r plot-standardized}
plot(fit, scale = "standardized")
```

The same estimates can be read three ways, and which one to use depends on the
question. Bars, with a 95% interval on each, answer *how far apart are the two
profiles on this indicator, and is the gap larger than its uncertainty*:

```{r plot-bars, fig.height = 4.6}
plot(fit, what = "bars", data = students)
```

A heatmap of standard deviations from each indicator's grand mean answers
*which indicators define the profiles* at a glance, and stays readable when
there are more indicators than a line plot can label:

```{r plot-heatmap, fig.height = 3.4}
plot(fit, what = "heatmap")
```

Every view is listed, so none of them has to be remembered:

```{r plot-catalogue}
multilpa_plot_types()
```

Profile labels are arbitrary. Nothing in the likelihood distinguishes "profile
1" from "profile 2", so a refit may swap them. Read the estimates against what
generated them, not against the label.

## How many profiles, and how many kinds of school?

Both counts are chosen, not estimated, so choose them in the open. Fit the grid:

```{r enumerate}
candidates <- enumerate_classes(students, vars, id = "school",
                                n_profiles = 1:3, n_group_classes = 1:2,
                                n_starts = 4, seed = 1)
```

```{r plot-enumeration}
plot(candidates, criterion = "bic_groups")
```

`summary()` reports which candidate each criterion prefers, rather than picking
one for you:

```{r summary-candidates}
summary(candidates)
```

They do not agree, and that is the normal case rather than a problem with the
data. The BIC family, CAIC, AWE and ICL all choose two profiles in two kinds of
school. AIC and CLC choose three profiles. This is the familiar pattern that
AIC under-penalizes in mixture models and tends to over-extract; the BIC family
is the more conservative read, and it happens to be the structure these data
were simulated from.

Note also that every sample-size-dependent criterion is reported twice, under
`groups` and under `individuals`. A two-level model has two defensible sample
sizes — 60 schools or 720 students — and neither is privileged. Pick one
convention and keep it for the whole analysis rather than choosing per table.

```{r criteria}
information_criteria(fit)
```

Lower is better on every row, including the first: the table reports the
deviance rather than the log likelihood, so `value` means one thing throughout
and equals `deviance + penalty`.

## The answer: the schools do differ

This is the quantity the second level exists for, and the one a single-level
model cannot report at all.

```{r prevalence}
as.data.frame(fit, what = "profile_probabilities")
```

```{r plot-probabilities}
plot(fit, what = "probabilities")
```

Two kinds of school, in roughly equal numbers, and the balance of profiles
reverses between them: the profile that is the clear majority in one kind is the
clear minority in the other. The pooled
`r round(100 * mean(students$engaged))`% from the first section is the average
of those two populations and a description of neither.

To see what pooling costs, fit the same measurement model with a single group
class — that *is* the single-level model:

```{r flat}
flat <- multilpa(students, vars, id = "school",
                 n_profiles = 2, n_group_classes = 1,
                 n_starts = 8, seed = 1)

as.data.frame(flat, what = "profile_probabilities")
```

One row per profile, one number each, and no way to ask the question. The
two-level model also fits better by a wide margin:

```{r compare}
lmr_lrt(flat, fit)
```

## How much should you trust it?

Three things are worth checking before any of the above is reported.

**Are the classes well separated?** Entropy near one means the posterior
assigns students confidently.

```{r entropy}
entropy_table(fit)
```

Individuals separate almost perfectly; schools less so, which is expected —
twelve students is not much evidence about a school.

Entropy is one number for the whole model, which hides where the uncertainty
actually sits. These two views show the distribution behind it. The first is
each student's contribution to the entropy sum, zero being a certain
assignment; the second is the posterior probability of the profile each student
was actually assigned to, where mass piled against one is a clean separation:

```{r plot-entropy, fig.height = 3.6}
plot(fit, what = "entropy")
```

```{r plot-posteriors, fig.height = 3.6}
plot(fit, what = "posteriors")
```

Each ridge is scaled to its own maximum, which is the usual ridgeline
convention: it compares the *shapes* and not the profile sizes, so prevalence
is printed in each label instead. A ridge with a long tail towards the middle
is where a single summary number would have misled you.

**Are the estimates precise?** Standard errors, with the measurement model:

```{r errors}
as.data.frame(fit, data = students)
```

Robust errors accumulate the score over independent schools rather than
students, which is the right unit when the measurement model may be
misspecified:

```{r robust}
head(parameter_inference(fit, data = students, vcov_type = "robust"), 6)
```

**Does the model's core assumption hold?** Latent profile analysis assumes the
indicators are independent *within* a profile. Where they are not, the profiles
are partly absorbing a correlation instead of describing a kind of student.

```{r residuals}
bivariate_residuals(fit, data = students)
```

`observed` and `expected` are both associations on the correlation scale, so
`residual` is their difference and the rows are comparable. All are small here;
a large one would argue for `covariance_model = "full"`.

## When students are measured more than once

If the same people are observed on several occasions, the question changes from
*what kinds are there* to *who moves between them*. That needs a different
dataset — in the data above, `term` indexes twelve different students within a
school, so a "transition" would be between two different people and would mean
nothing.

```{r panel}
set.seed(11)
n_students <- 120
n_waves <- 4
persistence <- 0.85

panel <- data.frame(
  student = rep(seq_len(n_students), each = n_waves),
  wave    = rep(seq_len(n_waves), times = n_students)
)

# Each wave depends on the one before, so the chain is built with
# Reduce(accumulate = TRUE) rather than a loop.
waves <- Reduce(function(previous, ignored) {
  ifelse(runif(n_students) < persistence, previous, 3L - previous)
}, seq_len(n_waves - 1L),
   init = sample(1:2, n_students, replace = TRUE), accumulate = TRUE)

panel$state <- as.vector(t(do.call(cbind, waves)))
high <- panel$state == 1L
panel$homework_hours <- round(rnorm(nrow(panel), ifelse(high, 8.0, 4.2), 1.2), 1)
panel$participation  <- round(rnorm(nrow(panel), ifelse(high, 7.2, 4.0), 1.1), 1)

moves <- fit_transitions(panel, c("homework_hours", "participation"),
                         id = "student", n_profiles = 2, time = "wave",
                         n_starts = 6, seed = 1)

transitions(moves)
```

Students were simulated to stay put with probability `r persistence`. The
diagonal of the fitted matrix is what `stable = TRUE` marks, and it recovers
that persistence to within a few points — close, not exact, which is what four
waves of 120 students supports.

## What the package refuses

A model family without standard errors says so, by raising a condition you can
catch by class rather than by matching message text:

```{r refuses, error = TRUE}
vcov(moves)
```

Every refusal in the package is a classed condition; `?"multilpa-conditions"`
lists all of them.
