---
title: "Multilevel latent profile analysis of course engagement"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multilevel latent profile analysis of course engagement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>",
                      fig.width = 7, fig.height = 4.2, fig.align = "center",
                      dpi = 96)
options(digits = 4, width = 92)
```

Multilevel latent profile analysis (MLPA) is used when observations are nested within higher-level units and the aim is to identify latent profiles at the observation level while also describing how the prevalence of those profiles varies across units. Examples include repeated measurements nested within persons, students nested within schools, patients nested within clinics, or employees nested within organizations. Because observations from the same higher-level unit may be dependent, MLPA extends standard latent profile analysis by modeling the data at two levels. Latent profiles are defined at Level 1, whereas latent classes at Level 2 describe differences between higher-level units in the composition of those profiles.

Consider a repeated-measures application. A person may occupy different profiles on different occasions. Let $C_{it}$ denote the profile of person $i$ at occasion $t$, and let $G_i$ denote the person's Level-2 latent class. The model estimates the conditional probability $P(C_{it}=k\mid G_i=m)$ that an observation belongs to profile $k$ given that the person belongs to class $m$. In this way, MLPA separates variation across repeated observations from variation between persons in the profiles they tend to occupy.

This distinction is also central to VaSSTra, which separates **state diversity** from **person heterogeneity** using a sequential estimation strategy. VaSSTra first establishes a common state space across all observations and then summarizes each person's repeated observations within that space. MLPA takes a different approach: the observation-level profiles and person-level classes are estimated jointly within a single model. Both approaches allow the same person to occupy several states across repeated observations, but they differ in how the states and the person-level representation are constructed.

The `multilpa` package fits MLPA models using maximum likelihood with multiple expectation-maximization starting values. In this vignette, we work through a three-profile, three-person-class example using longitudinal data on course engagement. The focus is on specifying, estimating, and interpreting the model. The companion guides cover [model evaluation](multilpa-evaluation.html), [covariates and staged estimation](multilpa-covariates.html), and [latent transitions](multilpa-transitions.html). 

```{r setup}
library(multilpa)
```

## Longitudinal course engagement


The engagement example revisits the setting studied by
[Saqr and López-Pernas (2021)](https://doi.org/10.1016/j.compedu.2021.104325),
*The longitudinal trajectories of online engagement over a full program*. The
study distinguished three engagement states within courses and three
longer-term student trajectories. Here, we use the bundled synthetic
`course_engagement` data to reproduce the same type of nested longitudinal
design. The purpose of the example is to illustrate the analytical questions
and workflow; the estimates reported below apply only to the synthetic package
data.

Each row represents one student's enrolment in one course. Enrolments are
therefore the Level-1 observations, while students are the Level-2 units within
which those observations are nested. A student who takes several courses
appears on several rows. The variable `student` identifies the Level-2 unit,
`course` identifies the course associated with each enrolment, and `sequence`
records the position of that course in the student's observed history.
Students need not contribute the same number of enrolments. The MLPA model
accounts for the grouping of enrolments within students; it does not include a
separate course-level random effect or latent structure.

Five activity indicators describe the student's engagement during each
enrolment. The indicators are stored as `log1p(count)` values standardized
within course. Each value can therefore be interpreted as the enrolment's
distance from the mean of that particular course, expressed in course-specific
standard deviations. A value above zero indicates more activity than the
course average, whereas a value below zero indicates less activity than the
course average.

This within-course standardization is important for the interpretation of the
profiles. Raw activity volumes can differ substantially across courses because
of differences in course design, duration, or expected workload. The indicators
therefore describe a student's engagement relative to other students taking
the same course rather than relative to the cohort as a whole. The profiles
identified below should consequently be interpreted as recurring patterns of
**relative engagement within courses**, not as absolute levels of activity
across the program.

| Variable | Definition |
|---|---|
| `browse` | Course page views. |
| `lectures` | Lecture videos viewed. |
| `forum_read` | Forum posts read. |
| `forum_post` | Forum posts written. |
| `attendance` | Days active in the course. |
| `previous_grade` | Standardized grade from the preceding course, with an entry grade at the first course. |

The indicator vector `vars` selects the five measures used to define engagement
profiles. Identifiers and `previous_grade` stay outside the measurement model.
The package accepts a character vector of column names.

```{r variables}
vars <- c("browse", "lectures", "forum_read", "forum_post", "attendance")
```

## Descriptive statistics and clustering within students

The `descriptives()` function summarizes the distributions of the activity indicators and estimates their intraclass correlations (ICCs), using `student` as the grouping variable. The ICC describes how much of the variation in an indicator is attributable to differences between students rather than variation across enrolments within the same student. It is defined as

$$
\mathrm{ICC}
=
\frac{\sigma^2_{\mathrm{between}}}
{\sigma^2_{\mathrm{between}}+\sigma^2_{\mathrm{within}}}.
$$


Here, $\sigma^2_{\mathrm{between}}$ is the variance between students and
$\sigma^2_{\mathrm{within}}$ is the variance across enrolments within students.
An ICC near zero indicates little between-student variation in an indicator's
average level, whereas an ICC near one indicates little variation across
enrolments within students. With the one-way random-effects estimator used
here, an estimated ICC can occasionally be negative because of sampling
variation; such estimates are generally interpreted as indicating little or no
detectable between-student clustering for that indicator.

ICCs describe the clustering of each activity indicator separately. They do not
determine how many Level-1 profiles or Level-2 student classes should be fitted.
In particular, students may differ in the **composition of their engagement
profiles** even when their average levels on the individual indicators are
similar. The ICCs are therefore descriptive measures of the multilevel
structure in the observed indicators rather than criteria for selecting the
latent structure.

The descriptive table reports means, standard deviations, ranges, and ICCs for
the activity indicators, together with sample sizes and missingness.


```{r descriptives}
descriptives(course_engagement, vars = vars, id = "student")
```

The data contain 1,422 enrolments from 106 students, with no missing values in
the activity indicators. The estimated ICCs range from 0.093 for `lectures` to
0.225 for `forum_read`, indicating that approximately one tenth to one quarter
of the variance in these indicators lies between students. Thus, most variation
occurs across enrolments within students, while a non-negligible share reflects
systematic differences between students. The repeated enrolments therefore
contain information about both variation in engagement across courses and
between-student heterogeneity in engagement.

## Three engagement profiles and three student classes

The `multilpa()` function estimates engagement profiles from the five activity indicators and
student classes from the distribution of those profiles across each student's
enrolments. A **3 × 3 model** contains three Level-1 enrolment profiles and
three Level-2 student classes. Each student class is characterized by a
different set of probabilities for the three enrolment profiles, while an
individual student may contribute enrolments to more than one profile.

Each engagement profile is defined by a mean and variance for every activity
indicator. These profile-specific parameters are shared across student classes.
Thus, a given profile represents the same pattern of engagement regardless of
which student class it occurs in. What differs across student classes is the
relative frequency with which their enrolments occupy the profiles.

The conditional profile probabilities introduced above can be written as

$$
\pi_{k\mid m}=P(C_{it}=k\mid G_i=m),
$$

where $\pi_{k\mid m}$ is the probability that an enrolment belongs to profile
$k$ when the student belongs to class $m$. For each student class, these
probabilities range from zero to one and sum to one across the three profiles.
The Level-2 class probability

$$
\omega_m=P(G_i=m)
$$

gives the proportion of students represented by class $m$.

The model makes several distributional and independence assumptions. Activity
indicators are assumed to follow Gaussian distributions within each profile,
and the profile definitions are assumed to be common across students. Students
are treated as independent Level-2 units, while enrolments are conditionally
independent given the student's latent class. Under the default diagonal
covariance structure, the five activity indicators are also assumed to be
independent within a profile after conditioning on profile membership. Residual
associations among the indicators that remain within profiles may therefore
signal a departure from this assumption and can influence the estimated
profile structure.

The choice of three profiles and three student classes is a candidate model
rather than an assumption about the true latent structure. The
[evaluation guide](multilpa-evaluation.html) compares this specification with
models containing different numbers of Level-1 profiles and Level-2 classes.

The analysis below uses ten random starts to assess whether estimation
repeatedly converges to the same likelihood solution. Setting seed = 1 makes
the set of starting values reproducible and restores the caller's random state
after estimation. A tighter tol is used to support the subsequent calculation
of standard errors. Finally, time = "sequence" retains the observed course
order for displaying students' enrolment histories. The present model describes
the composition of profiles across a student's courses; it does not model
transitions from one profile to another. Transition probabilities are estimated
with the separate longitudinal model.


```{r fit}
set.seed(1)
fit <- multilpa(course_engagement, vars, id = "student", time = "sequence", n_profiles = 3, n_group_classes = 3, n_starts = 10, max_iter = 2000, tol = 1e-10, seed = 1)
fit
```

The estimated profiles are ordered by their overall level of activity. **Profile 1**
has the highest mean on all five indicators, **Profile 2** has intermediate means,
and **Profile 3** has the lowest means. For `attendance`, for example, the profile
means are 1.133, 0.256, and -0.940 standard deviations relative to the
corresponding course means. For `browse`, the means are 0.709, 0.387, and
-0.777. The profiles therefore primarily distinguish different levels of
engagement, although the size of the differences varies across indicators. The
contrast between profiles 1 and 2 is particularly pronounced for `attendance`.

Profile numbers are arbitrary identifiers assigned during estimation; they do
not have an intrinsic substantive meaning. For this fitted model, we refer to
profiles 1, 2, and 3 as **higher activity**, **intermediate activity**, and
**lower activity**, respectively. These labels are based on the estimated
indicator means and are used only to make the subsequent interpretation easier.
A separately estimated model may assign the same substantive profiles different
numbers, so profile labels should always be based on the estimated parameters
rather than on the numeric identifiers themselves.

Before interpreting the student classes, it is useful to check whether the
multiple random starts converged to the same solution. `get_data()` extracts
the underlying result tables from a fitted model. The `"starts"` table reports
the final log likelihood, convergence status, and variance-boundary status for
each starting value. When several starts converge to the same highest log
likelihood, this provides evidence that the selected solution is numerically
stable with respect to the starting values used. It does not, however, prove
that the solution is the global maximum.


```{r starts}
get_data(fit, "starts")
```

All ten starts converge to a log likelihood of approximately -, with no
active variance bounds. The agreement supports the numerical stability of
this fitted solution.

`plot()` compares the estimated profile means across the five activity
indicators. The vertical axis is in standard deviations from each course's
mean, and line types and point
symbols distinguish the profiles as well as colour.

```{r profiles}
plot(fit, subtitle = "Profile means, standard deviations from each course mean")
```

The three lines retain the same ordering across all five indicators. Profiles
1 and 2 are closer on page browsing than on attendance, consistent with their
means above. Profile 3 is lower throughout, so the main contrast concerns the
amount of recorded activity.

## How students differ in their profile composition

The `"profile_probabilities"` table reports $\pi_{k\mid m}$ and $\omega_m$.
The first probability describes enrolments within a student class; the second
describes the prevalence of that class among students.

```{r profile-probabilities}
get_data(fit, "profile_probabilities")
```

Student class 1 has probabilities of 0.424, 0.507, and 0.068 for profiles 1,
2, and 3. Its enrolments are predominantly in the higher and intermediate
activity profiles. Student class 2 has a more even composition: 0.325, 0.340,
and 0.335. Student class 3 has probability 0.854 of the lower-activity profile.
The estimated proportions of students in these classes are 0.304, 0.410,
and 0.286.

The probability plot compares these compositions on a common zero-to-one scale.
Each line represents a student class, and each point gives the probability of
an enrolment occupying that profile.

```{r probabilities}
plot(fit, what = "probabilities")
```

Class 2 has a nearly level line because its three profile probabilities are
similar. Class 3 rises to 0.854 at profile 3, while class 1 falls to 0.068 there.
The contrast shows how students can differ in their overall engagement pattern
while still having enrolments in several profiles.

## Reading each student's course history

The `"sequences"` table arranges each enrolment's assigned profile in course
order. Assignment selects the profile with the highest posterior probability,
so the sequence is a summary that omits classification uncertainty. The wide
format places students in rows and successive courses in columns.

```{r sequences}
get_data(fit, "sequences", format = "wide") |> head(4)
```

Student 1 occupies profile 1 in the first two courses and profile 3 in the
remaining observed courses. Student 2 occupies profile 3 throughout. Students
3 and 4 move between profiles 1 and 2. The trailing `NA` values mark
unobserved later courses.

The `"sequence_summary"` table reports the number of students, enrolments, and
observed courses in each assigned student class. `gaps` counts students missing
an occasion within their observed span; a shorter history alone is not a gap.

```{r sequence-summary}
get_data(fit, "sequence_summary")
```

The assigned classes contain 34, 42, and 30 students. Mean sequence lengths
are 13.35, 13.45, and 13.43 courses, respectively, so the differences in profile
composition occur among classes with similar observation lengths. No sequence
has an internal gap.

## Interpretation

The descriptive statistics show variation both within and between students.
The MLPA model expresses that variation as three activity profiles and three
student classes with different profile compositions. The ordered assignments
add an individual perspective: a student can change engagement profile across
courses while retaining one student-level class.

The composition and sequence summaries answer different questions. Two students
can have similar proportions of the profiles but occupy them in a different
order. The [transition guide](multilpa-transitions.html) models that order by
estimating probabilities of moving between profiles.

## Limitations

The bundled data support an illustrative analysis of the published case, rather
than a numerical replication of the study. The study used latent class analysis
followed by hidden Markov modeling; the joint MLPA here estimates a different
model. Its three student classes describe profile composition across enrolments.

The 3 × 3 specification requires evaluation before substantive interpretation.
Overlapping indicators can influence profile definitions, and modal assignments
omit uncertainty. The [evaluation guide](multilpa-evaluation.html) examines
these issues with the same data and candidate model. The model also omits
shared course effects and assumes stable profile definitions across courses.

## Reference

Saqr, M., & López-Pernas, S. (2021). The longitudinal trajectories of online
engagement over a full program. *Computers & Education, 175*, 104325.
[doi:10.1016/j.compedu.2021.104325](https://doi.org/10.1016/j.compedu.2021.104325).
