---
title: "Multilevel latent profile analysis of course engagement"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multilevel latent profile analysis of course engagement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>",
                      fig.width = 7, fig.height = 4.2, fig.align = "center",
                      dpi = 96)
options(digits = 4, width = 92)
```

Multilevel latent profile analysis (MLPA) estimates profiles of observations and
differences in their prevalence across higher-level units. These units may be
persons with repeated measurements, schools containing students, clinics
containing patients, or organizations containing employees. Observations from
the same unit can be dependent. A two-level mixture model represents this
structure through latent profiles at Level 1 and latent classes at Level 2.
The lower level describes heterogeneity among observations; the higher level
describes how their composition differs between units.

In a repeated-measure application, a person can occupy different profiles on
different occasions. Let $C_{it}$ denote the profile of person $i$ at occasion
$t$, and $G_i$ their person-level class. MLPA estimates the probability
$P(C_{it}=k\mid G_i=m)$ of profile $k$ within person class $m$. This separates
diversity across repeated observations from differences between people in the
profiles they tend to occupy.

VASSTRA makes this distinction between **state diversity** and **person
heterogeneity** explicit through sequential estimation. It first establishes a
common state space across observations, then represents each person's repeated
observations within that space. The joint MLPA model estimates the profiles and
person classes together. Both approaches allow a person to occupy several
states; they differ in how the state definitions and person-level representation
are estimated.

The `multilpa` package fits MLPA through maximum likelihood with multiple
expectation-maximization starts. This guide develops a three-profile,
three-student-class example from longitudinal course engagement. The companion
guides cover [model evaluation](multilpa-evaluation.html),
[covariates and staged estimation](multilpa-covariates.html), and
[latent transitions](multilpa-transitions.html).

```{r setup}
library(multilpa)
```

## Longitudinal course engagement

The engagement example revisits the case in
[Saqr and López-Pernas (2021)](https://doi.org/10.1016/j.compedu.2021.104325),
*The longitudinal trajectories of online engagement over a full program*.
The study distinguished three engagement states within courses and three
longer-term student trajectories. Here, the bundled synthetic `course_engagement` data
provide a reproducible version of that nested design. The example illustrates
the analytical questions; its estimates are specific to these package data.

An enrolment records one student's activity in one course. Enrolments are
Level-1 observations, and the student taking those courses is the Level-2 unit.
The same student therefore appears on several rows. `student` identifies the
nesting, `course` identifies the course, and `sequence` gives the course's
position in that student's observed history. Students can have different
numbers of enrolments. The model groups observations by student; it does not
estimate a separate course effect.

The five activity indicators describe what happened during each enrolment.
They are stored as `log1p(count)` standardized within course, so each value is
that enrolment's distance from its own course's average, in that course's
standard deviations. Activity volume is not comparable across courses, so a
student is compared with their coursemates rather than with the whole cohort.

| Variable | Definition |
|---|---|
| `browse` | Course page views. |
| `lectures` | Lecture videos viewed. |
| `forum_read` | Forum posts read. |
| `forum_post` | Forum posts written. |
| `attendance` | Days active in the course. |
| `previous_grade` | Standardized grade from the preceding course, with an entry grade at the first course. |

The indicator vector `vars` selects the five measures used to define engagement
profiles. Identifiers and `previous_grade` stay outside the measurement model.
The package accepts a character vector of column names.

```{r variables}
vars <- c("browse", "lectures", "forum_read", "forum_post", "attendance")
```

## Descriptive statistics and clustering within students

`descriptives()` summarizes the activity distributions and estimates their
intraclass correlations (ICCs) using `student` as the grouping variable. The
ICC is the between-student variance divided by total variance:

$$\mathrm{ICC}=\frac{\sigma^2_{\mathrm{between}}}{\sigma^2_{\mathrm{between}}+\sigma^2_{\mathrm{within}}}.$$

An ICC near zero indicates little between-student variation in an indicator's
mean; an ICC near one indicates little variation within students. The one-way
random-effects estimate can be negative because of sampling variation. ICCs
describe each indicator separately and do not determine how many profiles or
student classes to fit. Students could differ in their profile composition even
when their indicator means are similar.

```{r descriptives}
descriptives(course_engagement, vars = vars, id = "student")
```

The data contain 1,422 enrolments from 106 students, with no missing values.
Because the indicators are standardized within course, each has mean zero and
standard deviation near one across the pooled enrolments. Their ICCs range from
0.088 for `lectures` to 0.224 for `forum_read`: roughly a tenth to a quarter of
each indicator's variance lies between students. The repeated enrolments
therefore carry information about both course-level engagement and differences
between students.

## Three engagement profiles and three student classes

`multilpa()` estimates engagement profiles from the five indicators and student
classes from the prevalence of those profiles across each student's enrolments.
A **3 × 3 model** has three enrolment profiles and three student classes. Each
student class has its own probabilities of occupying the three profiles, and
each student can have enrolments in more than one profile.

An engagement profile has a mean and variance for each activity indicator.
These parameters have the same meaning in every student class: a high-activity
profile describes the same pattern whether it occurs frequently or occasionally
in a student's history. Student classes differ in how often their enrolments
occupy each profile.

Three profiles and three student classes are a candidate specification, which
the [evaluation guide](multilpa-evaluation.html) compares with alternatives.

The analysis uses ten random starts to check whether estimation repeatedly
reaches the same solution. `seed = 1` makes those starts reproducible and
restores the caller's random state afterward. The tighter `tol` supports the
later calculation of standard errors. `time = "sequence"` retains course
order for displaying histories; this model estimates profile composition across
courses. Transition probabilities require the separate longitudinal model.

```{r fit}
set.seed(1)
fit <- multilpa(course_engagement, vars, id = "student", time = "sequence", n_profiles = 3, n_group_classes = 3, n_starts = 10, max_iter = 2000, tol = 1e-10, seed = 1)
```

### Profile means

Printing the fit reports one row per profile: its mean on each indicator, and
how many enrolments it holds.

```{r means}
fit
```

Profile 1 has the highest mean on every indicator, profile 2 is intermediate,
and profile 3 has the lowest means. For `attendance`, their means are 1.133,
0.256, and -0.940 standard deviations from each course's average. For `browse`,
they are 0.709, 0.387, and -0.777. The profiles describe different amounts of
recorded activity, and the attendance gap between profiles 1 and 2 is wider
than the browsing gap.

The profile numbers are identifiers assigned during estimation. In this fit,
we refer to profiles 1, 2, and 3 as **higher activity**, **intermediate
activity**, and **lower activity**, respectively. A separately fitted model can
use the numbers in a different order, so its means must be checked before
naming it.

### Profile and class sizes

`get_results()` extracts tables from a fitted model. The `"counts"` table reports
effective memberships at both levels: how many enrolments fall in each profile,
and how many students in each student class. These are summed posterior
probabilities rather than modal counts, so they are not whole numbers.

```{r counts}
get_results(fit, "counts")
```

The lower-activity profile is the largest, holding 575.0 enrolments, or 40.4%.
The higher-activity profile holds 369.2, or 26.0%. Among students, class 2 is
the largest at 47.3 students, or 44.6%, with classes 1 and 3 holding 27.7 and
31.0.

### How students differ in their profile composition

The `"profile_probabilities"` table reports the probability of each profile
within each student class, alongside the prevalence of that class among
students. Reading a row therefore answers two separate questions: how a class
composes its enrolments, and how common that class is.

```{r profile-probabilities}
get_results(fit, "profile_probabilities")
```

Student class 1 has probabilities of 0.398, 0.547, and 0.055 for profiles 1, 2,
and 3. Its enrolments are predominantly higher and intermediate activity.
Student class 2 has a more even composition: 0.295, 0.388, and 0.317. Student
class 3 has probability 0.847 of the lower-activity profile. The estimated
proportions of students in these classes are 0.261, 0.446, and 0.293.

The probability plot compares these compositions on a common zero-to-one scale.
Each line represents a student class, and each point gives the probability of
an enrolment occupying that profile.

```{r probabilities}
plot(fit, what = "probabilities")
```

Class 2 has a nearly level line because its three profile probabilities are
similar. Class 3 rises to 0.847 at profile 3, while class 1 falls to 0.055
there. The contrast shows how students can differ in their overall engagement
pattern while still having enrolments in several profiles.

## Checking the estimation

The `"starts"` table reports the log likelihood, convergence status, and
variance-boundary status for each start. Repeated likelihoods support numerical
stability, although they cannot prove that a global maximum has been found.

```{r starts}
get_results(fit, "starts")
```

All ten starts converge to a log likelihood of approximately -8348.4, with no
active variance bounds. The agreement supports the numerical stability of this
fitted solution.

## Profile means as a plot

`plot()` compares the estimated profile means across the five activity
indicators. The vertical axis is in standard deviations from each course's
mean, and line types and point symbols distinguish the profiles as well as
colour.

```{r profiles}
plot(fit, subtitle = "Profile means, standard deviations from each course mean")
```

The three lines retain the same ordering across all five indicators. Profiles
1 and 2 are closer on page browsing than on attendance, consistent with their
means above. Profile 3 is lower throughout, so the main contrast concerns the
amount of recorded activity.

## Reading each student's course history

The `"sequences"` table arranges each enrolment's assigned profile in course
order. Assignment selects the profile with the highest posterior probability,
so the sequence is a summary that omits classification uncertainty. The wide
format places students in rows and successive courses in columns.

```{r sequences}
get_results(fit, "sequences", format = "wide") |> head(4)
```

Students 1 and 2 are assigned to student class 3: student 1 begins in the
intermediate profile and settles into the lower-activity profile, and student 2
occupies the lower-activity profile throughout. Students 3 and 4 belong to
class 1 and move between the higher- and intermediate-activity profiles. The
trailing `NA` values mark unobserved later courses.

The `"sequence_summary"` table reports the number of students, enrolments, and
observed courses in each assigned student class. `gaps` counts students missing
an occasion within their observed span; a shorter history alone is not a gap.

```{r sequence-summary}
get_results(fit, "sequence_summary")
```

The assigned classes contain 25, 51, and 30 students. Mean sequence lengths are
13.20, 13.51, and 13.43 courses, so the differences in profile composition
occur among classes with similar observation lengths. No sequence has an
internal gap.

## Interpretation

The descriptive statistics show variation both within and between students.
The MLPA model expresses that variation as three activity profiles and three
student classes with different profile compositions. The ordered assignments
add an individual perspective: a student can change engagement profile across
courses while retaining one student-level class.

The composition and sequence summaries answer different questions. Two students
can have similar proportions of the profiles but occupy them in a different
order. The [transition guide](multilpa-transitions.html) models that order by
estimating probabilities of moving between profiles.

## Limitations

The bundled data support an illustrative analysis of the published case, rather
than a numerical replication of the study. The study used latent class analysis
followed by hidden Markov modeling; the joint MLPA here estimates a different
model. Its three student classes describe profile composition across enrolments.

The 3 × 3 specification requires evaluation before substantive interpretation.
The default diagonal covariance treats the indicators as independent within a
profile, and a shared association between activity measures can violate that
and affect the estimated profiles. Modal assignments omit uncertainty. The
[evaluation guide](multilpa-evaluation.html) examines these issues with the
same data and candidate model.

## Reference

Saqr, M., & López-Pernas, S. (2021). The longitudinal trajectories of online
engagement over a full program. *Computers & Education, 175*, 104325.
[doi:10.1016/j.compedu.2021.104325](https://doi.org/10.1016/j.compedu.2021.104325).
