---
title: "Multilevel latent profile analysis of course engagement"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multilevel latent profile analysis of course engagement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>",
                      fig.width = 7, fig.height = 4.2, fig.align = "center",
                      dpi = 96)
options(digits = 4, width = 92)
```

Multilevel latent profile analysis (MLPA) estimates profiles of observations and
differences in their prevalence across higher-level units. These units may be
persons with repeated measurements, schools containing students, clinics
containing patients, or organizations containing employees. Observations from
the same unit can be dependent. A two-level mixture model represents this
structure through latent profiles at Level 1 and latent classes at Level 2.
The lower level describes heterogeneity among observations; the higher level
describes how their composition differs between units.

In a repeated-measure application, a person can occupy different profiles on
different occasions. Let $C_{it}$ denote the profile of person $i$ at occasion
$t$, and $G_i$ their person-level class. MLPA estimates the probability
$P(C_{it}=k\mid G_i=m)$ of profile $k$ within person class $m$. This separates
diversity across repeated observations from differences between people in the
profiles they tend to occupy.

VASSTRA makes this distinction between **state diversity** and **person
heterogeneity** explicit through sequential estimation. It first establishes a
common state space across observations, then represents each person's repeated
observations within that space. The joint MLPA model estimates the profiles and
person classes together. Both approaches allow a person to occupy several
states; they differ in how the state definitions and person-level representation
are estimated.

The `multilpa` package fits MLPA through maximum likelihood with multiple
expectation-maximization starts. This guide develops a three-profile,
three-student-class example from longitudinal course engagement. The companion
guides cover [model evaluation](multilpa-evaluation.html),
[covariates and staged estimation](multilpa-covariates.html), and
[latent transitions](multilpa-transitions.html).

```{r setup}
library(multilpa)
```

## Longitudinal course engagement

The engagement example revisits the case in
[Saqr and López-Pernas (2021)](https://doi.org/10.1016/j.compedu.2021.104325),
*The longitudinal trajectories of online engagement over a full program*.
The study distinguished three engagement states within courses and three
longer-term student trajectories. Here, the bundled synthetic `course_engagement` data
provide a reproducible version of that nested design. The example illustrates
the analytical questions; its estimates are specific to these package data.

An enrolment records one student's activity in one course. Enrolments are
Level-1 observations, and the student taking those courses is the Level-2 unit.
The same student therefore appears on several rows. `student` identifies the
nesting, `course` identifies the course, and `sequence` gives the course's
position in that student's observed history. Students can have different
numbers of enrolments. The model groups observations by student; it does not
estimate a separate course effect.

The five activity indicators describe what happened during each enrolment.
They are stored as `log1p(count)` standardized within course, so each value is
that enrolment's distance from its own course's average, in that course's
standard deviations. Activity volume is not comparable across courses, so a
student is compared with their coursemates rather than with the whole cohort.

| Variable | Definition |
|---|---|
| `browse` | Course page views. |
| `lectures` | Lecture videos viewed. |
| `forum_read` | Forum posts read. |
| `forum_post` | Forum posts written. |
| `attendance` | Days active in the course. |
| `previous_grade` | Standardized grade from the preceding course, with an entry grade at the first course. |

The indicator vector `vars` selects the five measures used to define engagement
profiles. Identifiers and `previous_grade` stay outside the measurement model.
The package accepts a character vector of column names.

```{r variables}
vars <- c("browse", "lectures", "forum_read", "forum_post", "attendance")
```

## Descriptive statistics and clustering within students

`descriptives()` summarizes the activity distributions and estimates their
intraclass correlations (ICCs) using `student` as the grouping variable. The
ICC is the between-student variance divided by total variance:

$$\mathrm{ICC}=\frac{\sigma^2_{\mathrm{between}}}{\sigma^2_{\mathrm{between}}+\sigma^2_{\mathrm{within}}}.$$

Here $\sigma^2_{\mathrm{between}}$ is variance between students and
$\sigma^2_{\mathrm{within}}$ is variance across enrolments within students.

An ICC near zero indicates little between-student variation in an indicator's
mean; an ICC near one indicates little variation within students. The one-way
random-effects estimate can be negative because of sampling variation. ICCs
describe each indicator separately and do not determine how many profiles or
student classes to fit. Students could differ in their profile composition even
when their indicator means are similar.

The descriptive table reports means, standard deviations, ranges, and ICCs for
the activity indicators, alongside sample sizes and missingness.

```{r descriptives}
descriptives(course_engagement, vars = vars, id = "student")
```

The data contain 1,422 enrolments from 106 students. The activity indicators have no
missing values. Their ICCs range from 0.093 for `lectures` to 0.225 for
`forum_read`: approximately a tenth to a quarter of their variance lies between
students. The repeated enrolments therefore contain information about both
course-level engagement and differences between students.

## Three engagement profiles and three student classes

`multilpa()` estimates engagement profiles from the five indicators and student
classes from the prevalence of those profiles across each student's enrolments.
A **3 × 3 model** has three enrolment profiles and three student classes. Each
student class has its own probabilities of occupying the three profiles, and
each student can have enrolments in more than one profile.

An engagement profile has a mean and variance for each activity indicator.
These parameters have the same meaning in every student class. For example, a
high-activity profile describes the same pattern whether it occurs frequently
or occasionally in a student's history. Student classes differ in how often
their enrolments occupy each profile.

The profile probabilities introduced above can be written as
$\pi_{k\mid m}=P(C_{it}=k\mid G_i=m)$. They range from zero to one and sum to
one across the three profiles within each student class. The class probability
$\omega_m=P(G_i=m)$ describes the proportion of students in class $m$.

The model assumes Gaussian indicator distributions within each profile,
independent students, common profile definitions across
students, and independent enrolments conditional on student class. Its default
diagonal covariance also assumes that the indicators are independent within a
profile. A shared association between activity measures can violate that
assumption and affect the estimated profiles. Three profiles and three student
classes are a candidate specification, which the
[evaluation guide](multilpa-evaluation.html) compares with alternatives.

The analysis uses ten random starts to check whether estimation repeatedly
reaches the same solution. `seed = 1` makes those starts reproducible and
restores the caller's random state afterward. The tighter `tol` supports the
later calculation of standard errors. `time = "sequence"` retains course
order for displaying histories; this model estimates profile composition across
courses. Transition probabilities require the separate longitudinal model.

```{r fit}
set.seed(1)
fit <- multilpa(course_engagement, vars, id = "student", time = "sequence", n_profiles = 3, n_group_classes = 3, n_starts = 10, max_iter = 2000, tol = 1e-10, seed = 1)
fit
```

Profile 1 has the highest mean on every indicator, profile 2 is intermediate,
and profile 3 has the lowest means. For `attendance`, their means are 1.133,
0.256, and -0.940 standard deviations from each course's average. For `browse`,
they are 0.709, 0.387, and -0.777. These profiles describe different levels of activity, with a particularly
clear attendance difference between profiles 1 and 2.

The profile numbers are identifiers assigned during estimation. In this fit,
we refer to profiles 1, 2, and 3 as **higher activity**, **intermediate activity**,
and **lower activity**, respectively. A separately fitted model can use the
numbers in a different order, so its means must be checked before naming it.

`get_data()` extracts tables from a fitted model. Its `"starts"` table reports
the log likelihood, convergence status, and variance-boundary status for each
start. Repeated likelihoods support numerical stability, although they cannot
prove that a global maximum has been found.

```{r starts}
get_data(fit, "starts")
```

All ten starts converge to a log likelihood of approximately -7526.274, with no
active variance bounds. The agreement supports the numerical stability of
this fitted solution.

`plot()` compares the estimated profile means across the five activity
indicators. The vertical axis is in standard deviations from each course's
mean, and line types and point
symbols distinguish the profiles as well as colour.

```{r profiles}
plot(fit, subtitle = "Profile means, standard deviations from each course mean")
```

The three lines retain the same ordering across all five indicators. Profiles
1 and 2 are closer on page browsing than on attendance, consistent with their
means above. Profile 3 is lower throughout, so the main contrast concerns the
amount of recorded activity.

## How students differ in their profile composition

The `"profile_probabilities"` table reports $\pi_{k\mid m}$ and $\omega_m$.
The first probability describes enrolments within a student class; the second
describes the prevalence of that class among students.

```{r profile-probabilities}
get_data(fit, "profile_probabilities")
```

Student class 1 has probabilities of 0.424, 0.507, and 0.068 for profiles 1,
2, and 3. Its enrolments are predominantly in the higher and intermediate
activity profiles. Student class 2 has a more even composition: 0.325, 0.340,
and 0.335. Student class 3 has probability 0.854 of the lower-activity profile.
The estimated proportions of students in these classes are 0.304, 0.410,
and 0.286.

The probability plot compares these compositions on a common zero-to-one scale.
Each line represents a student class, and each point gives the probability of
an enrolment occupying that profile.

```{r probabilities}
plot(fit, what = "probabilities")
```

Class 2 has a nearly level line because its three profile probabilities are
similar. Class 3 rises to 0.854 at profile 3, while class 1 falls to 0.068 there.
The contrast shows how students can differ in their overall engagement pattern
while still having enrolments in several profiles.

## Reading each student's course history

The `"sequences"` table arranges each enrolment's assigned profile in course
order. Assignment selects the profile with the highest posterior probability,
so the sequence is a summary that omits classification uncertainty. The wide
format places students in rows and successive courses in columns.

```{r sequences}
get_data(fit, "sequences", format = "wide") |> head(4)
```

Student 1 occupies profile 1 in the first two courses and profile 3 in the
remaining observed courses. Student 2 occupies profile 3 throughout. Students
3 and 4 move between profiles 1 and 2. The trailing `NA` values mark
unobserved later courses.

The `"sequence_summary"` table reports the number of students, enrolments, and
observed courses in each assigned student class. `gaps` counts students missing
an occasion within their observed span; a shorter history alone is not a gap.

```{r sequence-summary}
get_data(fit, "sequence_summary")
```

The assigned classes contain 34, 42, and 30 students. Mean sequence lengths
are 13.35, 13.45, and 13.43 courses, respectively, so the differences in profile
composition occur among classes with similar observation lengths. No sequence
has an internal gap.

## Interpretation

The descriptive statistics show variation both within and between students.
The MLPA model expresses that variation as three activity profiles and three
student classes with different profile compositions. The ordered assignments
add an individual perspective: a student can change engagement profile across
courses while retaining one student-level class.

The composition and sequence summaries answer different questions. Two students
can have similar proportions of the profiles but occupy them in a different
order. The [transition guide](multilpa-transitions.html) models that order by
estimating probabilities of moving between profiles.

## Limitations

The bundled data support an illustrative analysis of the published case, rather
than a numerical replication of the study. The study used latent class analysis
followed by hidden Markov modeling; the joint MLPA here estimates a different
model. Its three student classes describe profile composition across enrolments.

The 3 × 3 specification requires evaluation before substantive interpretation.
Overlapping indicators can influence profile definitions, and modal assignments
omit uncertainty. The [evaluation guide](multilpa-evaluation.html) examines
these issues with the same data and candidate model. The model also omits
shared course effects and assumes stable profile definitions across courses.

## Reference

Saqr, M., & López-Pernas, S. (2021). The longitudinal trajectories of online
engagement over a full program. *Computers & Education, 175*, 104325.
[doi:10.1016/j.compedu.2021.104325](https://doi.org/10.1016/j.compedu.2021.104325).
