---
title: "Multilevel latent profile analysis of course engagement"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multilevel latent profile analysis of course engagement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>",
                      fig.width = 7, fig.height = 4.2, fig.align = "center",
                      dpi = 96)
options(digits = 4, width = 92)
```

Multilevel latent profile analysis (MLPA) is used when observations are nested within higher-level units and the aim is to identify latent profiles at the observation level while also describing how the prevalence of those profiles varies across units. Examples include repeated measurements nested within persons, students nested within schools, patients nested within clinics, or employees nested within organizations. Because observations from the same higher-level unit may be dependent, MLPA extends standard latent profile analysis by modeling the data at two levels. Latent profiles are defined at Level 1, whereas latent classes at Level 2 describe differences between higher-level units in the composition of those profiles.

Consider a repeated-measures application. A person may occupy different profiles on different occasions. Let $C_{it}$ denote the profile of person $i$ at occasion $t$, and let $G_i$ denote the person's Level-2 latent class. The model estimates the conditional probability $P(C_{it}=k\mid G_i=m)$ that an observation belongs to profile $k$ given that the person belongs to class $m$. In this way, MLPA separates variation across repeated observations from variation between persons in the profiles they tend to occupy.

This distinction is also central to VaSSTra, which separates **state diversity** from **person heterogeneity** using a sequential estimation strategy. VaSSTra first establishes a common state space across all observations and then summarizes each person's repeated observations within that space. MLPA takes a different approach: the observation-level profiles and person-level classes are estimated jointly within a single model. Both approaches allow the same person to occupy several states across repeated observations, but they differ in how the states and the person-level representation are constructed.

The `multilpa` package fits MLPA models using maximum likelihood with multiple expectation-maximization starting values. In this vignette, we work through a three-profile, three-person-class example using longitudinal data on course engagement. The focus is on specifying, estimating, and interpreting the model. The companion guides cover [model evaluation](multilpa-evaluation.html), [covariates and staged estimation](multilpa-covariates.html), and [latent transitions](multilpa-transitions.html). 

```{r setup}
library(multilpa)
```

## Longitudinal course engagement


The engagement example revisits the setting studied by
[Saqr and López-Pernas (2021)](https://doi.org/10.1016/j.compedu.2021.104325),
*The longitudinal trajectories of online engagement over a full program*. The
study distinguished three engagement states within courses and three
longer-term student trajectories. Here, we use the bundled synthetic
`course_engagement` data to reproduce the same type of nested longitudinal
design. The purpose of the example is to illustrate the analytical questions
and workflow; the estimates reported below apply only to the synthetic package
data.

Each row represents one student's enrolment in one course. Enrolments are
therefore the Level-1 observations, while students are the Level-2 units within
which those observations are nested. A student who takes several courses
appears on several rows. The variable `student` identifies the Level-2 unit,
`course` identifies the course associated with each enrolment, and `sequence`
records the position of that course in the student's observed history.
Students need not contribute the same number of enrolments. The MLPA model
accounts for the grouping of enrolments within students; it does not include a
separate course-level random effect or latent structure.

Five activity indicators describe the student's engagement during each
enrolment. The indicators are stored as `log1p(count)` values standardized
within course. Each value can therefore be interpreted as the enrolment's
distance from the mean of that particular course, expressed in course-specific
standard deviations. A value above zero indicates more activity than the
course average, whereas a value below zero indicates less activity than the
course average.

This within-course standardization is important for the interpretation of the
profiles. Raw activity volumes can differ substantially across courses because
of differences in course design, duration, or expected workload. The indicators
therefore describe a student's engagement relative to other students taking
the same course rather than relative to the cohort as a whole. The profiles
identified below should consequently be interpreted as recurring patterns of
**relative engagement within courses**, not as absolute levels of activity
across the program.

| Variable | Definition |
|---|---|
| `browse` | Course page views. |
| `lectures` | Lecture videos viewed. |
| `forum_read` | Forum posts read. |
| `forum_post` | Forum posts written. |
| `attendance` | Days active in the course. |
| `previous_grade` | Standardized grade from the preceding course, with an entry grade at the first course. |

The indicator vector `vars` selects the five measures used to define engagement
profiles. Identifiers and `previous_grade` stay outside the measurement model.
The package accepts a character vector of column names.

```{r variables}
vars <- c("browse", "lectures", "forum_read", "forum_post", "attendance")
```

## Descriptive statistics and clustering within students

The `descriptives()` function summarizes the distributions of the activity indicators and estimates their intraclass correlations (ICCs), using `student` as the grouping variable. The ICC describes how much of the variation in an indicator is attributable to differences between students rather than variation across enrolments within the same student. It is defined as

$$
\mathrm{ICC}
=
\frac{\sigma^2_{\mathrm{between}}}
{\sigma^2_{\mathrm{between}}+\sigma^2_{\mathrm{within}}}.
$$


Here, $\sigma^2_{\mathrm{between}}$ is the variance between students and
$\sigma^2_{\mathrm{within}}$ is the variance across enrolments within students.

An ICC near zero indicates little between-student variation in an indicator's
average level, whereas an ICC near one indicates little variation across
enrolments within students. With the one-way random-effects estimator used
here, an estimated ICC can occasionally be negative because of sampling
variation; such estimates are generally interpreted as indicating little or no
detectable between-student clustering for that indicator.

ICCs describe the clustering of each activity indicator separately. They do not
determine how many Level-1 profiles or Level-2 student classes should be fitted.
In particular, students may differ in the **composition of their engagement
profiles** even when their average levels on the individual indicators are
similar. The ICCs are therefore descriptive measures of the multilevel
structure in the observed indicators rather than criteria for selecting the
latent structure.

The descriptive table reports means, standard deviations, ranges, and ICCs for
the activity indicators, together with sample sizes and missingness.

```{r descriptives}
descriptives(course_engagement, vars = vars, id = "student")
```

The data comprise 1,422 enrolments from 106 students, with complete observations
on all activity indicators. Because the indicators were standardized within
course, their pooled means are approximately zero and their standard deviations
are close to one. The estimated ICCs range from 0.088 for `lectures` to 0.224
for `forum_read`, indicating that about 9% to 22% of the variance in the
individual indicators is attributable to differences between students. Most
variation therefore occurs across enrolments within students, while a meaningful
share reflects between-student differences. This combination of within- and
between-student variation is the structure that the multilevel model is designed
to represent.


## Three engagement profiles and three student classes

The `multilpa()` function estimates engagement profiles from the five activity indicators and
student classes from the distribution of those profiles across each student's
enrolments. A **3 × 3 model** contains three Level-1 enrolment profiles and
three Level-2 student classes. Each student class is characterized by a
different set of probabilities for the three enrolment profiles, while an
individual student may contribute enrolments to more than one profile.

Each engagement profile is defined by a mean and variance for every activity
indicator. These profile-specific parameters are shared across student classes.
Thus, a given profile represents the same pattern of engagement regardless of
which student class it occurs in. What differs across student classes is the
relative frequency with which their enrolments occupy the profiles.

The conditional profile probabilities introduced above can be written as

$$
\pi_{k\mid m}=P(C_{it}=k\mid G_i=m),
$$

where $\pi_{k\mid m}$ is the probability that an enrolment belongs to profile
$k$ when the student belongs to class $m$. For each student class, these
probabilities range from zero to one and sum to one across the three profiles.
The Level-2 class probability

$$
\omega_m=P(G_i=m)
$$

gives the proportion of students represented by class $m$.

The model relies on several distributional and independence assumptions. Within
each profile, the activity indicators are modeled as Gaussian, and the same
profile definitions apply to all students. Students are treated as independent
Level-2 units, and enrolments within a student are conditionally independent
given the student's latent class. With the default diagonal covariance
structure, the activity indicators are additionally assumed to be independent
within each profile. Substantial residual associations between indicators after
accounting for profile membership may indicate that this assumption is too
restrictive and can affect the resulting profile solution.

The **3 × 3 model** considered here is one candidate representation of the
latent structure rather than a prespecified truth. The
[evaluation guide](multilpa-evaluation.html) shows how this model can be
compared with alternatives containing different numbers of Level-1 profiles
and Level-2 student classes.

The model is estimated from ten random starts to check whether different
initial values lead to the same likelihood solution. Setting `seed = 1` makes
these starts reproducible while restoring the caller's random state after
estimation. A tighter convergence criterion, specified through `tol`, is used
to support the subsequent calculation of standard errors. The argument
`time = "sequence"` retains the order of each student's courses so that
enrolment histories can be displayed in sequence. Course order does not enter
the present model, however: the model describes the **composition of profiles**
across a student's enrolments rather than transitions between successive
profiles. Profile-to-profile transitions are addressed by the separate
longitudinal model.



```{r fit}
set.seed(1)
fit <- multilpa(course_engagement, vars, id = "student", time = "sequence", n_profiles = 3, n_group_classes = 3, n_starts = 10, max_iter = 2000, tol = 1e-10, seed = 1)
```
### Profile means

Printing the fitted model provides a compact summary of the Level-1 profiles.
Each row corresponds to one profile and reports its estimated mean on each
activity indicator, together with the number of enrolments assigned to that
profile.

```{r means}
fit
```



Profile 1 has the highest estimated mean on all five activity indicators,
profile 2 has intermediate means, and profile 3 has the lowest means. For
`attendance`, the profile means are 1.133, 0.256, and -0.940 standard
deviations from the corresponding course mean. For `browse`, they are 0.709,
0.387, and -0.777. The profiles therefore primarily distinguish different
levels of recorded activity, although the separation varies across indicators.
For example, the difference between profiles 1 and 2 is substantially larger
for `attendance` than for `browse`.

Profile numbers are arbitrary labels assigned during estimation and have no
inherent substantive meaning. In this fitted model, we refer to profiles 1, 2,
and 3 as **higher activity**, **intermediate activity**, and **lower activity**,
respectively. The numbering may change when the model is fitted again or when a
different specification is estimated. Profile labels should therefore be based
on the estimated indicator means rather than on the profile numbers themselves.


### Profile and class sizes

The `get_data()` function extracts result tables from the fitted model. The `"counts"` table
summarizes **effective membership** at both levels: the estimated number of
enrolments associated with each Level-1 profile and the estimated number of
students associated with each Level-2 class. These counts are obtained by
summing posterior membership probabilities across observations or students,
rather than by assigning each case to its most likely class. They therefore
need not be whole numbers.


```{r counts}
get_data(fit, "counts")
```

 
The lower-activity profile is the largest, holding 575.0 enrolments, or 40.4%.
The higher-activity profile holds 369.2, or 26.0%. Among students, class 2 is
the largest at 47.3 students, or 44.6%, with classes 1 and 3 holding 27.7 and
31.0.

### How students differ in their profile composition

The `"profile_probabilities"` table reports the probability of each profile
within each student class, alongside the prevalence of that class among
students. Reading a row therefore answers two separate questions: how a class
composes its enrolments, and how common that class is.

```{r profile-probabilities}
get_data(fit, "profile_probabilities")
```

Student class 1 has probabilities of 0.398, 0.547, and 0.055 for profiles 1, 2,
and 3. Its enrolments are predominantly higher and intermediate activity.
Student class 2 has a more even composition: 0.295, 0.388, and 0.317. Student
class 3 has probability 0.847 of the lower-activity profile. The estimated
proportions of students in these classes are 0.261, 0.446, and 0.293.

The probability plot compares these compositions on a common zero-to-one scale.
Each line represents a student class, and each point gives the probability of
an enrolment occupying that profile.

```{r probabilities}
plot(fit, what = "probabilities")
```

Class 2 has a nearly level line because its three profile probabilities are
similar. Class 3 rises to 0.847 at profile 3, while class 1 falls to 0.055
there. The contrast shows how students can differ in their overall engagement
pattern while still having enrolments in several profiles.

## Checking the estimation

The `"starts"` table reports the log likelihood, convergence status, and
variance-boundary status for each start. Repeated likelihoods support numerical
stability, although they cannot prove that a global maximum has been found.

```{r starts}
get_data(fit, "starts")
```

All ten starts converge to a log likelihood of approximately -8348.4, with no
active variance bounds. The agreement supports the numerical stability of this
fitted solution.

## Profile means as a plot

`plot()` compares the estimated profile means across the five activity
indicators. The vertical axis is in standard deviations from each course's
mean, and line types and point symbols distinguish the profiles as well as
colour.

```{r profiles}
plot(fit, subtitle = "Profile means, standard deviations from each course mean")
```

The three lines retain the same ordering across all five indicators. Profiles
1 and 2 are closer on page browsing than on attendance, consistent with their
means above. Profile 3 is lower throughout, so the main contrast concerns the
amount of recorded activity.

## Reading each student's course history

The `"sequences"` table arranges each enrolment's assigned profile in course
order. Assignment selects the profile with the highest posterior probability,
so the sequence is a summary that omits classification uncertainty. The wide
format places students in rows and successive courses in columns.

```{r sequences}
get_data(fit, "sequences", format = "wide") |> head(4)
```

Students 1 and 2 are assigned to student class 3: student 1 begins in the
intermediate profile and settles into the lower-activity profile, and student 2
occupies the lower-activity profile throughout. Students 3 and 4 belong to
class 1 and move between the higher- and intermediate-activity profiles. The
trailing `NA` values mark unobserved later courses.

The `"sequence_summary"` table reports the number of students, enrolments, and
observed courses in each assigned student class. `gaps` counts students missing
an occasion within their observed span; a shorter history alone is not a gap.

```{r sequence-summary}
get_data(fit, "sequence_summary")
```

The assigned classes contain 25, 51, and 30 students. Mean sequence lengths are
13.20, 13.51, and 13.43 courses, so the differences in profile composition
occur among classes with similar observation lengths. No sequence has an
internal gap.

## Interpretation

The descriptive statistics show variation both within and between students.
The MLPA model expresses that variation as three activity profiles and three
student classes with different profile compositions. The ordered assignments
add an individual perspective: a student can change engagement profile across
courses while retaining one student-level class.

The composition and sequence summaries answer different questions. Two students
can have similar proportions of the profiles but occupy them in a different
order. The [transition guide](multilpa-transitions.html) models that order by
estimating probabilities of moving between profiles.

## Limitations

The bundled data support an illustrative analysis of the published case, rather
than a numerical replication of the study. The study used latent class analysis
followed by hidden Markov modeling; the joint MLPA here estimates a different
model. Its three student classes describe profile composition across enrolments.

The 3 × 3 specification requires evaluation before substantive interpretation.
The default diagonal covariance treats the indicators as independent within a
profile, and a shared association between activity measures can violate that
and affect the estimated profiles. Modal assignments omit uncertainty. The
[evaluation guide](multilpa-evaluation.html) examines these issues with the
same data and candidate model.

## Reference

Saqr, M., & López-Pernas, S. (2021). The longitudinal trajectories of online
engagement over a full program. *Computers & Education, 175*, 104325.
[doi:10.1016/j.compedu.2021.104325](https://doi.org/10.1016/j.compedu.2021.104325).
