---
title: "Do schools differ in the students they serve?"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Do schools differ in the students they serve?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>",
                      fig.width = 7, fig.height = 4.2, fig.align = "center",
                      dpi = 96)
options(digits = 4, width = 92)
```

```{r library}
library(multilpa)
```

Every refusal in this package is a condition with a class, so the examples below
that show one catch it by class rather than by matching message text. This
helper only wraps the message to the width of the page:

```{r refusal-helper}
refusal <- function(condition) {
  writeLines(strwrap(conditionMessage(condition), width = 88))
}
```

## The question is about schools, not students

Latent profile analysis asks *what kinds of people are there*. This package
exists for the question one level up: given that there are kinds of people, **do
the groups they sit in differ in the mix of people they contain?**

That is not a rephrasing. It is a different parameter, and a single-level model
cannot report it at all.

The package ships the data used here, so nothing below depends on a simulation
you have to read first:

```{r data}
str(school_engagement)

vars <- c("homework_hours", "participation", "interest")
```

720 students in 60 schools, three engagement indicators each. Because the data
are simulated, the column `engaged` records the kind each student was *generated*
from — the truth behind the observations, which a real study never has. We will
fit the model without it and only afterwards ask whether the model found it.

Here is why the second level is not a technicality. Across the whole sample,
**`r round(100 * mean(school_engagement$engaged))`%** of students are the engaged
kind — a tidy headline number. Now look at the schools one at a time:

```{r the-hook}
per_school <- aggregate(engaged ~ school, data = school_engagement, FUN = mean)

summary(per_school)
```

The engaged share runs from **`r round(min(per_school$engaged), 2)`** to
**`r round(max(per_school$engaged), 2)`**, and its median is
**`r round(median(per_school$engaged), 2)`**. The sample average describes no
school in the data. Any analysis that pools the students and reports one number has
averaged away the thing worth knowing.

That hook used `engaged`, which a real study would not have. The check that
works without it is the intraclass correlation — the share of each indicator's
variance lying *between* schools rather than within them:

```{r descriptives}
descriptives(school_engagement, vars = vars, id = "school")
```

Around an eighth of the variance in each indicator is between schools. That is
what makes the rest of this worth doing: had the ICCs come back near zero, the
schools would not differ, and a model built to tell them apart would have
nothing to find. It costs one line and is the step most often skipped.

## Look at the profiles before reading anything about them

The default convergence tolerance is set for the likelihood, and the likelihood
is flat near its maximum. Wald standard errors need the *score* at zero, which
is a stricter requirement, so this fit asks for `tol = 1e-10`. It costs two
extra EM iterations here; at the default the package warns that the intervals
below should not be trusted, which is exactly the kind of warning worth
believing.

```{r fit}
fit <- multilpa(school_engagement, vars, id = "school",
                n_profiles = 2, n_group_classes = 2,
                n_starts = 8, tol = 1e-10, seed = 1)
```

```{r plot-profiles}
plot(fit)
```

Two clearly separated profiles: one doing around four hours of homework with
low participation and interest, one doing around eight with both high. The
separation is what makes the rest of the analysis worth doing; if the lines
crossed or sat on top of each other, no amount of fit statistics would rescue
it.

The numbers behind that picture, one row per profile and indicator:

```{r measurement}
as.data.frame(fit)
```

The indicators are on different scales, so the raw plot partly shows that
homework hours are simply bigger numbers. Standardizing puts them on a common
footing and shows the *shape* of each profile:

```{r plot-standardized}
plot(fit, scale = "standardized")
```

The same estimates can be read three ways, and which one to use depends on the
question. Bars, with a 95% interval on each, answer *how far apart are the two
profiles on this indicator, and is the gap larger than its uncertainty*. The
intervals need the data the model was fitted to, and the fit is carrying it:

```{r plot-bars, fig.height = 4.6}
plot(fit, what = "bars")
```

A heatmap of standard deviations from each indicator's grand mean answers
*which indicators define the profiles* at a glance, and stays readable when
there are more indicators than a line plot can label:

```{r plot-heatmap, fig.height = 3.4}
plot(fit, what = "heatmap")
```

Every view is listed, so none of them has to be remembered:

```{r plot-catalogue}
multilpa_plot_types()
```

Profile labels are arbitrary. Nothing in the likelihood distinguishes "profile
1" from "profile 2", so a refit may swap them. Read the estimates against what
generated them, not against the label.

## Did it find what made the data?

This is the check simulated data exist for, and the reason `engaged` ships
beside the indicators. Asking for the assignments returns every observation with
the profile it was put in, so the truth column and the estimate are in the same
row and nothing has to be lined up by hand:

```{r recovery}
assigned <- assignments(fit, data = school_engagement)

xtabs(~ profile + engaged, data = assigned)
```

One student out of 720 on the wrong side. That is not a claim about this
package's accuracy in general — the profiles here are far apart by construction,
and real indicators overlap much more. It is a statement that the estimator
does what it says when the assumptions hold, which is the only thing a
simulation can establish.

## How many profiles, and how many kinds of school?

Both counts are chosen, not estimated, so choose them in the open. Fit the grid:

```{r enumerate}
candidates <- enumerate_classes(school_engagement, vars, id = "school",
                                n_profiles = 1:3, n_group_classes = 1:2,
                                n_starts = 4, max_iter = 10000, tol = 1e-10,
                                seed = 1)
```

```{r plot-enumeration}
plot(candidates, criterion = "bic_groups")
```

`summary()` reports which candidate each criterion prefers, rather than picking
one for you:

```{r summary-candidates}
summary(candidates)
```

They do not agree, and that is the normal case rather than a problem with the
data. The BIC family, CAIC, AWE and ICL all choose two profiles in two kinds of
school. AIC and CLC choose three profiles. This is the familiar pattern that
AIC under-penalizes in mixture models and tends to over-extract; the BIC family
is the more conservative read, and it happens to be the structure these data
were simulated from.

Note also that every sample-size-dependent criterion is reported twice, under
`groups` and under `individuals`. A two-level model has two defensible sample
sizes — 60 schools or 720 students — and neither is privileged. Pick one
convention and keep it for the whole analysis rather than choosing per table.

### Testing the candidate the criteria disagreed about

The grid kept every model it fitted, so the candidate AIC preferred can be taken
out of it by name. Nothing has to be indexed by its position in the table:

```{r candidate}
three_profiles <- candidate_fit(candidates, n_profiles = 3, n_group_classes = 2)

three_profiles
```

Two of its three profiles are the two we already have. The third pins
participation near the top of the scale with almost no variance around it, which
is what a profile looks like when it is absorbing a ceiling rather than
describing a kind of student. Is it real? A parametric bootstrap answers that
without a chi-square approximation: it simulates data from the smaller model,
refits both models to every simulated dataset, and reads the observed statistic
against the distribution that actually came out.

```{r bootstrap}
two_profiles <- candidate_fit(candidates, n_profiles = 2, n_group_classes = 2)

comparison <- bootstrap_lrt(two_profiles, three_profiles,
                            iter = 9, n_starts = 2, max_iter = 2000, seed = 1)

comparison
```

Nine replicates is far too few to decide anything, and it is what keeps this
vignette building in seconds. The Monte Carlo standard error printed beside the
p-value is the honest measure of how coarse the answer is; use several hundred
replicates in real work. Even at nine the answer is not encouraging for the
third profile: a dataset simulated from the two-profile model already produced a
statistic as large as the one observed.

```{r plot-bootstrap, fig.height = 3.8}
plot(comparison)
```

A p-value is reported at all only because every replicate was usable. When one
is not, the p-value is withheld rather than computed from the replicates that
happened to work, so the per-replicate record is part of the result:

```{r bootstrap-replicates}
as.data.frame(comparison, what = "replicates")
```

One more thing the grid will not do: hand back a candidate it never fitted. A
cell that is not in the grid is an error rather than a `NULL` that would flow
silently into whatever came next.

```{r candidate-refused}
tryCatch(candidate_fit(candidates, n_profiles = 4, n_group_classes = 2),
         multilpa_unknown_candidate = refusal)
```

The criteria for the chosen model come back in the shape a model-comparison
table is published in — one row, one column per criterion:

```{r criteria}
information_criteria(fit)
```

Those are the same column names `enumerate_classes()` uses, so this row sits
directly under the grid it was chosen from without renaming anything.

The same numbers are available one row per criterion, which is the form to use
when the convention needs to be a value you can filter on rather than half of a
column name:

```{r criteria-long}
information_criteria(fit, format = "long")
```

Lower is better on every row here, including the first: the long form reports
the deviance rather than the log likelihood, so `value` points one way
throughout. `convention` is `NA` for the three criteria that use no sample size
at all — the question does not arise for them, rather than an answer being
missing.

## The answer: the schools do differ

This is the quantity the second level exists for, and the one a single-level
model cannot report at all.

```{r prevalence}
as.data.frame(fit, what = "profile_probabilities")
```

```{r plot-probabilities}
plot(fit, what = "probabilities")
```

Two kinds of school, in roughly equal numbers, and the balance of profiles
reverses between them: the profile that is the clear majority in one kind is the
clear minority in the other. That reversal is why the lines cross. The pooled
`r round(100 * mean(school_engagement$engaged))`% from the first section is the
average of those two populations and a description of neither.

To see what pooling costs, fit the same measurement model with a single group
class — that *is* the single-level model:

```{r flat}
flat <- multilpa(school_engagement, vars, id = "school",
                 n_profiles = 2, n_group_classes = 1,
                 n_starts = 8, tol = 1e-10, seed = 1)

as.data.frame(flat, what = "profile_probabilities")
```

One row per profile, one number each, and no way to ask the question. The
two-level model also fits better by a wide margin:

```{r compare}
lmr_lrt(flat, fit)
```

The statistic is reported without a p-value on purpose. The Vermunt–Lo–Mendell–
Rubin reference distribution is not implemented here, and a chi-square
approximation to it is not a substitute, so the column is `NA` rather than
misleading.

## How much should you trust it?

Every classification diagnostic comes back from one call:

```{r diagnostics}
diagnostics(fit)
```

That is a summary of four tables, each reachable as a data frame with
`as.data.frame(quality, what = )`. The rest of this section is those tables.

**Are the classes well separated?** Entropy near one means the posterior
assigns students confidently.

```{r entropy}
entropy_table(fit)
```

Individuals separate almost perfectly; schools less so, which is expected —
twelve students is not much evidence about a school.

Entropy is one number for the whole model, which hides where the uncertainty
actually sits. These two views show the distribution behind it. The first is
each student's contribution to the entropy sum, zero being a certain
assignment; the second is the posterior probability of the profile each student
was actually assigned to, where mass piled against one is a clean separation:

```{r plot-entropy, fig.height = 3.6}
plot(fit, what = "entropy")
```

```{r plot-posteriors, fig.height = 3.6}
plot(fit, what = "posteriors")
```

Each ridge is scaled to its own maximum, which is the usual ridgeline
convention: it compares the *shapes* and not the profile sizes, so prevalence
is printed in each label instead. A ridge with a long tail towards the middle
is where a single summary number would have misled you.

**How big is each class, and how confidently?** The classification table is one
row per class at each level. It reports the class size twice — once from modal
assignment, once as the sum of the posteriors — because the two differ by
exactly the classification uncertainty entropy summarised. The odds of correct
classification put that on a scale where five is conventionally adequate:

```{r classification-table}
classification_table(fit, level = "both")
```

The `average_posterior` column above is the diagonal of a square table. The
off-diagonal is which class a unit is confused *with*, and each assigned class's
row is a distribution summing to one:

```{r average-posteriors}
average_posteriors(fit, level = "both")
```

A school assigned to group class 1 carries around eight percent of its posterior
on the other class, which is the 0.81 group entropy written out per class rather
than pooled into one number.

**Are the estimates precise?** Standard errors, with the measurement model:

```{r errors}
as.data.frame(fit, data = school_engagement)
```

Robust errors accumulate the score over independent schools rather than
students, which is the right unit when the measurement model may be
misspecified:

```{r robust}
head(parameter_inference(fit, data = school_engagement, vcov_type = "robust"), 6)
```

**Does the model's core assumption hold?** Latent profile analysis assumes the
indicators are independent *within* a profile. Where they are not, the profiles
are partly absorbing a correlation instead of describing a kind of student.

```{r residuals}
bivariate_residuals(fit, data = school_engagement)
```

`observed` and `expected` are both associations on the correlation scale, so
`residual` is their difference and the rows are comparable. All are small here;
a large one would argue for `covariance_model = "full"`.

That was one row per profile and pair, which is the default. `by = "overall"`
pools the profiles into one residual per pair, and `diagnostics()` takes the
same argument and forwards it, so the summary describes the residuals you asked
for rather than a different set under the same name:

```{r diagnostics-overall}
diagnostics(fit, by = "overall")
```

**Does the profile structure account for the nesting?** The ICCs from the first
section said roughly an eighth of each indicator's variance lay between schools.
Split the same calculation by assigned profile:

```{r descriptives-by-profile}
descriptives(fit, by = "profile")
```

The ICCs collapse towards zero. Within a profile the schools no longer differ,
which is precisely the model's claim: schools differ in the *mix* of students
they serve, not in the students themselves. The raw data agree with the model
without being asked to.

Everything printed in this section and the ones before it is also available as a
single call, which is what to run first and what not to stop at. `report()` is a
first look; every table it prints is what one of the verbs above returns, and
those are where the numbers come from when you need them as data:

```{r report}
report(fit, plots = FALSE)
```

## Relating the profiles to something the model never saw

A profile earns its keep by predicting something outside the indicators that
defined it. Two shortcuts are tempting and both distort the answer. Estimating
the outcome jointly with the measurement model lets the outcome pull the profiles
towards itself, so the profiles being reported are no longer the ones that were
inspected. Assigning each student to their modal profile and comparing group
means instead throws away the fact that some assignments are wrong, and
attenuates every difference it then reports. The three-step family does neither.

These data carry no outcome, so here is one, drawn with a twelve-point gap
between the two kinds of student, and a covariate drawn to predict which kind a
student is. Both are simulated, both are seeded, and neither was available to
the model that was fitted above:

```{r outcome}
set.seed(2024)
with_extra <- transform(school_engagement,
                        exam = round(55 + 12 * engaged + rnorm(720, sd = 8), 1),
                        prior_grade = round(rnorm(720, 2 * engaged), 2))

str(with_extra)
```

The correction starts from how often a student who truly belongs to one profile
is assigned to the other. This is not the average-posterior table from the last
section: that conditioned on the profile a student was *assigned* to, this
conditions on the profile they are actually in, and the correction needs the
second:

```{r classification-errors}
classification_errors(fit)
```

Those probabilities become one weight per student per profile. The weights are
not probabilities: a student assigned to one profile carries a small *negative*
weight for the other, which is the arithmetic that undoes the attenuation.

```{r bch}
head(bch_weights(fit), 4)
```

`three_step()` applies them. Asked for a contrast, it reports the difference
between the profiles rather than a mean per profile, because a difference is the
quantity that has a null worth testing:

```{r three-step}
three_step(fit, data = with_extra, outcome = "exam", contrast = "pairs")
```

Close to the twelve points the outcome was drawn with. So is the naive modal
comparison, shown for contrast:

```{r three-step-modal}
three_step(fit, data = with_extra, outcome = "exam", contrast = "pairs",
           method = "modal")
```

The two agree here, and the reason is in the table two chunks above: under a
half of one percent of students are misclassified, so there is almost nothing
for the correction to undo. That is the useful lesson rather than a
disappointing one. The gap between these two rows is a function of how bad the
classification is, so `classification_errors()` is what tells you in advance
whether the correction will matter.

`three_step()` asks what a profile predicts. `r3step()` asks the reverse: what
predicts the profile. It regresses membership on the covariates with the same
correction applied, reporting log odds against the last profile as reference:

```{r r3step}
r3step(fit, data = with_extra, covariates = "prior_grade")
```

Profile 1 is the low-engagement profile, so a negative slope says a higher prior
grade makes the low profile less likely, which is the direction the covariate
was drawn with. Sixty schools are enough to ask for a sandwich clustered on
schools instead of the model-based information:

```{r r3step-robust}
r3step(fit, data = with_extra, covariates = "prior_grade",
       vcov_type = "robust")
```

### A standard error that cannot exist is refused

A cluster-robust variance is a sum of one contribution per independent group,
and those contributions sum to zero at the estimate — that sum *is* the
condition the estimate solves. With too few groups the result is singular, and
with one group it is exactly zero. Neither is a small standard error. Fit the
same model to two schools and ask for the clustered version:

```{r refuse-clusters}
two_schools <- subset(with_extra, school <= 2)

tiny <- multilpa(two_schools, vars, id = "school", n_profiles = 2,
                 n_group_classes = 1, n_starts = 4, seed = 1)

tryCatch(three_step(tiny, data = two_schools, outcome = "exam"),
         multilpa_too_few_groups = refusal)
```

The variance that *can* exist here is the unclustered one, and it
has to be asked for by name, so that a report of these numbers has to say the
nesting was ignored:

```{r independent}
three_step(tiny, data = two_schools, outcome = "exam",
           vcov_type = "independent")
```

### The one-step alternative

The three-step methods hold the measurement fixed on purpose. The one-step model
estimates the covariate effect and the profiles together, so the covariate
*does* help shape the profiles. That is the trade, and the package offers both
rather than naming one correct:

```{r covariates}
covariate_fit <- fit_covariates(with_extra, vars, id = "school",
                                n_profiles = 2, n_group_classes = 2,
                                profile_covariates = "prior_grade",
                                n_starts = 2, seed = 1)

as.data.frame(covariate_fit, what = "coefficients")
```

The slope has the same sign and nearly the same size as `r3step()` gave it,
which is what should happen when the profiles are this well separated: there is
little for the covariate to move. A covariate fit is a fit like any other, so
the same inference verb works on it. The membership coefficients are the last
rows it returns, after the measurement model:

```{r covariate-inference}
tail(parameter_inference(covariate_fit), 4)
```

The slope's standard error is what the one-step model buys over the coefficient
table above, which carries the estimate alone.

## Building the same model a different way

### Deciding the measurement before asking about groups

Everything so far estimated the measurement model and the group structure at the
same time, which means adding a group class can move the profiles it was meant
to describe. Staging forbids that: the measurement is fitted once, on its own,
and then held while the group classes are estimated.

```{r staged}
staged <- fit_staged(school_engagement, vars, id = "school", n_profiles = 2,
                     n_group_classes = 2, n_starts = 6, tol = 1e-10, seed = 1)

as.data.frame(staged, what = "stages")
```

One row per stage, and two parameter counts. `parameters` counts what that stage
estimated, which is what to use when comparing staged fits with one another;
`parameters_with_measurement` adds the held blocks back, which is what to use
against a jointly estimated fit. First-stage uncertainty is not propagated into
the second, so everything the second stage reports is conditional on a
measurement solution treated as known.

The staged fit is a `multilpa` object, so the answer comes back from the same
accessor as before:

```{r staged-probabilities}
as.data.frame(staged, what = "profile_probabilities")
```

What a stage hands to the next one is a starting-value object. It is also what
to keep when a solution should be reproduced exactly later, or handed to a fit
on new data:

```{r starts}
starting_values(fit, what = "measurement")
```

Its blocks are indexed by position, because that is what the fitter expects; the
labelled view is the accessor rather than the list:

```{r starts-table}
as.data.frame(starting_values(fit))
```

A staged fit cannot be bootstrapped against a freely estimated one. One holds
its measurement and the other does not, so the smaller model is not a
restriction of the larger and the likelihood-ratio statistic would compare two
parameter counts that do not correspond:

```{r refuse-nesting}
tryCatch(bootstrap_lrt(flat, staged, iter = 9, seed = 1),
         multilpa_bad_nesting = refusal)
```

### When schools differ by degree rather than by kind

Two latent classes of school is one answer to *how do schools differ*. The other
is that they do not fall into kinds at all: each school simply sits a little
higher or lower on every indicator. That is a random intercept, and it is a
different model rather than a variant of this one.

```{r random-intercept}
shifted <- fit_random_intercept(school_engagement, vars, id = "school",
                                n_profiles = 2, n_starts = 2, seed = 1)

summary(shifted)
```

The estimated random-intercept standard deviation is about two hundredths of a
point on indicators that span roughly ten: the model was given the chance to say
schools sit at different levels and found essentially nothing. Its BIC is worse
than the two-class model's, and the whole of this vignette says why. The schools
in these data do not differ in *where* their students sit, they differ in *which
students* they contain, and only a model with a second latent variable can put
that difference anywhere.

The posterior intercept for each school is a table like everything else:

```{r random-intercepts}
head(as.data.frame(shifted, what = "random_intercepts"))
```

This family provides no standard errors either, and asking for them raises a
classed condition rather than returning something that looks like one. The last
section shows that refusal on another family.

## When students are measured more than once

If the same people are observed on several occasions, the question changes from
*what kinds are there* to *who moves between them*. That needs a different
dataset — in `school_engagement`, `term` indexes twelve different students within
a school, so a "transition" would be between two different people and would mean
nothing.

The package ships one with the right shape:

```{r panel}
str(engagement_panel)
```

120 students, four waves each, and again a truth column: `state` is the state
each row was generated from, with students staying put from one wave to the next
with probability 0.85.

The first thing to do with repeated measures is to look at them in order. Fitted
with `time =`, the model stores the ordering without using it, and the
assignments can then be read back as sequences. Here the grouping variable is
the student rather than the school, so a latent group class is a kind of
*trajectory*:

```{r panel-fit}
panel_fit <- multilpa(engagement_panel, c("homework_hours", "participation"),
                      id = "student", time = "wave", n_profiles = 2,
                      n_group_classes = 2, n_starts = 6, seed = 1)

head(sequences(panel_fit))
```

One row per observation is the shape to compute on. One row per student, one
column per wave, is the shape to look at, and the shape sequence-analysis
packages expect:

```{r sequences-wide}
head(sequences(panel_fit, format = "wide"))
```

```{r plot-sequences, fig.height = 4.6}
plot(panel_fit, what = "sequences")
```

Before reading anything into the picture, check that the two kinds of trajectory
are built on comparable amounts of data. A class whose students are
systematically shorter or gappier is a different finding from a class whose
students simply differ in profile mix, and the prevalence table cannot show it:

```{r sequence-summary}
sequence_summary(panel_fit)
```

Every student is observed at every wave and no sequence has a hole in it, so the
two classes are comparing like with like.

The model above still ignores the ordering it stored. The model that uses it
estimates where students move:

```{r transitions}
moves <- fit_transitions(engagement_panel, c("homework_hours", "participation"),
                         id = "student", time = "wave", n_profiles = 2,
                         n_starts = 6, seed = 1)

transitions(moves, stable = TRUE)
```

`stable = TRUE` marks the diagonal — staying in the same profile — and it
recovers the 0.85 it was generated with to within a few points. Close, not
exact, which is what four waves of 120 students supports.

## What the package refuses

Three refusals have already appeared in this vignette: a cluster-robust standard
error with too few groups, a bootstrap comparison whose two models are not
nested, and a candidate that was never enumerated. Here is the fourth. A model
family for which no standard error is implemented says so, rather than returning
a number that would look like one:

```{r refuses}
tryCatch(vcov(moves), multilpa_no_inference = refusal)
```

Every refusal in the package is a classed condition, which is why every
`tryCatch()` above names the class it is catching instead of matching message
text. `?"multilpa-conditions"` lists all of them.
