---
title: "Two-level latent profiles, end to end"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Two-level latent profiles, end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
options(digits = 4, width = 100)
```

Every table below is printed exactly as the package returns it. Nothing is
reshaped, subsetted or reformatted on the way out: if a verb does not already
return what a reader needs, that is a defect in the verb rather than something
for the caller to patch afterwards.

```{r library}
library(multilpa)
```

## Data with a known answer

Students are nested in schools. Each student belongs to one of two engagement
profiles, and each school belongs to one of two classes that differ in how its
students are distributed across those profiles. Simulating it means the truth
is known, so every estimate below can be read against something.

```{r simulate}
set.seed(2024)

n_schools <- 60
students_per_school <- 12

school_class <- rep(1:2, length.out = n_schools)
high_engagement_share <- ifelse(school_class == 1L, 0.75, 0.25)

students <- data.frame(
  school = rep(seq_len(n_schools), each = students_per_school),
  term = rep(seq_len(students_per_school), times = n_schools)
)
students$true_school_class <- rep(school_class, each = students_per_school)
students$true_profile <- 1L + as.integer(
  runif(nrow(students)) > rep(high_engagement_share, each = students_per_school)
)

engaged <- students$true_profile == 1L
students$reading    <- rnorm(nrow(students), ifelse(engaged,  1.2, -1.2), 0.8)
students$maths      <- rnorm(nrow(students), ifelse(engaged,  0.9, -0.9), 0.8)
students$engagement <- rnorm(nrow(students), ifelse(engaged,  1.5, -1.5), 0.9)
students$passed     <- ifelse(runif(nrow(students)) < ifelse(engaged, 0.85, 0.3),
                              "yes", "no")

str(students)
```

True profile prevalence is 0.75 and 0.25 in the two kinds of school, and the
measurement means are 1.2, 0.9 and 1.5 against their negatives.

## Fitting

```{r fit}
indicators <- c("reading", "maths", "engagement")

fit <- multilpa(students, indicators, group = "school",
                n_profiles = 2, n_group_classes = 2,
                n_starts = 10, seed = 1)
fit
```

## The measurement model

`as.data.frame()` gives the Gaussian measurement model, one row per profile and
indicator.

```{r measurement}
as.data.frame(fit)
```

Profile labels are arbitrary: nothing in the likelihood distinguishes "profile 1"
from "profile 2", so a fit may recover the simulated profiles in either order.
Read the estimates against the values that generated them rather than against
the labels, and align labels explicitly before comparing two fits.

Passing the fitting data adds a standard error beside every estimate, so the
measurement model reads complete from one call rather than a join against
`parameter_inference()`.

```{r measurement-errors}
as.data.frame(fit, data = students)
```

## Where the second level shows up

Profile prevalence differs by school class. This is the quantity the second
level exists to estimate, and the one a single-level model cannot report at all.

```{r prevalence}
as.data.frame(fit, what = "profile_probabilities")
```

## Fit statistics

Every information criterion the package computes, in one table. The five that
depend on a sample size appear under both conventions, because a two-level model
has two defensible sample sizes and neither is privileged. `aic` and `kic` depend
on none. `clc` depends on no sample size either, but its convention selects which
level's classification uncertainty it penalizes.

```{r criteria}
information_criteria(fit)
```

## Classification quality

```{r classification}
classification_table(fit, level = "both")
```

```{r entropy}
entropy_table(fit)
```

## Standard errors

```{r inference}
parameter_inference(fit, data = students)
```

Robust errors accumulate the score over independent groups, which is the
appropriate choice when the measurement model may be misspecified.

```{r robust}
parameter_inference(fit, data = students, vcov_type = "robust")
```

## Is two the right number of classes?

```{r enumerate}
candidates <- enumerate_classes(students, indicators, group = "school",
                                profiles = 1:3, group_classes = 1:2,
                                n_starts = 4, seed = 1)
as.data.frame(candidates)
```

## Does conditional independence hold?

Bivariate residuals report where the assumption fails. The p-values are
approximate and unadjusted; a multiplicity correction belongs on top of them
when several pairs are compared.

```{r residuals}
bivariate_residuals(fit, data = students)
```

## Relating classes to variables that did not define them

The classification carries error, and the three-step verbs carry it forward
rather than pretending modal assignment is the truth.

```{r errors}
classification_errors(fit)
```

```{r bch}
bch_weights(fit)
```

## Mixed indicators

Naming a subset of the indicators as categorical fits a mixed measurement model.
Everything above works on it unchanged.

```{r mixed}
mixed_fit <- multilpa(students, c(indicators, "passed"), group = "school",
                      n_profiles = 2, n_group_classes = 2,
                      categorical = "passed", n_starts = 6, seed = 1)
as.data.frame(mixed_fit, what = "responses", data = students)
```

## Membership covariates

```{r covariates}
students$prior_attainment <- rnorm(nrow(students))

covariate_fit <- fit_covariates(students, indicators, group = "school",
                                n_profiles = 2, n_group_classes = 2,
                                profile_covariates = "prior_attainment",
                                n_starts = 6, seed = 1)
as.data.frame(covariate_fit, what = "coefficients")
```

## Staged estimation

`fit_staged()` estimates the measurement model on its own, then the group-class
structure with that measurement held fixed, so adding group classes cannot move
the profiles they are meant to describe.

```{r staged}
staged <- fit_staged(students, indicators, group = "school",
                     n_profiles = 2, n_group_classes = 2,
                     n_starts = 6, seed = 1)
as.data.frame(staged, what = "stages")
```

Both parameter counts are reported. Compare staged fits with one another using
`parameters`; compare a staged fit against a jointly estimated one using
`parameters_with_measurement`, remembering that the staged likelihood is not the
joint maximum and first-stage uncertainty is not propagated.

## Profiles over time

Sequences and transitions ask about the same unit measured repeatedly, which the
data above are not: there `term` merely indexes students within a school, and a
move from one student to the next means nothing. So this section uses a second
dataset, where the same students are measured across six terms and mostly stay
where they are.

```{r panel}
set.seed(11)

n_students <- 80
n_terms <- 6
persistence <- 0.8

paths <- lapply(seq_len(n_students), function(student) {
  state <- Reduce(function(previous, step) {
    if (runif(1) < persistence) previous else 3L - previous
  }, seq_len(n_terms - 1L), init = 1L + student %% 2L, accumulate = TRUE)
  data.frame(student = student, term = seq_len(n_terms), state = unlist(state))
})

panel <- do.call(rbind, paths)
engaged_now <- panel$state == 1L
panel$reading    <- rnorm(nrow(panel), ifelse(engaged_now,  1.2, -1.2), 0.7)
panel$maths      <- rnorm(nrow(panel), ifelse(engaged_now,  0.9, -0.9), 0.7)
panel$engagement <- rnorm(nrow(panel), ifelse(engaged_now,  1.5, -1.5), 0.7)

str(panel)
```

Fitting the cross-sectional model with `time` records the ordering without using
it, so the assignments can be read back in sequence.

```{r sequences}
timed <- multilpa(panel, indicators, group = "student",
                  n_profiles = 2, n_group_classes = 1,
                  time = "term", n_starts = 6, seed = 1)
head(sequences(timed))
```

```{r sequence-summary}
sequence_summary(timed)
```

## Latent transitions

`sequences()` reports where the model put each observation. It does not estimate
how observations move. `fit_transitions()` does: the diagonal below is the
probability of staying, against the 0.8 that generated the data.

```{r transitions}
moves <- fit_transitions(panel, indicators, group = "student",
                         n_profiles = 2, time = "term",
                         n_starts = 6, seed = 1)
transitions(moves)
```

```{r transitions-initial}
as.data.frame(moves, what = "initial")
```

## What the package refuses

Standard errors are not available for every model family. Where they are not,
the request raises a condition with a stable class rather than returning a
number that would be wrong. `?"multilpa-conditions"` lists all of them.

```{r refusal}
tryCatch(parameter_inference(moves, students),
         multilpa_no_inference = function(condition) conditionMessage(condition))
```

## Session

```{r session}
sessionInfo()
```
