---
title: "Do students differ in the courses they engage in?"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Do students differ in the courses they engage in?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>",
                      fig.width = 7, fig.height = 4.2, fig.align = "center",
                      dpi = 96)
options(digits = 4, width = 92)
```

```{r library}
library(multilpa)
```

Every refusal in this package is a condition with a class, so the examples below
that show one catch it by class rather than by matching message text. This
helper only wraps the message to the width of the page:

```{r refusal-helper}
refusal <- function(condition) {
  writeLines(strwrap(conditionMessage(condition), width = 88))
}
```

## The question is about students, not enrolments

Latent profile analysis asks *what kinds of thing are there*. This package
exists for the question one level up: given that there are kinds, **do the units
they are nested in differ in the mix of kinds they contain?**

That is not a rephrasing. It is a different parameter, and a single-level model
cannot report it at all.

The package ships one dataset, and the whole of this vignette runs on it — the
two-level model, the sequence model and the covariate model all on the same
rows, so nothing below depends on a simulation you have to read first:

```{r data}
str(course_engagement)

activity <- c("browse", "lectures", "forum_read", "forum_post", "attendance")
```

One row is one student's enrolment in one course: 1422 enrolments, 106 students,
32 courses, up to fifteen courses each. Five activity indicators, all
`log1p` counts of what the student did in that course. Because the data are
simulated, `engagement` records the pattern each enrolment was *generated* from
— the truth behind the observations, which a real study never has. We will fit
the model without it and only afterwards ask whether the model found it.

Here is why the second level is not a technicality. Pooled across the sample:

```{r the-hook-pooled}
xtabs(~ engagement, data = course_engagement)
```

837 engaged enrolments against 585 disengaged — a tidy headline number, a bit
under three in five. Now look at the students one at a time:

```{r the-hook}
engaged_share <- aggregate(engaged ~ student,
                           data = transform(course_engagement,
                                            engaged = as.integer(engagement == "engaged")),
                           FUN = mean)

summary(engaged_share)
```

The engaged share runs the whole way from 0 to 1, with a quarter of students
below 0.29 and a quarter above 0.86. The sample average, 0.589, describes almost
no student in the data. Any analysis that pools the enrolments and reports one
number has averaged away the thing worth knowing.

That hook used `engagement`, which a real study would not have. The check that
works without it is the intraclass correlation — the share of each indicator's
variance lying *between* students rather than within them:

```{r descriptives}
descriptives(course_engagement, vars = activity, id = "student")
```

Between a tenth and a quarter of the variance in each indicator is between
students. That is what makes the rest of this worth doing: had the ICCs come
back near zero, the students would not differ, and a model built to tell them
apart would have nothing to find. It costs one line and is the step most often
skipped.

## Look at the profiles before reading anything about them

The default convergence tolerance is set for the likelihood, and the likelihood
is flat near its maximum. Wald standard errors need the *score* at zero, which
is a stricter requirement, so this fit asks for `tol = 1e-10`. It costs a couple
of extra EM iterations here; at the default the package warns that the intervals
below should not be trusted, which is exactly the kind of warning worth
believing.

```{r fit}
fit <- multilpa(course_engagement, activity, id = "student",
                n_profiles = 2, n_group_classes = 2,
                n_starts = 10, tol = 1e-10, seed = 1)

fit
```

A mixture likelihood has local maxima, so a single start is a guess. Ten were
run, and what each of them did is part of the fit rather than something to take
on trust:

```{r starts-diagnostics}
as.data.frame(fit, what = "starts")
```

All ten converged, none of them on a boundary, and all ten reached the same
likelihood — the minimum evidence that this one is the maximum. `summary()` and
`report()` print the same check condensed into a single replication count.

```{r plot-profiles}
plot(fit)
```

Two clearly separated profiles: one doing little of everything, one doing a lot
of everything. The separation is what makes the rest of the analysis worth
doing; if the lines crossed or sat on top of each other, no amount of fit
statistics would rescue it.

The numbers behind that picture, one row per profile and indicator:

```{r measurement}
as.data.frame(fit)
```

The indicators are all log counts, so the raw plot is already close to a common
footing. Standardizing removes the last of the scale differences and shows the
*shape* of each profile:

```{r plot-standardized}
plot(fit, scale = "standardized")
```

The same estimates can be read three ways, and which one to use depends on the
question. Bars, with a 95% interval on each, answer *how far apart are the two
profiles on this indicator, and is the gap larger than its uncertainty*. The
intervals need the data the model was fitted to, and the fit is carrying it:

```{r plot-bars, fig.height = 4.6}
plot(fit, what = "bars")
```

A heatmap of standard deviations from each indicator's grand mean answers
*which indicators define the profiles* at a glance, and stays readable when
there are more indicators than a line plot can label:

```{r plot-heatmap, fig.height = 3.4}
plot(fit, what = "heatmap")
```

Every view is listed, so none of them has to be remembered:

```{r plot-catalogue}
multilpa_plot_types()
```

Profile labels are arbitrary. Nothing in the likelihood distinguishes "profile
1" from "profile 2", so a refit may swap them. Read the estimates against what
generated them, not against the label.

## Did it find what made the data?

This is the check simulated data exist for, and the reason `engagement` ships
beside the indicators. Asking for the assignments returns every observation with
the profile it was put in, so the truth column and the estimate are in the same
row and nothing has to be lined up by hand:

```{r recovery}
xtabs(~ profile + engagement, data = assignments(fit, data = course_engagement))
```

Twenty-four enrolments out of 1422 on the wrong side. The truth column is named
`engagement` rather than `profile` precisely so that this is one call: a dataset
carrying a column called `profile` would collide with the one `assignments()`
adds, and the verb would refuse rather than overwrite it.

The same call carries the group classes, so the second level can be checked in
the same way. Here the labels *have* switched — group class 1 is the kind of
student the simulation called wavering — which is the arbitrariness warned about
above, seen in the wild:

```{r recovery-groups}
xtabs(~ group_class + student_type, data = assignments(fit, data = course_engagement))
```

Neither table is a claim about this package's accuracy in general — the profiles
here are far apart by construction, and real indicators overlap much more. They
are a statement that the estimator does what it says when the assumptions hold,
which is the only thing a simulation can establish.

## How many profiles, and how many kinds of student?

Both counts are chosen, not estimated, so choose them in the open. Fit the grid:

```{r enumerate}
candidates <- enumerate_classes(course_engagement, activity, id = "student",
                                n_profiles = 1:3, n_group_classes = 1:2,
                                n_starts = 2, max_iter = 1000, seed = 1)
```

```{r plot-enumeration}
plot(candidates, criterion = "bic_groups")
```

`summary()` reports which candidate each criterion prefers, rather than picking
one for you:

```{r summary-candidates}
summary(candidates)
```

Almost every criterion wants three profiles in two kinds of student; only the
three individual-level criteria that penalise entropy heavily — AWE, ICL and CLC
computed on 1422 individuals — hold out for two. Hold that disagreement; the
diagnostics section settles it, and not by picking a favourite criterion.

Note also that every sample-size-dependent criterion is reported twice, under
`groups` and under `individuals`. A two-level model has two defensible sample
sizes — 106 students or 1422 enrolments — and neither is privileged. Pick one
convention and keep it for the whole analysis rather than choosing per table.

The grid also records the cell it could not fit at all, rather than dropping it:
one profile in two kinds of student is not identifiable, because with a single
profile there is no mix for the kinds to differ in.

```{r enumeration-table}
as.data.frame(candidates)
```

### Looking at the candidate the criteria preferred

The grid kept every model it fitted, so the candidate can be taken out of it by
name. Nothing has to be indexed by its position in the table:

```{r candidate}
three_profiles <- candidate_fit(candidates, n_profiles = 3, n_group_classes = 2)

as.data.frame(three_profiles)
```

Look at what the third profile bought. Profiles 1 and 2 are near-copies of each
other on `browse`, `lectures`, `forum_read` and `forum_post` — 4.02 against
3.79, 4.08 against 3.72 — and separate almost entirely on `attendance`, 3.74
against 2.90, with unusually small variances there. A profile that exists to
split one indicator is not usually a kind of enrolment; it is usually something
the measurement model is failing to represent. The entropy in the grid above
falls from 0.946 to 0.800 when it is added, which is what happens when a class
has to be split with a coin.

One more thing the grid will not do: hand back a candidate it never fitted. A
cell that is not in the grid is an error rather than a `NULL` that would flow
silently into whatever came next.

```{r candidate-refused}
tryCatch(candidate_fit(candidates, n_profiles = 4, n_group_classes = 2),
         multilpa_unknown_candidate = refusal)
```

The criteria for the chosen model come back in the shape a model-comparison
table is published in — one row, one column per criterion:

```{r criteria}
information_criteria(fit)
```

Those are the same column names `enumerate_classes()` uses, so this row sits
directly under the grid it was chosen from without renaming anything.

The same numbers are available one row per criterion, which is the form to use
when the convention needs to be a value you can filter on rather than half of a
column name:

```{r criteria-long}
information_criteria(fit, format = "long")
```

Lower is better on every row here, including the first: the long form reports
the deviance rather than the log likelihood, so `value` points one way
throughout. `convention` is `NA` for the three criteria that use no sample size
at all — the question does not arise for them, rather than an answer being
missing.

## The answer: the students do differ

This is the quantity the second level exists for, and the one a single-level
model cannot report at all.

```{r prevalence}
as.data.frame(fit, what = "profile_probabilities")
```

```{r plot-probabilities}
plot(fit, what = "probabilities")
```

Two kinds of student, about a third and two thirds of the cohort, and the
balance of profiles reverses between them: the profile that is the clear
majority in one kind is the clear minority in the other. That reversal is why
the lines cross. The pooled 0.589 from the first section is the average of those
two populations and a description of neither.

To see what pooling costs, fit the same measurement model with a single group
class — that *is* the single-level model:

```{r flat}
flat <- multilpa(course_engagement, activity, id = "student",
                 n_profiles = 2, n_group_classes = 1,
                 n_starts = 10, tol = 1e-10, seed = 1)

as.data.frame(flat, what = "profile_probabilities")
```

One row per profile, one number each, and no way to ask the question. The
two-level model also fits better by a wide margin:

```{r compare}
lmr_lrt(flat, fit)
```

The statistic is reported without a p-value on purpose. The Vermunt–Lo–Mendell–
Rubin reference distribution is not implemented here, and a chi-square
approximation to it is not a substitute, so the column is `NA` rather than
misleading.

A parametric bootstrap gives the same comparison a reference distribution that
is not assumed. It simulates data from the single-class model, refits both
models to every simulated dataset, and reads the observed statistic against the
distribution that actually came out:

```{r bootstrap}
comparison <- bootstrap_lrt(flat, fit, iter = 9, n_starts = 2,
                            max_iter = 2000, seed = 1)

comparison
```

```{r plot-bootstrap, fig.height = 3.8}
plot(comparison)
```

Read the p-value and the Monte Carlo standard error together. Nine replicates
cannot produce a p-value below 0.1 — one tenth is the smallest value the
arithmetic allows — so 0.1 here is the floor of the procedure and not a measure
of the evidence. The evidence is in the plot: the observed statistic is 360, and
the largest statistic any of the nine simulated datasets produced is under 3.
Nine replicates is what keeps this vignette building in seconds; use several
hundred in real work, where the floor is low enough to be informative.

A p-value is reported at all only because every replicate was usable. When one
is not, the p-value is withheld rather than computed from the replicates that
happened to work, so the per-replicate record is part of the result:

```{r bootstrap-replicates}
as.data.frame(comparison, what = "replicates")
```

## How much should you trust it?

Every classification diagnostic comes back from one call:

```{r diagnostics}
diagnostics(fit)
```

That is a summary of four tables, each reachable as a data frame with
`as.data.frame(quality, what = )`. The rest of this section is those tables, and
the last of them will turn out to say something the summary above already
flagged.

**Are the classes well separated?** Entropy near one means the posterior
assigns units confidently.

```{r entropy}
entropy_table(fit)
```

Enrolments separate almost perfectly, students very nearly so — thirteen
enrolments is a good deal of evidence about a student.

Entropy is one number for the whole model, which hides where the uncertainty
actually sits. These two views show the distribution behind it. The first is
each enrolment's contribution to the entropy sum, zero being a certain
assignment; the second is the posterior probability of the profile each
enrolment was actually assigned to, where mass piled against one is a clean
separation:

```{r plot-entropy, fig.height = 3.6}
plot(fit, what = "entropy")
```

```{r plot-posteriors, fig.height = 3.6}
plot(fit, what = "posteriors")
```

Each ridge is scaled to its own maximum, which is the usual ridgeline
convention: it compares the *shapes* and not the profile sizes, so prevalence
is printed in each label instead. A ridge with a long tail towards the middle
is where a single summary number would have misled you.

**How big is each class, and how confidently?** The classification table is one
row per class at each level. It reports the class size twice — once from modal
assignment, once as the sum of the posteriors — because the two differ by
exactly the classification uncertainty entropy summarised. The odds of correct
classification put that on a scale where five is conventionally adequate:

```{r classification-table}
classification_table(fit, level = "both")
```

The `average_posterior` column above is the diagonal of a square table. The
off-diagonal is which class a unit is confused *with*, and each assigned class's
row is a distribution summing to one:

```{r average-posteriors}
average_posteriors(fit, level = "both")
```

A student assigned to group class 1 carries about three percent of their
posterior on the other class, which is the 0.938 group entropy written out per
class rather than pooled into one number.

**Are the estimates precise?** Standard errors, with the measurement model:

```{r errors}
as.data.frame(fit, data = course_engagement)
```

Robust errors accumulate the score over independent students rather than
enrolments, which is the right unit when the measurement model may be
misspecified:

```{r robust}
head(parameter_inference(fit, vcov_type = "robust"), 6)
```

### Does the model's core assumption hold? (No, and that is the finding)

Latent profile analysis with a diagonal residual covariance assumes the
indicators are independent *within* a profile. Where they are not, the profiles
are partly absorbing a correlation instead of describing a kind of enrolment.
`by = "overall"` pools the profiles into one residual per pair:

```{r residuals}
bivariate_residuals(fit, by = "overall")
```

`observed` and `expected` are both associations on the correlation scale, so
`residual` is their difference and the rows are comparable. Four of the ten
pairs are large and unambiguous, and every one of the four involves
`attendance`. That is not noise: a student is recorded as active on a day
*because* they clicked something, so `attendance` shares variance with each
click measure over and above the profile they are both in.

`diagnostics()` takes the same argument and forwards it, so the summary
describes the residuals you asked for rather than a different set under the same
name:

```{r diagnostics-overall}
diagnostics(fit, by = "overall")
```

The fix is to let the residuals correlate, which is what
`covariance_model = "full"` estimates:

```{r corrected}
corrected <- multilpa(course_engagement, activity, id = "student",
                      n_profiles = 2, n_group_classes = 2,
                      covariance_model = "full",
                      n_starts = 10, tol = 1e-10, seed = 1)

bivariate_residuals(corrected, by = "overall")
```

The residuals are gone — a full residual covariance reproduces every pairwise
association exactly, which is why this table is a check that the model was
fitted and not a test that it is true. The test is whether the twenty extra
parameters pay for themselves, and whether the third profile survives them. Run
the grid again under the corrected measurement model:

```{r enumerate-full}
corrected_candidates <- enumerate_classes(course_engagement, activity,
                                          id = "student",
                                          n_profiles = 2:3, n_group_classes = 2,
                                          covariance_model = "full",
                                          n_starts = 2, max_iter = 1000, seed = 1)

summary(corrected_candidates)
```

That settles the disagreement from the enumeration section. The two-profile
model with a full residual covariance has a BIC of 15190 against 15249 for the
three-profile diagonal model, and beats the three-profile *full* model on almost
every criterion as well. The third profile was the `attendance` correlation in
disguise: given no way to say that attendance and clicks move together, the
model said it with a class instead. This is the most common way an extra profile
appears in real latent profile work, and the residual table is what names it.

Everything from here uses `corrected`.

```{r corrected-entropy}
entropy_table(corrected)
```

Note that the entropy went *down*, from 0.946 to 0.917. A model that fits better
classifies slightly less sharply here, because some of what was separating the
profiles was the correlation the diagonal model had nowhere else to put. Fit and
classification quality are different questions, and a drop in entropy is not by
itself a reason to prefer the worse model.

**Does the profile structure account for the nesting?** The ICCs from the first
section said between a tenth and a quarter of each indicator's variance lay
between students. Split the same calculation by assigned profile:

```{r descriptives-by-profile}
descriptives(corrected, by = "profile")
```

The ICCs collapse towards zero. Within a profile the students no longer differ,
which is precisely the model's claim: students differ in the *mix* of enrolments
they produce, not in the enrolments themselves. The raw data agree with the
model without being asked to.

Everything printed in this section and the ones before it is also available as a
single call, which is what to run first and what not to stop at. `report()` is a
first look; every table it prints is what one of the verbs above returns, and
those are where the numbers come from when you need them as data:

```{r report}
report(corrected, plots = FALSE)
```

## Relating the profiles to something the model never saw

A profile earns its keep by relating to something outside the indicators that
defined it. Two shortcuts are tempting and both distort the answer. Estimating
the outside variable jointly with the measurement model lets it pull the
profiles towards itself, so the profiles being reported are no longer the ones
that were inspected. Assigning each enrolment to its modal profile and comparing
group means instead throws away the fact that some assignments are wrong, and
attenuates every difference it then reports. The three-step family does neither.

The variable here is `previous_grade`: the grade the student earned in the
course *before* this one, standardised across the cohort. It ships with the data
and no model above has seen it. Its causal direction is the interesting part —
the grade comes first, so it is a predictor of engagement rather than a
consequence of it — and the two verbs below read the same association from the
two ends.

The correction starts from how often an enrolment that truly belongs to one
profile is assigned to the other. This is not the average-posterior table from
the last section: that conditioned on the profile an enrolment was *assigned*
to, this conditions on the profile it is actually in, and the correction needs
the second:

```{r classification-errors}
classification_errors(corrected)
```

Those probabilities become one weight per enrolment per profile. The weights are
not probabilities: an enrolment assigned to one profile carries a small
*negative* weight for the other, which is the arithmetic that undoes the
attenuation.

```{r bch}
head(bch_weights(corrected), 4)
```

`three_step()` applies them. Asked for a contrast, it reports the difference
between the profiles rather than a mean per profile, because a difference is the
quantity that has a null worth testing:

```{r three-step}
three_step(corrected, data = course_engagement, outcome = "previous_grade",
           contrast = "pairs")
```

Enrolments in the disengaged profile follow grades more than half a standard
deviation below those the engaged enrolments follow. The naive modal comparison
is shown for contrast:

```{r three-step-modal}
three_step(corrected, data = course_engagement, outcome = "previous_grade",
           contrast = "pairs", method = "modal")
```

The two nearly agree, and the reason is in the table three chunks above: under
three percent of enrolments are misclassified, so there is almost nothing for
the correction to undo. That is the useful lesson rather than a disappointing
one. The gap between these two rows is a function of how bad the classification
is, so `classification_errors()` is what tells you in advance whether the
correction will matter.

`three_step()` asks what a profile predicts. `r3step()` asks the reverse: what
predicts the profile. That is the direction these data were built with, so it is
the one to report. It regresses membership on the covariates with the same
correction applied, reporting log odds against the last profile as reference:

```{r r3step}
r3step(corrected, data = course_engagement, covariates = "previous_grade")
```

Profile 1 is the engaged profile, so a positive slope says a higher previous
grade makes engagement in the next course more likely. A hundred and six
students are enough to ask for a sandwich clustered on students instead of the
model-based information:

```{r r3step-robust}
r3step(corrected, data = course_engagement, covariates = "previous_grade",
       vcov_type = "robust")
```

The estimate is unchanged, as it must be — clustering changes the variance, not
the solution — and the standard error tightens slightly.

### A standard error that cannot exist is refused

A cluster-robust variance is a sum of one contribution per independent group,
and those contributions sum to zero at the estimate — that sum *is* the
condition the estimate solves. With too few groups the result is singular, and
with one group it is exactly zero. Neither is a small standard error. Fit the
same model to two students and ask for the clustered version:

```{r refuse-clusters}
two_students <- subset(course_engagement, student <= 2)

tiny <- multilpa(two_students, activity, id = "student", n_profiles = 2,
                 n_group_classes = 1, n_starts = 4, seed = 1)

tryCatch(three_step(tiny, data = two_students, outcome = "previous_grade"),
         multilpa_too_few_groups = refusal)
```

The variance that *can* exist here is the unclustered one, and it has to be
asked for by name, so that a report of these numbers has to say the nesting was
ignored:

```{r independent}
three_step(tiny, data = two_students, outcome = "previous_grade",
           vcov_type = "independent")
```

### The one-step alternative

The three-step methods hold the measurement fixed on purpose. The one-step model
estimates the covariate effect and the profiles together, so the covariate
*does* help shape the profiles. That is the trade, and the package offers both
rather than naming one correct:

```{r covariates}
covariate_fit <- fit_covariates(course_engagement, activity, id = "student",
                                n_profiles = 2, n_group_classes = 2,
                                profile_covariates = "previous_grade",
                                covariance_model = "full",
                                n_starts = 2, seed = 1)

as.data.frame(covariate_fit, what = "coefficients")
```

The labels switched again, and the measurement table is what says so:

```{r covariate-measurement}
as.data.frame(covariate_fit)
```

`profile_1` is the low-activity profile in this refit, so the membership slope
of -0.456 points the same way as the +0.60 `r3step()` gave for engagement, at
about three quarters of its size. Some of the covariate's association has
been spent moving the profiles rather than describing them, which is exactly
what the one-step model is for and exactly why the three-step estimate is the
one to quote when the measurement model is the thing you inspected.

A covariate fit is a fit like any other, so the same inference verb works on it.
The membership coefficients are the last rows it returns, after the measurement
model:

```{r covariate-inference}
tail(parameter_inference(covariate_fit), 4)
```

The slope's standard error is what the one-step model buys over the coefficient
table above, which carries the estimate alone.

## Building the same model a different way

### Deciding the measurement before asking about groups

Everything so far estimated the measurement model and the group structure at the
same time, which means adding a group class can move the profiles it was meant
to describe. Staging forbids that: the measurement is fitted once, on its own,
and then held while the group classes are estimated.

```{r staged}
staged <- fit_staged(course_engagement, activity, id = "student", n_profiles = 2,
                     n_group_classes = 2, covariance_model = "full",
                     n_starts = 6, tol = 1e-10, seed = 1)

as.data.frame(staged, what = "stages")
```

One row per stage, and two parameter counts. `parameters` counts what that stage
estimated, which is what to use when comparing staged fits with one another;
`parameters_with_measurement` adds the held blocks back, which is what to use
against a jointly estimated fit. First-stage uncertainty is not propagated into
the second, so everything the second stage reports is conditional on a
measurement solution treated as known.

The staged fit is a `multilpa` object, so the answer comes back from the same
accessor as before, and lands within a few thousandths of the joint fit's:

```{r staged-probabilities}
as.data.frame(staged, what = "profile_probabilities")
```

What a stage hands to the next one is a starting-value object. It is also what
to keep when a solution should be reproduced exactly later, or handed to a fit
on new data:

```{r starts}
starting_values(corrected, what = "measurement")
```

Its blocks are indexed by position, because that is what the fitter expects; the
labelled view is the accessor rather than the list:

```{r starts-table}
as.data.frame(starting_values(corrected))
```

A staged fit cannot be bootstrapped against a freely estimated one. One holds
its measurement and the other does not, so the smaller model is not a
restriction of the larger and the likelihood-ratio statistic would compare two
parameter counts that do not correspond. The model below differs from `staged`
in exactly the one class a bootstrap comparison is allowed to differ in, and is
still refused:

```{r refuse-nesting}
flat_corrected <- multilpa(course_engagement, activity, id = "student",
                           n_profiles = 2, n_group_classes = 1,
                           covariance_model = "full",
                           n_starts = 6, tol = 1e-10, seed = 1)

tryCatch(bootstrap_lrt(flat_corrected, staged, iter = 9, seed = 1),
         multilpa_bad_nesting = refusal)
```

### When students differ by degree rather than by kind

Two latent classes of student is one answer to *how do students differ*. The
other is that they do not fall into kinds at all: each student simply sits a
little higher or lower on every indicator. That is a random intercept, and it is
a different model rather than a variant of this one.

The likelihood has an integral in it, done by Gaussian quadrature. Thirty-one
nodes are enough here, and the fit checks that for itself by recomputing the
likelihood on a finer grid and reporting the difference:

```{r random-intercept}
shifted <- fit_random_intercept(course_engagement, activity, id = "student",
                                n_profiles = 2, n_starts = 2,
                                quadrature_nodes = 31L, seed = 1)

summary(shifted)
```

The estimated random-intercept standard deviation is about six hundredths of a
point on indicators that span six: the model was given the chance to say
students sit at different levels and found almost nothing. Its BIC is far worse
than the two-class model's, and the whole of this vignette says why. The
students in these data do not differ in *where* their enrolments sit, they
differ in *which* enrolments they produce, and only a model with a second latent
variable can put that difference anywhere.

The posterior intercept for each student is a table like everything else:

```{r random-intercepts}
head(as.data.frame(shifted, what = "random_intercepts"))
```

This family provides no standard errors either, and asking for them raises a
classed condition rather than returning something that looks like one. The last
section shows that refusal on another family.

## The same students, in the order they took their courses

Nothing changes dataset here. `sequence` orders each student's courses within
their own programme, so the same 1422 rows that answered *what kinds of
enrolment are there* can answer *does a student's engagement carry from one
course to the next*. That continuity is the point of shipping one dataset: the
transition model below is fitted to the same rows, the same students and the
same indicators as the model above.

The first thing to do with ordered data is to look at it in order. Fitted with
`time =`, the model stores the ordering without using it, and the assignments
can then be read back as sequences:

```{r pace-fit}
pace <- multilpa(course_engagement, activity, id = "student",
                 time = "sequence", n_profiles = 2, n_group_classes = 2,
                 covariance_model = "full", n_starts = 6, seed = 1)

head(sequences(pace))
```

One row per observation is the shape to compute on. One row per student, one
column per course, is the shape to look at, and the shape sequence-analysis
packages expect. The trailing `NA`s are real: students who took fewer courses
have shorter sequences, so the panel is ragged rather than padded.

```{r sequences-wide}
head(sequences(pace, format = "wide"), 4)
```

```{r plot-sequences, fig.height = 4.6}
plot(pace, what = "sequences")
```

Before reading anything into the picture, check that the two kinds of student
are built on comparable amounts of data. A class whose students are
systematically shorter or gappier is a different finding from a class whose
students simply differ in profile mix, and the prevalence table cannot show it:

```{r sequence-summary}
sequence_summary(pace)
```

The two classes average thirteen and a half courses each and no sequence has a
hole in it, so they are comparing like with like. Only the ragged tails differ,
and they differ by one course.

The model above still ignores the ordering it stored. The model that uses it
estimates where students move — and estimating it separately in each group class
asks whether the two kinds of student differ in how *sticky* engagement is, not
just in how much of it there is:

```{r transitions}
moves <- fit_transitions(course_engagement, activity, id = "student",
                         time = "sequence", n_profiles = 2, n_group_classes = 2,
                         covariance_model = "full", n_starts = 6, seed = 1)

transitions(moves)
```

`stable` marks the diagonal — staying in the same profile. Both labellings in
that table are the model's own, so read them against the truth columns before
reading anything into the numbers:

```{r transitions-recovery}
xtabs(~ profile + engagement, data = assignments(moves, data = course_engagement))

xtabs(~ group_class + student_type, data = assignments(moves, data = course_engagement))
```

Profile 1 is the engaged pattern here, and group class 2 is the committed
student. The two kinds of student are mirror images on the diagonal. Among
committed students — group class 2, just under two thirds of the cohort —
engagement persists (0.88) while disengagement does not (0.58): a lapse is
temporary. Among wavering students it is the other way round (0.76 and 0.93): a
lapse tends to stick. The mix estimated at the start of this vignette and the
persistence estimated here are two views of the same two kinds of student, and
only the ordered model can see the second.

## What the package refuses

Three refusals have already appeared in this vignette: a cluster-robust standard
error with too few groups, a bootstrap comparison whose two models are not
nested, and a candidate that was never enumerated. Here is the fourth. A model
family for which no standard error is implemented says so, rather than returning
a number that would look like one:

```{r refuses}
tryCatch(vcov(moves), multilpa_no_inference = refusal)
```

Every refusal in the package is a classed condition, which is why every
`tryCatch()` above names the class it is catching instead of matching message
text. `?"multilpa-conditions"` lists all of them.
