---
title: "Covariates and staged estimation in multilevel LPA"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Covariates and staged estimation in multilevel LPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4, width = 100)
```

Three-step covariate analysis estimates associations between latent profile
membership and variables outside the measurement model, with an adjustment for
classification error. The course-engagement example asks whether previous
grades are associated with a student's engagement in the next course. This
guide compares membership regression, differences in previous grades across
profiles, and joint estimation with grades included. It also shows how staged
estimation holds profile definitions fixed while estimating student classes.

```{r setup}
library(multilpa)
```

## Data and engagement profiles

The bundled synthetic `course_engagement` data record enrolments nested within
students. The five indicators are page views (`browse`), lecture videos viewed
(`lectures`), forum posts read (`forum_read`), posts written (`forum_post`), and
active days (`attendance`), all stored as `log1p` counts standardized within
course. `previous_grade` is the
standardized grade from the preceding course; an entry grade is used for the
first course. Its temporal position makes it a predictor of subsequent
engagement. The [introductory guide](multilpa.html) explains the longitudinal
case and the two levels in more detail.

The indicator vector `vars` defines the activity measures used to estimate the
profiles. Keeping `previous_grade` separate lets us first establish the
engagement profiles and then examine their association with grades.

```{r variables}
vars <- c("browse", "lectures", "forum_read", "forum_post", "attendance")
```

`multilpa()` estimates profile means and variances together with the probability
of each profile within each student class. Write that probability as
$\pi_{k\mid m}=P(C_{it}=k\mid G_i=m)$, where $C_{it}$ is student $i$'s
profile in course $t$ and $G_i$ is their student class. The probabilities sum
to one across profiles within a student class. The model below has three
profiles and three student classes, using independent Gaussian indicators
within each profile and common profile definitions across student classes.

The fit assumes independent students and independent enrolments conditional on
student class. Indicator dependence can change the profiles, and multiple
likelihood maxima can occur. The [evaluation guide](multilpa-evaluation.html)
examines these issues; here the same candidate model provides a consistent
basis for comparing covariate methods. Ten starts and a fixed seed make the
numerical result reproducible.

```{r fit}
set.seed(1)
fit <- multilpa(course_engagement, vars, id = "student", time = "sequence", n_profiles = 3, n_group_classes = 3, n_starts = 10, max_iter = 2000, tol = 1e-10, seed = 1)
fit
```

The profiles have higher, intermediate, and lower activity in that order.
Their attendance means are 3.753, 2.915, and 1.840. Profile 3 is therefore the
lower-activity reference when interpreting the membership regressions below.
These labels apply to this fit; each subsequent model's means need checking.

## Accounting for classification error

The classification-error matrix estimates the probability of an assigned
profile given the underlying profile:
$D_{ks}=P(\widehat C=s\mid C=k)$. Here $\widehat C$ is the profile with the
largest posterior probability. Each row of $D$ sums to one; diagonal entries
represent correct assignment under the fitted model. `get_data()` extracts
this matrix through `"classification_errors"`.

The adjustment assumes that assignment errors are independent of the auxiliary
variable conditional on the underlying profile. The error matrix is estimated
from the measurement model and then treated as fixed. Consequently, later
intervals omit uncertainty in that first-stage estimate. Poor separation can
also make the correction unstable.

```{r classification-errors}
get_data(fit, "classification_errors", level = "individuals")
```

The estimated correct-assignment probabilities are 0.863, 0.852, and 0.984.
Most ambiguity concerns profiles 1 and 2: the estimated probability of assigning
profile 1 to profile 2 is 0.137, and the reverse probability is 0.121.
Treating these assignments as certain would lose that uncertainty.

## Previous grade as a membership predictor

`r3step()` estimates log-odds coefficients relating previous grade to profile
membership while keeping the profile definitions and error matrix fixed. With
profile 3 as the reference and $x$ denoting previous grade, the regression is

$$\log\frac{P(C=k\mid x)}{P(C=3\mid x)}=\alpha_k+\beta_k x,\qquad k=1,2.$$

The intercept $\alpha_k$ describes the log odds at average previous grade,
$x=0$. The slope $\beta_k$ describes the change per one standard deviation of
previous grade. Positive slopes favor profile $k$ relative to profile 3.
The regression assumes a linear effect on the log-odds scale. For enrolment-level
analysis, `vcov_type = "robust"` clusters the score contributions by student;
it requires enough independent students for an estimable covariance matrix.
This regression relates marginal profile membership to grades, whereas the
one-step model below estimates membership conditional on student class.
The reported 95% intervals use the normal approximation. The adjusted p-values
apply the Benjamini–Hochberg correction across the slopes; the intervals are
pointwise and do not receive that correction.

```{r membership-predictor}
r3step(fit, data = course_engagement, covariates = "previous_grade", vcov_type = "robust")
```

A one-standard-deviation increase in previous grade is associated with a 0.647
increase in the log odds of higher activity versus lower activity, and a 0.521
increase for intermediate versus lower activity. Their confidence intervals
are 0.492 to 0.801 and 0.381 to 0.660. Both associations favor more active
engagement after higher previous grades. They are observational associations
and do not establish that grades cause the subsequent engagement pattern.

## Comparing previous grades across profiles

`three_step()` estimates class-specific means of a numeric auxiliary variable
and differences between those means. The default BCH method uses signed
weights derived from the inverse classification-error matrix. For grades
$x_u$ and profile weights $w_{uk}$ across enrolments $u$, the estimated mean is
$\widehat\mu_k=\sum_u w_{uk}x_u/\sum_u w_{uk}$. A pairwise contrast compares
$\widehat\mu_k-\widehat\mu_l$ for profiles $k$ and $l$.

The method requires an invertible error matrix and usable weighted class
sizes. Its default standard errors account for enrolments clustered within
students, while treating the classification-error matrix as fixed. Although
this method also supports distal outcomes, the variable here precedes
engagement. The contrasts describe differences in prior grades across profiles.
The 95% intervals are pointwise normal intervals. Adjusted p-values apply the
Benjamini–Hochberg correction across the pairwise contrasts.

```{r grade-contrasts}
three_step(fit, data = course_engagement, outcome = "previous_grade", contrast = "pairs")
```

Previous grades in profile 3 are 0.601 standard deviations below profile 1 and
0.486 below profile 2. The profile-2 versus profile-1 difference is -0.115,
with an interval from -0.286 to 0.057. The contrast between the two more active
profiles is therefore less certain than their contrasts with lower activity.

Modal analysis computes the same contrasts after assigning every enrolment to
one profile with certainty. It provides a direct comparison with the
classification-adjusted estimates, but its standard errors omit assignment
uncertainty.

```{r modal-contrasts}
three_step(fit, data = course_engagement, outcome = "previous_grade", contrast = "pairs", method = "modal")
```

The modal profile-2 versus profile-1 difference is -0.095, compared with -0.115
under BCH. Its standard error is 0.064, compared with 0.088 after correction.
The adjustment changes both the estimated contrast and its uncertainty where
the profiles overlap most.

## Estimating profiles and grade associations together

One-step estimation jointly estimates the profile measurement model and the
membership regression. `multilpa(profile_covariates = "previous_grade")`
allows the grade association to contribute to estimation of the profiles.
For student class $m$, the model is

$$\log\frac{P(C=k\mid G=m,x)}{P(C=3\mid G=m,x)}=\alpha_{km}+\beta_k x.$$

Profile-specific intercepts vary by student class; grade slopes are shared
across student classes. The model retains the Gaussian and conditional
independence assumptions of the original fit. Because the profiles can change
and the coefficients are conditional on student class, their magnitudes need
not equal the three-step coefficients. The fitted means establish which
profile each coefficient describes.

```{r one-step}
covariate_fit <- multilpa(course_engagement, vars, id = "student", n_profiles = 3, n_group_classes = 3, profile_covariates = "previous_grade", n_starts = 6, max_iter = 2000, tol = 1e-10, seed = 1)
covariate_fit
```

The profiles retain their higher, intermediate, and lower activity ordering.
Their attendance means are now 3.767, 2.926, and 1.842. The profile definitions
have changed slightly from the original fit.

The `"coefficients"` table reports the estimated intercepts and grade slopes.
Profile 3 is the reference for the profile-level coefficients.

```{r one-step-coefficients}
get_data(covariate_fit, "coefficients")
```

The previous-grade slopes are 0.503 for higher activity and 0.317 for
intermediate activity relative to lower activity. Both point in the same
direction as the three-step slopes of 0.647 and 0.521. Their smaller magnitudes
cannot be attributed solely to changing profile definitions because the
one-step regression also conditions on student class.

## Holding profile definitions fixed across student classes

Staged estimation estimates a common set of profiles first, then estimates
student-class probabilities with those measurement parameters held fixed.
`fit_staged()` begins with one student class and subsequently estimates the
requested three-class composition. The second stage estimates
$\pi_{k\mid m}$ and $\omega_m=P(G=m)$, while retaining the first stage's
profile means and variances. This separates profile estimation from estimation
of the higher-level structure.

The first-stage uncertainty is not propagated into the second stage. The
result is conditional on the fixed measurement model and is not the joint
maximum-likelihood solution. Profile and student-class numeric labels can differ
from those in the joint fit, so comparisons require matching their meanings.

```{r staged}
staged <- fit_staged(course_engagement, vars, id = "student", n_profiles = 3, n_group_classes = 3, n_starts = 6, max_iter = 2000, tol = 1e-10, seed = 1)
staged
```

The staged fit labels intermediate activity as profile 1, lower activity as
profile 2, and higher activity as profile 3. Their attendance means are 2.872,
1.809, and 3.710. The activity patterns remain recognizable, with different
numeric labels and slightly different means.

The `"stages"` table distinguishes the parameters estimated at each stage.
`parameters_with_measurement` also counts the measurement parameters held
fixed during the membership stage.

```{r stages}
get_data(staged, "stages")
```

The measurement stage estimates 32 parameters; the membership stage estimates
8, with 38 parameters when the held measurement parameters are included.
The second-stage log likelihood is -7529.197, compared with -7526.274 for the
joint fit printed earlier. Holding the profiles fixed constrains the final
solution. A descriptive comparison with a joint fit uses
`parameters_with_measurement`, which includes the held profile parameters.

## Interpretation and limitations

The membership regressions and grade contrasts agree that lower activity is
associated with lower previous grades. The distinction between higher and
intermediate activity has more classification uncertainty and a less certain
grade contrast. Three-step and joint estimation address different conditional
relationships; the appropriate estimate depends on whether the goal is to
retain previously defined profiles or estimate them together with covariates.

All these results depend on the candidate measurement model. The
[evaluation guide](multilpa-evaluation.html) shows why its diagonal covariance
and class counts require scrutiny. Classification adjustments do not repair a
misspecified profile model, and clustering standard errors by student does not
make observational associations causal. Staged and three-step intervals also
omit uncertainty in the quantities held fixed from the first stage.
