---
title: "Evaluating multilevel latent profile models"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Evaluating multilevel latent profile models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Multilevel latent profile analysis (MLPA) estimates observation-level profile
means and the probabilities of those profiles within higher-level latent
classes. Model evaluation asks whether the estimated profiles separate the
observations, whether the measurement assumptions describe their associations,
and whether a different number of profiles or classes gives a better fit. The
three-profile, three-class model from [the introductory vignette](multilpa.html)
is the starting point for these checks.

```{r setup, message = FALSE}
library(multilpa)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 6, width = 100)
vars <- c("browse", "lectures", "forum_read", "forum_post", "attendance")
```

## Data and starting model

`course_engagement` contains repeated course enrolments nested within students.
The bundled data are synthetic; their relationship to the longitudinal
engagement study is described in the [introductory vignette](multilpa.html).
The indicators measure course-page views (`browse`), lecture-video views
(`lectures`), forum posts read (`forum_read`), forum posts written
(`forum_post`), and days active (`attendance`), each a `log1p` count
standardized within course.
`student` identifies the higher-level unit, and `sequence` orders each
student's courses. The model uses student nesting; courses are not an
additional grouping level.

The starting model estimates three Gaussian profiles and three student classes.
For enrolment profile $k$ and student class $m$, its central quantity is
$\pi_{k\mid m}=P(C_{it}=k\mid G_i=m)$, where $C_{it}$ is student $i$'s
profile at occasion $t$ and $G_i$ is that student's class. Each probability
lies between zero and one, and the probabilities sum to one within a student
class. Profile means and variances are shared across student classes.

The diagonal covariance model assumes independent indicators within a profile
and independent enrolments conditional on the student's class. Students are
independent of one another. These assumptions can fail when indicators share
activity counts, when engagement persists across adjacent courses, or when
students share unmodeled course effects. Multiple starts reduce sensitivity to
local likelihood maxima; they do not guarantee a global maximum.

`multilpa()` estimates this model by maximum likelihood. The stricter `tol`
helps the likelihood score approach zero before the standard-error calculation.

```{r fit}
set.seed(1)
fit <- multilpa(course_engagement, vars, id = "student", time = "sequence", n_profiles = 3, n_group_classes = 3, n_starts = 10, max_iter = 2000, tol = 1e-10, seed = 1)
fit
```

The fit describes 1,422 enrolments from 106 students. Profile 1 has the highest
mean on every indicator, profile 2 has intermediate means, and profile 3 has
the lowest means. Attendance means are 1.133, 0.256, and -0.940, respectively,
in standard deviations from each course's mean. These differences define the fitted profiles; the
checks below assess their support.

`get_results(fit, "starts")` reports convergence and the maximized log likelihood
for each initialization. Agreement across converged starts supports the
numerical stability of the chosen solution; distinct maxima or active variance
bounds require further investigation.

```{r starts}
get_results(fit, "starts")
```

All ten starts converged without a reported boundary and reached a log
likelihood of approximately -8348.429. The class-count comparisons below
therefore begin with a replicated solution.

## Classification uncertainty

Relative entropy measures the concentration of posterior class probabilities.
For $N$ units, $K$ classes, and posterior probability $p_{uk}$ that unit $u$
belongs to class $k$, it is

$$
E=1-\frac{-\sum_{u=1}^{N}\sum_{k=1}^{K}p_{uk}\log p_{uk}}{N\log K}.
$$

The value is zero when every unit has equal probability of every class and one
when every unit has probability one of a single class. The calculation is
separate for enrolments and students. It is conditional on the fitted model:
incorrect measurement assumptions can still produce concentrated probabilities.
Relative entropy is undefined for a level with a single class.

`get_results(fit, "entropy")` reports both the unnormalized entropy sum and the
relative entropy for the two levels.

```{r entropy}
get_results(fit, "entropy")
```

Relative entropy is 0.792 for enrolments and 0.797 for students. Both levels
retain classification uncertainty, so replacing every posterior distribution
with its most probable class discards information.

Average posterior probability measures certainty within each assigned class.
For the set $A_k$ of units assigned to class $k$, it is
$\bar p_k=|A_k|^{-1}\sum_{u\in A_k}p_{uk}$. Values closer to one indicate
more certain assignments. The quantity conditions on the fitted classes and
their modal assignments; it cannot establish that the class interpretation is
correct. `get_results(fit, "classification")` also reports how many units were
assigned to each class.

```{r classification}
classification <- get_results(fit, "classification")
classification[, c("level", "class", "n_modal", "average_posterior")]
```

Profile 2 has 478 assigned enrolments and an average posterior probability of
0.853, compared with 0.978 for the 576 enrolments assigned to profile 3.
Student class 2 also has less certain assignments than class 3: their average
posterior probabilities are 0.864 and 0.977. Later analyses should retain this
uncertainty; [the covariate vignette](multilpa-covariates.html) describes
classification-error adjustments.

## Residual dependence

Bivariate residual correlations measure the association left unexplained by
the profile measurement model. For indicators $a$ and $b$ within profile $k$,
the residual is $r_{ab,k}^{\mathrm{observed}}-r_{ab,k}^{\mathrm{expected}}$.
The observed correlation uses posterior profile weights; the expected
correlation is zero under the diagonal model. A residual near zero indicates
agreement for that pair, while its sign identifies the direction of the
unexplained association.

Residual correlations depend on the estimated profile membership. Their
reported p-values are descriptive approximations: they do not account for
estimated measurement parameters, posterior memberships, or dependence within
students. The residual magnitudes identify measurement assumptions to examine;
the p-values are not calibrated tests of those assumptions.

`get_results(fit, "residuals")` returns indicator pairs ranked by absolute residual.
The first four rows identify the largest discrepancies within profiles.

```{r residuals}
residuals <- get_results(fit, "residuals")
head(residuals[, c("profile", "indicator_1", "indicator_2", "observed", "expected", "residual")], 4)
```

The largest residual is 0.211 between forum reading and attendance in profile
3, followed by 0.177 between browsing and attendance in the same profile.
All four displayed pairs include attendance. The three-profile model has left
an association between recorded activity and active days within its profiles.

## Class counts and covariance assumptions

Class enumeration compares fitted models with different numbers of profiles
and student classes. The Bayesian information criterion is
$\mathrm{BIC}=-2\ell+q\log N$, where $\ell$ is the maximized log likelihood,
$q$ is the number of free parameters, and $N$ is the chosen sample size.
Lower values favor a model's balance of fit and parameter count within the
candidate set. Here `bic_groups` uses the number of students for $N$ throughout.

The covariance comparison asks whether allowing indicators to correlate within
profiles changes the preferred class counts. `structure = "VVI"` permits
profile-specific variances and fixes residual covariances to zero.
`structure = "VVV"` estimates an unrestricted covariance matrix for each
profile. Both models retain Gaussian profiles and common measurement
parameters across student classes. Full covariance models need more parameters
and may have poorly separated maxima, especially when a profile has few
observations.

`enumerate_classes()` fits every requested combination to the same enrolments
and indicators. `get_results()` returns BIC and the numerical diagnostics for
each candidate. These comparisons require converged fits without active bounds;
the best likelihood should recur across starts. BIC alone does not validate
the Gaussian assumptions or establish a unique number of classes.

```{r enumeration}
candidates <- enumerate_classes(course_engagement, vars, id = "student", time = "sequence", n_profiles = 2:3, n_group_classes = 2:3, structure = c("VVI", "VVV"), n_starts = 6, max_iter = 2000, tol = 1e-10, seed = 1)
comparison <- get_results(candidates, "candidates")
comparison[, c("n_profiles", "n_group_classes", "structure", "bic_groups", "converged", "boundary", "n_best_replicated")]
```

The diagonal models favor three profiles and three student classes, with BIC
16874.1. Allowing residual covariance changes the preferred profile count: the
two-profile, three-class model has BIC 16800.6, the lowest in the grid, and its
best solution recurs across all six starts. The full covariance model with three
profiles and three classes has BIC 16867.3. Only the three-profile, two-class
full-covariance model fails to replicate, reaching its best solution once in six
starts. Thus the requested three-by-three model is a working description whose
support depends on the measurement assumptions.

`candidate_fit()` retrieves a fitted candidate by its class counts and
covariance structure. Its residual table checks whether the estimated
covariances reproduce the within-profile correlations.

```{r covariance-check}
full <- candidate_fit(candidates, n_profiles = 2, n_group_classes = 3, structure = "VVV")
full_residuals <- get_results(full, "residuals")
head(full_residuals[, c("profile", "indicator_1", "indicator_2", "observed", "expected", "residual")], 4)
```

The full covariance model estimates a forum-reading–attendance correlation of
approximately 0.267 in profile 2, almost exactly its observed value. The
displayed residuals are close to zero because these associations are now
estimated. Their agreement checks the fitted covariance calculation; it does
not independently validate the profile count or the student classes.

## Uncertainty in profile means

Parameter inference estimates the sampling uncertainty of a fitted parameter
$\hat\theta$. The default standard error is the square root of the relevant
diagonal element of the inverse observed information matrix. A 95% Wald
interval is $\hat\theta\pm1.96\operatorname{SE}(\hat\theta)$.

These local approximations require a converged solution, inactive parameter
bounds, and an invertible information matrix. They condition on the specified
number of classes and measurement model. An interval can therefore be narrow
even when an alternative covariance model supports a different profile count.
`parameter_inference()` calculates intervals for the fitted model; the table
below retains attendance means from the original three-by-three fit.

```{r inference}
inference <- parameter_inference(fit)
subset(inference, parameter == "mean" & term == "attendance", select = c(outcome, term, estimate, standard_error, conf_low, conf_high))
```

The attendance mean in profile 1 is 1.133 with a standard error of 0.061 and
an interval from 1.013 to 1.252. Profile 2's interval is 0.155 to 0.357, and
profile 3's is -0.993 to -0.887. The means are distinct within this fitted model;
these intervals do not resolve the covariance sensitivity identified above.

## Limits of the interpretation

The class-count comparison covers a restricted candidate set and depends on
the chosen BIC sample-size convention. The full covariance candidates with
unreplicated best solutions need additional starts before substantive use.
All reported classifications and intervals depend on the corresponding
measurement specification.

Student nesting accounts for differences in profile composition, but the
model does not estimate serial dependence between adjacent enrolments.
`time = "sequence"` preserves their order for sequence summaries.
[The transition vignette](multilpa-transitions.html) estimates the probabilities
of movement between profiles when that is the scientific question.
