---
title: "Changes in engagement across successive courses"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Changes in engagement across successive courses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(multilpa)
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
options(digits = 4, width = 120)
```

Latent transition analysis estimates the probability of moving between latent
profiles over successive observations. Here, the observations are students'
engagement in successive courses, and student classes distinguish patterns of
persistence and change. The analysis asks whether engagement tends to persist
from one course to the next and whether that persistence differs between
students. The [main vignette](multilpa.html) explains the corresponding
two-level profile model; this analysis adds the ordering of courses.

## Course engagement over time

The longitudinal case follows the question examined by Saqr and López-Pernas
(2021): how does online engagement develop across a programme? Their study
identified three course-level engagement states and three longer-term
trajectories using latent class analysis and hidden Markov models. The example
here uses the bundled synthetic `course_engagement` data to examine that
question with three engagement profiles and three student classes. These are
candidate dimensions of the model; they do not establish the number of
profiles or reproduce the paper's estimates.

Course enrolments are nested within students. `student` identifies the student,
`course` identifies the course, and `sequence` gives its position in that
student's course history. Students can have different numbers of enrolments.
The model follows each student's sequence, so a transition refers to movement
between successive observed courses.

Five indicators describe engagement within each course. `browse` measures
course page views, `lectures` measures lecture videos viewed, `forum_read`
measures forum posts read, `forum_post` measures posts written, and `attendance`
measures days of online activity. Each indicator is stored as `log1p` of its
activity count, standardized within course: a value is that enrolment's
distance from its own course's average, in that course's standard deviations,
and higher values indicate more activity. The indicator names are
collected in `vars` for use in the model.

```{r variables}
vars <- c("browse", "lectures", "forum_read", "forum_post", "attendance")
```

## Estimating persistence and change

The transition model represents engagement at course position $t$ for student
$i$ by a latent profile $C_{it}$, and the student's class by $G_i$. An initial
probability $\delta_{k\mid m} = P(C_{i1}=k\mid G_i=m)$ describes the chance of
starting in profile $k$ for a student in class $m$. A transition probability
describes what follows a particular profile:

$$
a_{kl\mid m} = P(C_{i,t+1}=l\mid C_{it}=k,G_i=m).
$$

Each probability lies between zero and one. For a given class and starting
profile, the probabilities across destination profiles sum to one. The
diagonal value $a_{kk\mid m}$ is the probability of remaining in profile $k$;
the other values describe changes to another profile. Each student class has
its own initial probabilities and transition matrix.

The model assumes independent students, and that the next profile depends on
the current profile and student class. Transition probabilities remain constant across course
positions. Profile means and variances are shared across positions and student
classes, so a profile has a common definition throughout the programme.
Continuous indicators follow Gaussian distributions within each profile. The
default diagonal covariance structure assumes that indicators are independent
within a profile. Residual associations, changing profile definitions, or
transitions that vary across courses can therefore distort the estimated
patterns.

`lta()` estimates the profile distributions, student-class
probabilities, initial probabilities, and transition probabilities jointly by
maximum likelihood. It uses the indicators in `vars`, groups observations by
`student`, and orders them by `sequence`. Course positions must be observed and
unique within each student. Multiple starts reduce sensitivity to initial
values, although they do not guarantee the global maximum. The model below
estimates three profiles and three student classes from six starts.

```{r transition-model}
set.seed(1)
moves <- lta(course_engagement, vars = vars, id = "student", time = "sequence", n_profiles = 3, n_group_classes = 3, n_starts = 6, max_iter = 2000, seed = 1)
moves
```

The model uses 1,422 enrolments from 106 students, with histories extending to
15 courses. The selected solution converges after 276 iterations. Its profile
means distinguish three activity patterns: profile 1 has the highest mean on
every indicator, profile 2 is intermediate, and profile 3 has the lowest means.
For example, their mean `attendance` values are 1.091, 0.244, and -0.928
standard deviations from each course's mean. We therefore refer to them as the higher-, intermediate-, and
lower-activity profiles. These labels describe this fitted model's means and
must be checked again when interpreting a different fit.

## Engagement at the beginning of the sequence

Initial profile probabilities describe how students enter their observed course
histories. `get_results(moves, "initial")` extracts these probabilities for each
student class. The `probability` column gives $\delta_{k\mid m}$, whereas
`prevalence` gives the posterior expected proportion of enrolments in each
profile within that class, across the observed histories.
`group_class_probability` gives the estimated proportion of students
in each class. These probabilities describe the fitted distributions;
uncertainty about an individual student's class remains part of the model.

```{r initial-probabilities}
get_results(moves, "initial")
```

Class 1 starts in the lower-activity profile with probability 0.841, and class
3 starts there with probability 0.747. Class 2 begins in the higher-activity or
intermediate-activity profiles, with probabilities 0.605 and 0.371, and almost
never in the lower-activity one. Its estimated student-class probability is
0.550, compared with 0.194 and 0.256 for classes 1 and 3. These differences describe starting positions;
the transition probabilities establish how engagement subsequently develops.

## Movement between engagement profiles

Transition probabilities describe the destination profile conditional on the
current profile and student class. `get_results(moves, "transitions")` extracts
the fitted $a_{kl\mid m}$ values. `from` and `to` identify the current and next
profiles, and `stable` marks a transition that stays in the same profile.
`expected_count` is the posterior expected number of transitions of that type,
so it can be fractional. It helps distinguish estimates supported by many
transitions from those based on little information.

The `estimated` flag identifies whether an outgoing transition row has data
support. A profile with no occupancy before the final observation supplies no
information about its next state. Such a row is assigned uniform probabilities
and flagged as unestimated; those values must not be interpreted as evidence
of equal transition chances. The following table gives all transitions for
the three student classes.

```{r transition-probabilities}
get_results(moves, "transitions")
```

Lower activity persists with probability 0.947 in class 3 and 0.746 in class 1.
These classes therefore combine frequent lower activity at entry with a high
probability of remaining there once it occurs. Class 2 has a lower probability
of remaining in that profile, 0.597; its probabilities of moving from lower
activity to higher or intermediate activity are 0.186 and 0.216.

Class 2 also moves between the two more active profiles. Its probability of
moving from higher to intermediate activity is 0.459, and the reverse
probability is 0.395. A student can therefore maintain relatively active
engagement while changing between those profiles. Student classes represent
different transition patterns, while profiles describe engagement within an
individual course.

Class 3's transition from higher to intermediate activity sits at its lower
bound, on an expected count of essentially zero: no observation supports it.
Class 1's transition from higher to lower activity is likewise near zero.
Such boundary estimates call for caution about sparse transitions and the
chosen number of student classes; they do not establish that a transition is
impossible in the population, only that these data carry no evidence for it.

## Limitations

The three-profile, three-class specification describes one candidate model.
The diagonal covariance assumption is consequential because the engagement
indicators can remain associated within profiles. Profile enumeration and
residual dependence should be considered together, as explained in
[model evaluation](multilpa-evaluation.html). A richer profile structure can
otherwise absorb indicator dependence and alter the apparent transitions.

The estimates concern successive observed courses, whose duration and content
may differ. Constant transition probabilities do not capture changes specific
to a particular course or programme stage. Class differences describe
associations in engagement histories and do not establish causes of change.
The package does not provide standard errors or likelihood-ratio tests for
this transition model family. Membership predictors and auxiliary-variable
analyses for the two-level profile model are covered separately in
[covariates and outcomes](multilpa-covariates.html).

## Reference

Saqr, M., & López-Pernas, S. (2021). The longitudinal trajectories of online
engagement over a full program. *Computers & Education, 175*, 104325.
<https://doi.org/10.1016/j.compedu.2021.104325>
