---
title: "Do schools differ in the students they serve?"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Do schools differ in the students they serve?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>",
                      fig.width = 7, fig.height = 4.2, fig.align = "center",
                      dpi = 96)
options(digits = 4, width = 92)
```

```{r library}
library(multilpa)
```

## The question is about schools, not students

Latent profile analysis asks *what kinds of people are there*. This package
exists for the question one level up: given that there are kinds of people, **do
the groups they sit in differ in the mix of people they contain?**

That is not a rephrasing. It is a different parameter, and a single-level model
cannot report it at all.

The package ships the data used here, so nothing below depends on a simulation
you have to read first:

```{r data}
str(school_engagement)

vars <- c("homework_hours", "participation", "interest")
```

720 students in 60 schools, three engagement indicators each. Because the data
are simulated, the column `engaged` records the kind each student was *generated*
from — the truth behind the observations, which a real study never has. We will
fit the model without it and only afterwards ask whether the model found it.

Here is why the second level is not a technicality. Across the whole sample,
**`r round(100 * mean(school_engagement$engaged))`%** of students are the engaged
kind — a tidy headline number. Now look at the schools one at a time:

```{r the-hook}
per_school <- aggregate(engaged ~ school, data = school_engagement, FUN = mean)

summary(per_school)
```

The engaged share runs from **`r round(min(per_school$engaged), 2)`** to
**`r round(max(per_school$engaged), 2)`**, and its median is
**`r round(median(per_school$engaged), 2)`**. The sample average describes no
school in the data. Any analysis that pools the students and reports one number has
averaged away the thing worth knowing.

That hook used `engaged`, which a real study would not have. The check that
works without it is the intraclass correlation — the share of each indicator's
variance lying *between* schools rather than within them:

```{r descriptives}
descriptives(school_engagement, vars = vars, id = "school")
```

Around an eighth of the variance in each indicator is between schools. That is
what makes the rest of this worth doing: had the ICCs come back near zero, the
schools would not differ, and a model built to tell them apart would have
nothing to find. It costs one line and is the step most often skipped.

## Look at the profiles before reading anything about them

The default convergence tolerance is set for the likelihood, and the likelihood
is flat near its maximum. Wald standard errors need the *score* at zero, which
is a stricter requirement, so this fit asks for `tol = 1e-10`. It costs two
extra EM iterations here; at the default the package warns that the intervals
below should not be trusted, which is exactly the kind of warning worth
believing.

```{r fit}
fit <- multilpa(school_engagement, vars, id = "school",
                n_profiles = 2, n_group_classes = 2,
                n_starts = 8, tol = 1e-10, seed = 1)
```

```{r plot-profiles}
plot(fit)
```

Two clearly separated profiles: one doing around four hours of homework with
low participation and interest, one doing around eight with both high. The
separation is what makes the rest of the analysis worth doing; if the lines
crossed or sat on top of each other, no amount of fit statistics would rescue
it.

The numbers behind that picture, one row per profile and indicator:

```{r measurement}
as.data.frame(fit)
```

The indicators are on different scales, so the raw plot partly shows that
homework hours are simply bigger numbers. Standardizing puts them on a common
footing and shows the *shape* of each profile:

```{r plot-standardized}
plot(fit, scale = "standardized")
```

The same estimates can be read three ways, and which one to use depends on the
question. Bars, with a 95% interval on each, answer *how far apart are the two
profiles on this indicator, and is the gap larger than its uncertainty*. The
intervals need the data the model was fitted to, and the fit is carrying it:

```{r plot-bars, fig.height = 4.6}
plot(fit, what = "bars")
```

A heatmap of standard deviations from each indicator's grand mean answers
*which indicators define the profiles* at a glance, and stays readable when
there are more indicators than a line plot can label:

```{r plot-heatmap, fig.height = 3.4}
plot(fit, what = "heatmap")
```

Every view is listed, so none of them has to be remembered:

```{r plot-catalogue}
multilpa_plot_types()
```

Profile labels are arbitrary. Nothing in the likelihood distinguishes "profile
1" from "profile 2", so a refit may swap them. Read the estimates against what
generated them, not against the label.

## Did it find what made the data?

This is the check simulated data exist for, and the reason `engaged` ships
beside the indicators. Asking for the assignments returns every observation with
the profile it was put in, so the truth column and the estimate are in the same
row and nothing has to be lined up by hand:

```{r recovery}
assigned <- assignments(fit, data = school_engagement)

xtabs(~ profile + engaged, data = assigned)
```

One student out of 720 on the wrong side. That is not a claim about this
package's accuracy in general — the profiles here are far apart by construction,
and real indicators overlap much more. It is a statement that the estimator
does what it says when the assumptions hold, which is the only thing a
simulation can establish.

## How many profiles, and how many kinds of school?

Both counts are chosen, not estimated, so choose them in the open. Fit the grid:

```{r enumerate}
candidates <- enumerate_classes(school_engagement, vars, id = "school",
                                n_profiles = 1:3, n_group_classes = 1:2,
                                n_starts = 4, max_iter = 10000, tol = 1e-10,
                                seed = 1)
```

```{r plot-enumeration}
plot(candidates, criterion = "bic_groups")
```

`summary()` reports which candidate each criterion prefers, rather than picking
one for you:

```{r summary-candidates}
summary(candidates)
```

They do not agree, and that is the normal case rather than a problem with the
data. The BIC family, CAIC, AWE and ICL all choose two profiles in two kinds of
school. AIC and CLC choose three profiles. This is the familiar pattern that
AIC under-penalizes in mixture models and tends to over-extract; the BIC family
is the more conservative read, and it happens to be the structure these data
were simulated from.

Note also that every sample-size-dependent criterion is reported twice, under
`groups` and under `individuals`. A two-level model has two defensible sample
sizes — 60 schools or 720 students — and neither is privileged. Pick one
convention and keep it for the whole analysis rather than choosing per table.

The criteria for the chosen model come back in the shape a model-comparison
table is published in — one row, one column per criterion:

```{r criteria}
information_criteria(fit)
```

Those are the same column names `enumerate_classes()` uses, so this row sits
directly under the grid it was chosen from without renaming anything.

The same numbers are available one row per criterion, which is the form to use
when the convention needs to be a value you can filter on rather than half of a
column name:

```{r criteria-long}
information_criteria(fit, format = "long")
```

Lower is better on every row here, including the first: the long form reports
the deviance rather than the log likelihood, so `value` points one way
throughout. `convention` is `NA` for the three criteria that use no sample size
at all — the question does not arise for them, rather than an answer being
missing.

## The answer: the schools do differ

This is the quantity the second level exists for, and the one a single-level
model cannot report at all.

```{r prevalence}
as.data.frame(fit, what = "profile_probabilities")
```

```{r plot-probabilities}
plot(fit, what = "probabilities")
```

Two kinds of school, in roughly equal numbers, and the balance of profiles
reverses between them: the profile that is the clear majority in one kind is the
clear minority in the other. That reversal is why the lines cross. The pooled
`r round(100 * mean(school_engagement$engaged))`% from the first section is the
average of those two populations and a description of neither.

To see what pooling costs, fit the same measurement model with a single group
class — that *is* the single-level model:

```{r flat}
flat <- multilpa(school_engagement, vars, id = "school",
                 n_profiles = 2, n_group_classes = 1,
                 n_starts = 8, tol = 1e-10, seed = 1)

as.data.frame(flat, what = "profile_probabilities")
```

One row per profile, one number each, and no way to ask the question. The
two-level model also fits better by a wide margin:

```{r compare}
lmr_lrt(flat, fit)
```

The statistic is reported without a p-value on purpose. The Vermunt–Lo–Mendell–
Rubin reference distribution is not implemented here, and a chi-square
approximation to it is not a substitute, so the column is `NA` rather than
misleading.

## How much should you trust it?

Every classification diagnostic comes back from one call:

```{r diagnostics}
diagnostics(fit)
```

That is a summary of four tables, each reachable as a data frame with
`as.data.frame(quality, what = )`. The rest of this section is those tables.

**Are the classes well separated?** Entropy near one means the posterior
assigns students confidently.

```{r entropy}
entropy_table(fit)
```

Individuals separate almost perfectly; schools less so, which is expected —
twelve students is not much evidence about a school.

Entropy is one number for the whole model, which hides where the uncertainty
actually sits. These two views show the distribution behind it. The first is
each student's contribution to the entropy sum, zero being a certain
assignment; the second is the posterior probability of the profile each student
was actually assigned to, where mass piled against one is a clean separation:

```{r plot-entropy, fig.height = 3.6}
plot(fit, what = "entropy")
```

```{r plot-posteriors, fig.height = 3.6}
plot(fit, what = "posteriors")
```

Each ridge is scaled to its own maximum, which is the usual ridgeline
convention: it compares the *shapes* and not the profile sizes, so prevalence
is printed in each label instead. A ridge with a long tail towards the middle
is where a single summary number would have misled you.

**Are the estimates precise?** Standard errors, with the measurement model:

```{r errors}
as.data.frame(fit, data = school_engagement)
```

Robust errors accumulate the score over independent schools rather than
students, which is the right unit when the measurement model may be
misspecified:

```{r robust}
head(parameter_inference(fit, data = school_engagement, vcov_type = "robust"), 6)
```

**Does the model's core assumption hold?** Latent profile analysis assumes the
indicators are independent *within* a profile. Where they are not, the profiles
are partly absorbing a correlation instead of describing a kind of student.

```{r residuals}
bivariate_residuals(fit, data = school_engagement)
```

`observed` and `expected` are both associations on the correlation scale, so
`residual` is their difference and the rows are comparable. All are small here;
a large one would argue for `covariance_model = "full"`.

**Does the profile structure account for the nesting?** The ICCs from the first
section said roughly an eighth of each indicator's variance lay between schools.
Split the same calculation by assigned profile:

```{r descriptives-by-profile}
descriptives(fit, by = "profile")
```

The ICCs collapse towards zero. Within a profile the schools no longer differ,
which is precisely the model's claim: schools differ in the *mix* of students
they serve, not in the students themselves. The raw data agree with the model
without being asked to.

## When students are measured more than once

If the same people are observed on several occasions, the question changes from
*what kinds are there* to *who moves between them*. That needs a different
dataset — in `school_engagement`, `term` indexes twelve different students within
a school, so a "transition" would be between two different people and would mean
nothing.

The package ships one with the right shape:

```{r panel}
str(engagement_panel)
```

120 students, four waves each, and again a truth column: `state` is the state
each row was generated from, with students staying put from one wave to the next
with probability 0.85.

```{r transitions}
moves <- fit_transitions(engagement_panel, c("homework_hours", "participation"),
                         id = "student", time = "wave", n_profiles = 2,
                         n_starts = 6, seed = 1)

transitions(moves, stable = TRUE)
```

`stable = TRUE` marks the diagonal — staying in the same profile — and it
recovers the 0.85 it was generated with to within a few points. Close, not
exact, which is what four waves of 120 students supports.

## What the package refuses

A model family without standard errors says so, by raising a condition you can
catch by class rather than by matching message text:

```{r refuses, error = TRUE}
vcov(moves)
```

Every refusal in the package is a classed condition; `?"multilpa-conditions"`
lists all of them.
