---
title: "Latent class growth analysis: classes of trajectories"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Latent class growth analysis: classes of trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5,
                      fig.height = 4.6, fig.align = "center", dpi = 96)
options(width = 90)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

When the same people are measured repeatedly, **latent class growth
analysis** (Nagin 2005; also called group-based trajectory modelling) finds a
small number of classes, each with its own trajectory, and assigns every
person to the class whose trajectory they follow. Everyone in a class is
assumed to follow that trajectory exactly; what remains is noise.

That assumption is its strength and its limit:

- **Strength.** The model is simple, fast and fits any outcome the regression
  mixture fits: continuous scores, pass/fail, counts, ordered ratings.
- **Limit.** People of the same kind rarely follow one curve exactly. When
  they vary around it, the model can only express that by adding classes, so
  it tends to find more classes than there are kinds (Bauer and Curran 2003).
  The growth mixture model adds random effects for that variation; it is the
  subject of `vignette("growth-mixture")`.

A useful habit is to fit both and compare them (`compare_models()`), as the
growth-mixture vignette does.

| Role | Argument | Here |
|---|---|---|
| Outcome, measured repeatedly | left of `~` | `score` |
| Time (any shape) | right of `~` | `wave`, `wave + I(wave^2)` |
| Person | `id`, with `class_level = "group"` | `student` |
| Kind of outcome | `family` | `"gaussian"`, `"binomial"`, `"poisson"`, `"negative_binomial"`, `"ordinal"` |
| What predicts the class | `membership` | `"motivation"` |

## The data

`growth_scores` is simulated so the answer is known: 300 students tested on up
to seven waves, of three kinds (improving, stable, declining), each student
varying a little around their kind's trajectory, with motivation predicting
the kind. Missed tests (about 12%) need no imputation.

```{r data}
library(latents)
head(growth_scores)
```

## Fitting and choosing the number of classes

Persons are the classified units, so `id` names them and
`class_level = "group"` keeps all of a person's measurements in one class.

```{r enumerate}
enumerate_regressions(score ~ wave, growth_scores, n_classes = 1:5,
                      id = "student", class_level = "group",
                      n_starts = 5, seed = 1)
```

The BIC keeps falling up to five classes, although only three kinds of
student generated the data: the students' own spread around their kind's
trajectory is being absorbed by extra classes. With these data the growth
mixture model, which models that spread, picks three. Here, to keep the
three kinds, the three-class model is read:

```{r fit}
fit <- mixture_regression(score ~ wave, growth_scores, n_classes = 3,
                          id = "student", class_level = "group",
                          membership = "motivation", n_starts = 5, seed = 1)
fit
```

Every table is a data frame from `get_results()`:

```{r tables}
get_results(fit, "classes")
get_results(fit, "membership")
get_results(fit, "recovery", data = growth_scores, truth = "trajectory")
```

## Seeing the classes

Every view of a trajectory fit is one `plot()` call. Each class's trajectory
with its confidence band, the band within which 80% of its students'
observations lie, and the observed class means:

```{r trajectories, eval = has_ggplot}
plot(fit)
```

The same with one panel per class:

```{r facet, eval = has_ggplot}
plot(fit, facet = TRUE)
```

Individual students, chosen across each class's range of classification
certainty: their observations and their class trajectory.

```{r individuals, eval = has_ggplot, fig.height = 6.5}
plot(fit, what = "individuals")
```

The coefficients with their intervals and p-values:

```{r coefficients, eval = has_ggplot}
plot(fit, what = "coefficients")
```

How confidently each student is classified, one bar per student:

```{r classification, eval = has_ggplot, fig.height = 4.2}
plot(fit, what = "classification")
```

## Curved trajectories

Any formula works. A quadratic lets each class bend; compare it with the
straight model fitted the same way (without `membership`):

```{r quadratic}
straight <- mixture_regression(score ~ wave, growth_scores, n_classes = 3,
                               id = "student", class_level = "group",
                               n_starts = 5, seed = 1)
curved <- mixture_regression(score ~ wave + I(wave^2), growth_scores,
                             n_classes = 3, id = "student", class_level = "group",
                             n_starts = 5, seed = 1)
compare_models(straight = straight, curved = curved)
```

These trajectories were generated straight, so the BIC does not reward the
curvature.

```{r quadratic-plot, eval = has_ggplot}
plot(curved)
```

## Outcomes that are not scores

This is where latent class growth analysis is needed: random effects for a
binary, count or ordered outcome need numerical integration, which the growth
mixture model does not do, so trajectories of such outcomes are fitted here.

**Pass or fail.** Whether each test reached 60 points, with a logistic
trajectory per class:

```{r binary}
passing <- transform(growth_scores, passed = as.integer(score >= 60))
pass_fit <- mixture_regression(passed ~ wave, passing, n_classes = 3,
                               family = "binomial", id = "student",
                               class_level = "group", n_starts = 5, seed = 1)
get_results(pass_fit, "coefficients")
```

Each class's probability of passing over the waves, and the coefficients:

```{r binary-plot, eval = has_ggplot}
plot(pass_fit)
plot(pass_fit, what = "coefficients")
```

**Ordered bands.** The score in four ordered bands, with a cumulative-logit
(proportional odds) trajectory per class (McCullagh 1980): each class has its
own thresholds between the bands and its own slope over the waves.

```{r ordinal}
banded <- transform(growth_scores,
                    band = cut(score, c(-Inf, 52, 58, 64, Inf),
                               labels = c("low", "middle", "high", "top"),
                               ordered_result = TRUE))
band_fit <- mixture_regression(band ~ wave, banded, n_classes = 3,
                               family = "ordinal", id = "student",
                               class_level = "group", n_starts = 5, seed = 1)
get_results(band_fit, "coefficients")
```

The plot shows each class's expected band (1 = low to 4 = top), and the
students' own bands over the waves:

```{r ordinal-plot, eval = has_ggplot}
plot(band_fit)
plot(band_fit, what = "individuals")
```

Counts work the same way with `family = "poisson"`, or
`family = "negative_binomial"` when the counts are more spread out than a
Poisson allows.

## Distal outcomes

Do the classes differ on something measured once per person? `three_step()`
compares class means with the bias-corrected three-step method (Vermunt 2010;
Bakk and Vermunt 2016), which accounts for persons being classified with
error. The trajectory model must be fitted without `membership` for this
correction:

```{r distal}
three_step(straight, data = growth_scores, outcome = "motivation",
           contrast = "pairs")
```

## Choices and warnings

- **Latent class growth analysis or a growth mixture model?** If persons of
  the same kind vary around their trajectory -- they almost always do for
  scores -- fit the growth mixture model too and compare them with
  `compare_models()`; prefer it when it fits better. Use latent class growth
  analysis for non-Gaussian outcomes, for a first look, and when within-class
  variation is not plausible.
- **Number of classes.** Read the BIC with the over-extraction above in mind;
  the bootstrap likelihood-ratio test (`enumerate_regressions(bootstrap =)`)
  is the stronger evidence.
- **Starts.** Use several (`n_starts`), and check that the best likelihood is
  reached more than once (the fit reports how many times).

## References

Bakk, Z., & Vermunt, J. K. (2016). Robustness of stepwise latent class
modeling with continuous distal outcomes. *Structural Equation Modeling*, 23,
20--31.

Bauer, D. J., & Curran, P. J. (2003). Distributional assumptions of growth
mixture models: Implications for overextraction of latent trajectory classes.
*Psychological Methods*, 8, 338--363.

McCullagh, P. (1980). Regression models for ordinal data. *Journal of the
Royal Statistical Society B*, 42, 109--142.

Nagin, D. S. (2005). *Group-based modeling of development*. Harvard University
Press.

Vermunt, J. K. (2010). Latent class modeling with covariates: Two improved
three-step approaches. *Political Analysis*, 18, 450--469.
