---
title: "Latent profile analysis: a complete workflow"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Latent profile analysis: a complete workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5,
                      fig.height = 4.6, fig.align = "center", dpi = 96)
options(digits = 3, width = 90)
```

Latent profile analysis (LPA) is a mixture model for continuous indicators.
It assumes that the observations come from a small number of unobserved
subpopulations, called profiles, each with its own mean on every indicator.
This vignette presents a complete analysis: description of the data,
selection of the covariance structure and the number of profiles,
estimation, interpretation, and evaluation of the classification.

```{r package}
library(latents)
set.seed(1)
```

The estimation uses random starting values, so the seed is set once to make
the results reproducible.

## Data

The `srl` data contain the scores of 300 respondents on five constructs of
the Motivated Strategies for Learning Questionnaire: cognitive strategy use,
intrinsic value, self-efficacy, self-regulation, and test anxiety. Each score
is the mean of the construct's items on a scale from 1 to 7. The responses
were generated by a large language model (GPT) in a study of how such models
answer psychological surveys; they serve here as a well-structured example
and are not evidence about students.

```{r data}
indicators <- c("cognitive_strategies", "intrinsic_value", "self_efficacy",
                "self_regulation", "test_anxiety")
descriptives(srl, indicators)
```

## Selecting the covariance structure and the number of profiles

A latent profile model must specify, besides the number of profiles, how
the indicators vary within a profile: whether the variances are equal
across profiles or differ between them, and whether the indicators are
allowed to covary within a profile. The four combinations are named by the
volume, shape and orientation of each profile's covariance matrix, each
equal (`E`) or varying (`V`) across profiles, with `I` for no covariances
(Celeux and Govaert, 1995): `EEI` has equal variances and no covariances,
`VVI` varying variances and no covariances, `EEE` one covariance matrix
shared by all profiles, and `VVV` a covariance matrix for each profile.
`enumerate_lpa()` estimates these four structures with every number of
profiles requested.

```{r enumerate}
models <- enumerate_lpa(srl, indicators, n_profiles = 1:6)
plot(models)
```

The choice of structure matters here because four of the five constructs are
strongly correlated. The structures without covariances (`EEI`, `VVI`) can
reproduce these correlations only by adding profiles, and their BIC continues
to fall up to six profiles. The structures with covariances account for the
correlations directly, and reach far lower BIC values with one or two
profiles. The `EEE` structure attains the lowest values; its candidates are:

```{r choose}
summary(enumerate_lpa(srl, indicators, n_profiles = 1:6, model = "EEE"))
```

The BIC is lowest for one profile, and the two-profile model is 4.7 points
higher; the information criteria therefore do not decide between one and two
profiles. The parametric bootstrap likelihood ratio test compares the two
models directly.

```{r bootstrap}
one <- lpa(srl, indicators, n_profiles = 1, model = "EEE")
fit <- lpa(srl, indicators, n_profiles = 2, model = "EEE")
lrt <- bootstrap_lrt(one, fit)
lrt
```

The test rejects the one-profile model (p = 0.005), and the two-profile model
with the `EEE` structure is retained.

## Estimation

`lpa()` fits a single-level latent profile model. The fitted model:

```{r fit}
fit
```

## Interpretation of the profiles

The profile plot shows the estimated mean of each profile on each indicator
with its 95% confidence interval.

```{r profiles}
plot(fit)
```

The larger profile, comprising 69% of the respondents, has higher means on
cognitive strategy use, intrinsic value, self-efficacy and self-regulation
than the smaller profile, comprising 31%. The two profiles differ little in
test anxiety. The profiles can be described as high and moderate
self-regulated learning. The estimates with their standard errors:

```{r profile-table}
get_results(fit, "profiles")
```

The same means as bars with their confidence intervals:

```{r bars}
plot(fit, what = "bars")
```

The heatmap expresses each mean as a distance from the indicator's grand
mean in standard deviations, which makes the differences comparable across
indicators:

```{r heatmap, fig.height = 3.2}
plot(fit, what = "heatmap")
```

The means summarize distributions. The raincloud plot shows, for the
respondents assigned to each profile, the density of each indicator, a bar
from its first to its third quartile with the median marked, and the
individual scores. It shows how much
the profiles overlap on each indicator: they are well separated on intrinsic
value and self-efficacy, and overlap almost completely on test anxiety.

```{r raincloud, fig.height = 6}
plot(fit, what = "raincloud")
```

## Classification quality

The usefulness of a profile solution depends on how accurately the
observations can be assigned to the profiles. `diagnostics()` reports the
relative entropy, the size of the smallest profile, the average posterior
probabilities, and the largest residual correlation within a profile, and
draws the four classification plots: the effective size of each profile
(the sum of the posterior probabilities of belonging to it), the
distribution of the posterior probability of each respondent's assigned
profile, each respondent's contribution to the classification entropy, and
the average posterior probability matrix, in which each row contains the
respondents assigned to one profile and each column the average posterior
probability they place on each profile.

```{r diagnostics, fig.height = 3.6}
diagnostics(fit)
```

A relative entropy of 0.74 and average posterior probabilities of at least
0.88 indicate an acceptable classification.

## Posterior assignments

`get_results(fit, "assignments")` returns each respondent with the assigned
profile, the posterior probability of every profile, and the classification
uncertainty, in the row order of the data.

```{r assignments}
head(get_results(fit, "assignments"))
```

## Stability across starting values

The likelihood of a mixture model can have several local maxima.
`sensitivity()` re-estimates the model with different random seeds and
reports the maximum reached by each run and the proportion of respondents
assigned to the same profile as in the reported solution.

```{r sensitivity}
sensitivity(fit)
```

## Further analyses

The profiles can be related to covariates with `r3step()` and to distal
outcomes with `three_step()`, both of which correct for classification
error. When the observations are nested in groups, `multilpa()` estimates
the two-level model, in which the groups are in addition grouped into latent
classes, and `enumerate_classes()` compares its numbers of profiles and group
classes; both are presented in `vignette("lpa")`. The ten other covariance
structures, which constrain the shape and orientation separately, are
estimated with `enumerate_lpa(model = "all")`.

## References

Celeux, G., & Govaert, G. (1995). Gaussian parsimonious clustering models.
*Pattern Recognition*, 28(5), 781–793.
