---
title: "Latent class analysis: a complete workflow"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Latent class analysis: a complete workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5,
                      fig.height = 4.6, fig.align = "center", dpi = 96)
options(digits = 3, width = 90)
```

Latent class analysis (LCA) is a mixture model for categorical indicators.
It assumes that the observations come from a small number of unobserved
classes, and that within each class the indicators are independent, each
with its own response probabilities. This vignette presents a complete
single-level analysis: description of the data, selection of the number of
classes, estimation, interpretation, evaluation of the classification, and
the relation of the classes to a distal outcome. The final section extends
the model to two levels. In the output of this package the classes of
observations are labelled `profile_1`, `profile_2`, and so on.

```{r package}
library(latents)
set.seed(1)
```

The estimation uses random starting values, so the seed is set once to make
the results reproducible.

## Data

The `student_esm` data contain 2,582 experience-sampling reports from
university students. At each report a student indicated whether they had
been engaged in each of eight activities, and rated how happy, relaxed,
worried and exhausted they felt. The aim of the analysis is to identify
typical combinations of activities.

```{r data}
activities <- c("time_with_friends", "on_social_media", "tv_video_games",
                "listened_music", "sports", "walking", "reading",
                "part_time_job")
descriptives(student_esm, activities)
```

The reports are nested in students. The first part of this vignette treats
each report as an independent observation, which corresponds to the
conventional single-level model. The nesting is taken into account in the
final section.

## Selecting the number of classes

`enumerate_lca()` estimates models with one to six classes and returns the
results in one table.

```{r enumerate}
models <- enumerate_lca(student_esm, activities, n_classes = 1:6)
plot(models)
```

```{r choose}
summary(models)
```

The BIC is lowest for four classes, 13.2 points below the three-class
model. The fourth class, however, contains 1.3% of the reports (about 32),
which describe many activities at once; a class of this size is poorly
supported, and the ICL, which penalizes classification uncertainty, is lower
for three classes than for four. The ICL is lowest for one class, which has
no uncertainty to penalize, and among the models with more than one class it
is lowest for two; the two-class solution, however, has a relative entropy
below 0.5 and therefore separates the reports poorly. The three-class model
is retained.

## Estimation

`lca()` fits a single-level latent class model.

```{r fit}
fit <- lca(student_esm, activities, n_classes = 3)
fit
```

## Interpretation of the classes

The heatmap shows, for each class, the estimated probability that the
activity was reported.

```{r heatmap, fig.height = 3.4}
plot(fit, what = "heatmap")
```

The largest class, comprising 58% of the reports, is characterized by
screen-based activities: social media, television and video games, and
music. The smallest class, 4% of the reports, consists almost entirely of
reports of working at a part-time job. The third class, 39% of the reports,
has elevated probabilities of spending time with friends, walking and
sport, and a probability of social media use close to zero. The same
probabilities are shown below as profiles across the activities, with 95%
confidence intervals where the probability is not on its boundary.

```{r responses}
plot(fit, what = "responses")
```

The probabilities with their standard errors. Several are estimated at zero
or one; a standard error is not defined on the boundary of the parameter
space, so those are held at their bound and have none.

```{r response-table}
head(get_results(fit, "responses"), 8)
```

## Classification quality

`diagnostics()` reports the classification measures and draws the four
classification plots: the effective size of each class, the distribution of
the posterior probability of each report's assigned class, each report's
contribution to the classification entropy, and the average posterior
probability matrix, which shows which classes are confused with which.

```{r diagnostics, fig.height = 3.6}
diagnostics(fit)
```

The relative entropy of 0.66 indicates a moderately clear classification.

## Classes and a distal outcome

`three_step()` estimates the mean of an outcome within each class. It uses
the BCH method, which weights each observation by its posterior class
probabilities and corrects for classification error (Bolck, Croon and
Hagenaars, 2004). The outcome here is the happiness rating.

```{r three-step}
happiness <- three_step(fit, student_esm, outcome = "happy")
happiness
```

Happiness is highest in the class of reports away from screens (5.30 on
the seven-point scale), and lower in the screen-based class (4.63) and the
work class (4.53).

## Extension to two levels

Each student contributes many reports, so the reports are not independent.
When the student identifier is supplied as `id`, `multilca()` estimates a
two-level model: the classes of reports are retained, and the students are
in addition grouped into latent classes that differ in the proportion of
their reports belonging to each class.

```{r two-level, message = FALSE}
two_level <- multilca(student_esm, activities, id = "student",
                      n_profiles = 3, n_group_classes = 2)
get_results(two_level, "profile_probabilities")
```

The two-level model is presented in detail in `vignette("lca")`.

## References

Bolck, A., Croon, M., & Hagenaars, J. (2004). Estimating latent structure
models with categorical variables: One-step versus three-step estimators.
*Political Analysis*, 12(1), 3–27.
