---
title: "Two levels of engagement: kinds of student, and the schools that mix them"
author: "multilpa"
date: "2026-09-20"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: flatly
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE, comment = "#>", fig.width = 7, fig.height = 5,
  dpi = 150, out.width = "100%"
)
options(digits = 4, width = 96)
set.seed(2026)
```

## Where this sits

The README fits `multilpa()` to show the syntax. `vignette("multilpa")` walks
the argument for the second level from a standing start. Neither is the thing an
analyst actually arrives wanting: one model, taken end to end, with every claim
followed by the check that would have caught it if it were wrong.

This tutorial is that pass. It is deliberately a *verification* document. Each
section states what the model says and then asks what would have to be true for
that statement to be an artefact — of the optimiser, of a class count nobody
chose in the open, of an assumption the model makes silently, or of a posterior
too blurred to act on.

Every number below is computed by this document when it renders, from data that
ship with the package. Nothing here depends on a file you do not have.

```{r library}
library(multilpa)
```

Two datasets carry the whole tutorial.

```{r data}
str(school_engagement)

str(engagement_panel)
```

`school_engagement` is the cross-sectional one: 720 students, twelve in each of
60 schools, with three engagement indicators. `engagement_panel` is the
repeated-measures counterpart: 120 students seen at four waves. Both are
simulated, and both ship a column recording the kind or state each row was
*generated* from — `engaged` and `state`. A real study never has that column.
Here it exists so that the model's answer can be checked against the thing that
produced the data, which is the one check no amount of fit statistics can
substitute for.

### Why the second level is not optional

A single-level latent profile analysis of `school_engagement` treats the twelve
students in a school as twelve unrelated people. That is wrong in two separate
ways, and only one of them is the familiar objection.

The familiar one is that the rows are not independent, so every standard error
is too small. The second is more interesting: a single-level model cannot
express the question worth asking. "Which schools serve mostly disengaged
students?" is not a statement about any one student. It is a statement about the
*distribution* of students belonging to a school, and a model with only one
level has nowhere to put it.

Vermunt's (2003) nonparametric formulation puts it in a second set of latent
classes. Schools belong to a class; a class is nothing more than a different mix
of the same student profiles. The measurement model — what an engaged student
looks like — is shared by everyone, so the classes cannot cheat by redefining
the profiles for each group. They differ only in how often each profile occurs.

### The route

1. **Before the model** — whether the grouping carries anything at all.
2. **The model as a whole** — whether it converged, on what, and whether a
   larger model can be found twice.
3. **The profiles** — what one student looks like, and whether the model found
   what made the data.
4. **The school classes** — what a whole school looks like, and what pooling
   costs.
5. **Order** — what these data can and cannot say about *when*, and what
   ignoring order costs when order exists.
6. **Testing the assumption the model makes** — whether the indicators are
   really independent inside a profile.
7. **How much to trust the assignments** — entropy, classification quality, and
   the one warning this fit raises.
8. **What this model cannot tell you.**

### The verbs used

| verb | the question it answers |
|---|---|
| `descriptives()` | is there any between-group variance to model |
| `multilpa()` | estimate the two-level model |
| `summary()` | the whole fit on one screen, restarts included |
| `information_criteria()` | what the parameters cost, under both sample-size conventions |
| `as.data.frame()` | any part of the fit, as a tidy table |
| `plot()` | the measurement model, the class mixes, the diagnostics |
| `assignments()` | every observation beside a column of your own |
| `lmr_lrt()` | whether the larger model is worth its parameters |
| `sequences()` / `sequence_summary()` | the assignments in occasion order |
| `fit_transitions()` / `transitions()` | movement between profiles over time |
| `bivariate_residuals()` | whether the independence assumption holds |
| `diagnostics()` | entropy and classification quality in one call |
| `parameter_inference()` | standard errors, with the qualification attached |

## 1. Before the model: does the grouping carry anything?

This is the step most often skipped, and it is the one that decides whether the
rest of the document is worth writing. A two-level model exists to tell groups
apart. If the groups do not differ, it has nothing to find, and it will still
return a fit.

Start with the question the data can answer directly, because they are
simulated and carry the truth column:

```{r hook}
per_school <- aggregate(engaged ~ school, data = school_engagement, FUN = mean)

summary(per_school)
```

Pooled across the sample, **`r with(school_engagement, round(100 * mean(engaged)))`%**
of students are the engaged kind — a tidy headline number. School by school the
share runs from **`r with(per_school, sprintf("%.2f", min(engaged)))`** to
**`r with(per_school, sprintf("%.2f", max(engaged)))`**. The sample average
describes no school in the data. Any analysis that pools the students and reports one
number has averaged away the thing worth knowing.

That hook used `engaged`, which a real study would not have. The check that
works without it is the intraclass correlation — the share of each indicator's
variance lying *between* schools rather than within them:

```{r descriptives}
vars <- c("homework_hours", "participation", "interest")

descriptives(school_engagement, vars = vars, id = "school")
```

Around an eighth of the variance in each indicator sits between schools. That is
modest, and it is enough: had these come back near zero, the schools would not
differ and the second level would be estimating noise. One line, before any
model.

## 2. The model as a whole

`multilpa()` takes the indicators, the column identifying the second level, and
the two class counts. `n_starts` matters more than it looks: mixture likelihoods
are multimodal, and a single start reports whichever hill it happened to climb.

```{r fit}
fit <- multilpa(school_engagement, vars = vars, id = "school",
                n_profiles = 2, n_group_classes = 2,
                n_starts = 20, seed = 2026)

summary(fit)
```

Every start reached the same maximum. That unanimity is the single most
reassuring line on the page: where multiple optima exist, starts scatter across
them, and a clean replication says the solution is not an artefact of where the
search began. No start hit a variance boundary, no start failed to converge, and
neither profile nor school class has an effective membership below one, so
nothing has collapsed.

```{r criteria}
information_criteria(fit)
```

The criteria are reported under both sample-size conventions because, for a
multilevel mixture, "n" is genuinely ambiguous — 60 schools or 720 students.
They are alternatives to be chosen between and declared, not a menu to pick the
agreeable one from. Pick one and keep it for the whole analysis.

### Could a third profile be found twice?

Two profiles was a choice, so make the alternative argue for itself.

```{r three-profiles}
three <- multilpa(school_engagement, vars = vars, id = "school",
                  n_profiles = 3, n_group_classes = 2,
                  n_starts = 20, seed = 2026)

summary(three)
```

The three-profile solution reaches a higher likelihood, and its AIC is lower.
Read the rest of the output before believing that.

The best likelihood is found by a minority of the starts; the others settle on
at least three other optima, and one exhausts its iteration budget without
converging at all. The third profile it finds is a spike: an effective
membership of about ten students out of 720, whose participation standard
deviation is a small fraction of the other two profiles'. That is the classic
signature of a mixture component fitting a handful of points rather than a kind
of student.

```{r three-criteria}
information_criteria(three)
```

Set that table beside section 2's. AIC, SABIC on the group convention, and CLC
under both conventions prefer three profiles; KIC, BIC, CAIC, AWE and ICL prefer
two, and SABIC reverses when the convention changes. This is the familiar
pattern — AIC under-penalises in mixture models and tends to over-extract, and
CLC, which charges for classification uncertainty rather than for parameters, is
rewarding a class that separates cleanly precisely because it is a spike. ICL
charges for both and keeps two. The disagreement is the normal case, not a
problem with the data.

The restart column breaks the tie, and it should. A criterion cannot prefer a
solution the optimiser locates in a minority of attempts. Two profiles is the
model this document reports, and the reason is the diagnostic, not the
criterion.

## 3. The profiles: what one student looks like

The profiles are the measurement model, and they are shared across every school
class. A profile describes one student.

```{r profiles-table}
as.data.frame(fit)
```

```{r profiles-plot}
plot(fit)
```

The picture is a level ordering, not a set of shapes. One profile sits high on
all three indicators at once; the other sits low on all three, and the gap
between them is between roughly two and a half and three and a half times the
within-profile standard deviation on every indicator. They separate in the same
direction every time.

That is worth stating plainly, because it is the kind of result that is easy to
over-read. These data contain no kind of student who does a great deal of
homework while participating little, or who is highly interested but does
nothing. Engagement here moves as a block. Two profiles is the right number for
describing *how much*, not *in what way*.

```{r heatmap, fig.height = 3.4}
plot(fit, what = "heatmap")
```

The heatmap says the same thing in standard deviations from each indicator's
grand mean, which is the view to use when the indicators are on different
scales or when there are more of them than a line plot can label.

### Did it find what made the data?

This is the check simulated data exist for. `assignments()` puts each student's
estimated profile in the same row as the column that generated it, so nothing
has to be lined up by hand:

```{r recovery}
xtabs(~ profile + engaged, data = assignments(fit, data = school_engagement))
```

One student out of 720 on the wrong side. That is not a claim about this
package's accuracy in general — the profiles here are far apart by construction,
and real indicators overlap much more. It is a statement that the estimator does
what it says when the assumptions hold, which is the only thing a simulation can
establish.

Verbs that bring in a column the model never saw get their own function with its
own documented arguments, rather than another `as.data.frame(what =)` value.
`assignments()` owns the alignment: a frame with the wrong number of rows is
refused, and a column its output would overwrite is an error rather than a
silent replacement.

## 4. The school classes: what a whole school looks like

Everything so far could have come from a single-level model. The second level is
where the students belonging to one school are tied together, and a school class
is defined by nothing except the mix of profiles its students produce.

```{r probabilities-table}
as.data.frame(fit, what = "profile_probabilities")
```

```{r probabilities-plot}
plot(fit, what = "probabilities")
```

Two kinds of school in roughly equal numbers, and the balance reverses between
them: the profile that is the clear majority in one kind is the clear minority
in the other. That reversal is why the lines cross, and it is the quantity a
single-level model cannot report at all.

Read the size of the minority carefully, because it is the honest limit of the
finding. Neither kind of school is homogeneous. Roughly three students in ten
in each kind belong to the *other* profile, and no school class is anywhere near
pure. These are schools that differ in their mix, not schools that contain only
one kind of student — which is exactly what the model claims and exactly what
the data were generated from.

To see what pooling costs, fit the same measurement model with a single school
class. That *is* the single-level model:

```{r flat}
flat <- multilpa(school_engagement, vars = vars, id = "school",
                 n_profiles = 2, n_group_classes = 1,
                 n_starts = 20, seed = 2026)

as.data.frame(flat, what = "profile_probabilities")
```

One row per profile, one number each, and no way to ask the question. The
two-level model also fits better by a wide margin:

```{r flat-criteria}
information_criteria(flat)
```

Every criterion, under both conventions, prefers the two-class model over this
one — which is a cleaner verdict than section 2's three-profile comparison
produced, and the contrast is the point: the criteria agree when the extra
structure is real.

```{r compare}
lmr_lrt(flat, fit)
```

The statistic is reported without a p-value on purpose. The
Vermunt–Lo–Mendell–Rubin reference distribution is not implemented here, and a
chi-square approximation to it is not a substitute, so the column is `NA` rather
than misleading. The statistic and its adjusted version are large against two
degrees of freedom, which is a reason to take the second level seriously; it is
not a test, and this document does not report it as one. The criteria above are
the verdict this document rests on.

### The raw data agree, without being asked to

Section 1 found roughly an eighth of each indicator's variance lying between
schools. If the model's claim is right — schools differ in the *mix* of students,
not in the students themselves — then that between-school variance should
disappear once you condition on the profile. Split the same calculation by
assigned profile:

```{r descriptives-by-profile}
descriptives(fit, by = "profile")
```

The intraclass correlations collapse to zero or below. Within a profile the
schools no longer differ. The model was never shown this calculation; it is the
data agreeing with the model on a quantity the model did not fit.

```{r group-posteriors}
head(as.data.frame(fit, what = "group_posteriors"))
```

One row per school, with the modal class, the school's own contribution to the
likelihood, and the posterior behind the assignment. The posterior is the honest
object here: the modal class is a summary of it, not a fact about the school.

## 5. Order: what these data can and cannot say about *when*

A school class is a probability mix, and a mix is an average over careers that
may look nothing like each other. The obvious next question is whether the mix
conceals a trajectory.

For `school_engagement` the answer is that the question does not apply, and the
package says so rather than drawing something meaningless:

```{r no-time, error = TRUE}
sequences(fit)
```

That refusal is correct. `term` in `school_engagement` indexes twelve *different*
students within a school, so a "transition" from term 1 to term 2 would be a
transition between two different people. The column exists; the ordering does
not. The fit was never given `time =`, and the verb will not invent one.

`engagement_panel` has the structure the question needs: the same 120 students at
four waves. Two models can be fitted to it, and the difference between them is
the whole point of this section.

### The model that stores order but does not use it

`multilpa()` with `time =` fits exactly the model of sections 2 to 4 — students
nested in a unit, the unit's class defined by its mix of profiles — and stores
the occasion so the assignments can be read back in order afterwards. Here the
unit is a student and the observations are that student's waves.

```{r panel-fit}
panel_vars <- c("homework_hours", "participation")

panel_fit <- multilpa(engagement_panel, vars = panel_vars, id = "student",
                      n_profiles = 2, n_group_classes = 2, time = "wave",
                      n_starts = 20, seed = 2026)

as.data.frame(panel_fit, what = "profile_probabilities")
```

Two kinds of student: one mostly in the low profile, one almost always in the
high one. Now lay the assignments out in wave order:

```{r sequences}
head(sequences(panel_fit, format = "wide"), 8)
```

```{r sequences-plot, fig.width = 8, fig.height = 7}
plot(panel_fit, what = "sequences")
```

Each row is one student, each column one wave, with students blocked by their
class. Read the wide table above rather than the picture if you want to check
this: the waves within a row come in runs. A student holds a profile for a
while and then switches, rather than alternating. The model that produced those
assignments knows nothing about that — it scored wave 4 the same way it scored
wave 1 — which is precisely the gap the next model closes.

```{r sequence-summary}
sequence_summary(panel_fit)
```

Every student contributes four complete waves with no gaps, so nothing in what
follows is about attrition.

### The model that uses order

`fit_transitions()` estimates first-order transition probabilities between
profiles across the occasions, by a forward–backward EM algorithm. On these data
it costs exactly the same number of parameters as the mixture above.

```{r transitions}
moves <- fit_transitions(engagement_panel, vars = panel_vars, id = "student",
                         time = "wave", n_profiles = 2,
                         n_starts = 20, seed = 2026)

transitions(moves, stable = TRUE)
```

`stable = TRUE` marks the diagonal — staying in the same profile from one wave
to the next. These data were generated with a persistence of 0.85, and both
diagonal entries land within a few points of it. Close, not exact, which is what
four waves of 120 students supports.

Now compare the two models on the criteria. Same data, same indicators, same
number of parameters; the only difference is whether the ordering is used:

```{r panel-criteria}
information_criteria(panel_fit)
```

```{r moves-criteria}
information_criteria(moves)
```

The ordered model reaches a higher likelihood for the same eleven parameters,
and every criterion — under either sample-size convention — falls. That is the
cost of ignoring order, measured rather than asserted. It is also the honest
answer to the objection people raise about latent profile analysis: the mixture
is not wrong, it answers a different question, and when order genuinely exists,
the model that uses it is the better description.

### And the school classes it does not need

A latent transition model can also carry a second level, letting different kinds
of student have different transition matrices. Ask for one here and watch what
happens:

```{r transitions-two}
moves_two <- fit_transitions(engagement_panel, vars = panel_vars, id = "student",
                             time = "wave", n_profiles = 2, n_group_classes = 2,
                             n_starts = 20, seed = 2026)

summary(moves_two)
```

```{r transitions-two-criteria}
information_criteria(moves_two)
```

Three things went wrong at once. One estimated persistence is 1 to four decimal
places — an absorbing state that no student ever leaves — which is a parameter
sitting on its boundary, not a finding. The likelihood barely moves for the four
extra parameters, so every criterion but one is worse than the one-class
model's; the single exception, CLC under the individual convention, is lower by
about one point, which is noise. And the best solution no longer replicates
across every start.

The second level is real in `school_engagement` and is not real here. The
package reports both outcomes the same way, and it is the reader's job to look
at the restart line and the boundary before reading the transition matrix.

## 6. Testing the assumption the model makes

Everything above rests on an assumption that has not yet been examined. A
profile model says that, once you know which profile a student belongs to, their
indicators are independent of one another: whatever made a student do a lot of
homework is the profile, not some separate thing that also made them
participate. The default `covariance_model = "diagonal"` writes that down as a
constraint, fixing every within-profile covariance at zero.

Nothing in the fit reports whether that constraint is reasonable. A model can be
badly wrong in exactly this way and still converge, replicate across every
start, and return a respectable entropy. `bivariate_residuals()` is the check.

```{r residuals}
bivariate_residuals(fit)
```

One row per pair of indicators per profile, worst first. `observed` is the
association still present inside the profile, `expected` is what the fitted
covariance model implies for it, and `residual` is the gap. Both are on the
correlation scale, so the rows are comparable.

The largest gap anywhere is `r with(head(bivariate_residuals(fit), 1), round(abs(residual), 3))`
in absolute value, and no pair is significant at the conventional level even
before any correction for the six tests. The assumption holds.

That is a result, not a non-result, and it is worth saying out loud: this
diagnostic can acquit. Running it and finding nothing is what licenses the rest
of the document; skipping it and finding nothing are not the same thing.

The same residuals can be pooled rather than split by profile, which asks
whether the profile structure as a whole has absorbed the associations:

```{r residuals-overall}
bivariate_residuals(fit, by = "overall")
```

### What the unconstrained model buys

The comparison that settles it is with the model that estimates the covariances
instead of fixing them.

```{r full-covariance}
full <- multilpa(school_engagement, vars = vars, id = "school",
                 n_profiles = 2, n_group_classes = 2,
                 n_starts = 20, covariance_model = "full", seed = 2026)

information_criteria(full)
```

Six extra parameters, and the likelihood improves by around two points. Set that
beside section 2's table: every criterion is worse except CLC, which is lower by
about four points under either convention — and CLC is the criterion that
rewards clean classification rather than parsimony, so it was never going to
object to six parameters that change no assignment.

Estimating associations that are not there costs parameters and buys nothing,
which is the mirror image of the usual warning and just as important:
`covariance_model = "full"` is the right choice when the residuals say so, and
the wrong choice when they do not.

Note one thing in passing, because it will matter in section 8:

```{r full-profiles}
as.data.frame(full)
```

The profiles are the same two profiles, and the labels have swapped. Nothing in
the likelihood distinguishes "profile 1" from "profile 2".

## 7. How much to trust the assignments

A profile solution can be well estimated and still be too blurred to act on.
Every classification diagnostic comes back from one call:

```{r diagnostics}
diagnostics(fit)
```

That is a summary of four tables. The rest of this section is those tables, plus
the one qualification this fit raises.

### The warning this fit raises

Ask for standard errors and the package pushes back:

```{r inference}
head(parameter_inference(fit), 6)
```

That warning is `multilpa_unconverged`, and it is worth believing rather than
muffling. The default convergence tolerance is set for the *likelihood*, and the
likelihood is flat near its maximum: stopping when it stops improving can leave
the parameters at a point where the score is still appreciably far from zero.
Wald standard errors are built from the curvature at a point where the score is
assumed to be zero, so that residual score is exactly the thing that invalidates
them. The estimates are fine; the intervals around them are what is being
questioned.

The fix is one argument: a tighter likelihood tolerance, which drives the
residual score down with it.

```{r precise}
precise <- multilpa(school_engagement, vars = vars, id = "school",
                    n_profiles = 2, n_group_classes = 2,
                    n_starts = 20, tol = 1e-10, seed = 2026)

head(parameter_inference(precise), 6)
```

No warning — and, this being the point, every estimate, standard error and
confidence bound is identical to the precision printed above. Only one test
statistic moves, in its final digit.

That is what makes the warning worth having rather than annoying. It was never a
claim that the estimates were wrong; it was the package declining to certify an
interval whose derivation assumed something it could not verify. Compare what
would have happened had it stayed quiet: the same numbers, no flag, and no way
for a reader to know the difference between an interval that was checked and one
that was not.

Robust errors accumulate the score over independent schools rather than
students, which is the right unit when the measurement model may be
misspecified:

```{r robust}
head(parameter_inference(precise, vcov_type = "robust"), 6)
```

### Entropy and classification quality

```{r entropy}
entropy_table(fit)
```

Relative entropy is
`r with(subset(entropy_table(fit), level == "individuals"), round(relative_entropy, 3))`
for students and
`r with(subset(entropy_table(fit), level == "groups"), round(relative_entropy, 3))`
for schools. Both are high enough that the modal assignments mean something, and
the ordering between them is the interesting part.

Students separate more sharply than the schools they sit in — the opposite of
what one might expect from the usual "more data, sharper answer" intuition, and
it is not a defect. A student is classified from three indicators measured
directly on them. A school is classified only from the *mix* of its twelve
students, and two kinds of school whose mixes are 27% and 71% will produce
overlapping samples of twelve fairly often. Twelve students is not much evidence
about a school.

```{r classification}
classification_table(fit, level = "both")
```

The average posterior for the student profiles is near one, and the odds of
correct classification are in the hundreds — against the value of 1 that a
useless classifier would give. For schools the same quantities are lower, and
consistently so, which is the same story the entropy told.

Note also that `n_modal` and `estimated_n` agree closely at the student level.
Where a model separates poorly these two diverge, because the modal count claims
a certainty the posteriors do not support. At the school level they diverge a
little — 30 and 30 by modal assignment against a fractional split — which is
that weaker separation showing up in a second place.

The two distributions behind the entropy number, which one number cannot show:

```{r plot-entropy, fig.height = 3.6}
plot(fit, what = "entropy")
```

```{r plot-posteriors, fig.height = 3.6}
plot(fit, what = "posteriors")
```

Each ridge is scaled to its own maximum, which is the usual ridgeline
convention: it compares *shapes* rather than profile sizes, so prevalence is
printed in each label instead. A ridge with a long tail towards the middle is
where a single summary number would have misled you.

## 8. What this model cannot tell you

Four limits are worth naming, because none of them is something the output announces on its own.

**The labels are arbitrary.** Section 6 showed it happening: the same two
profiles came back from the full-covariance fit with their numbers swapped.
Nothing in the likelihood distinguishes them, so any comparison between two fits
has to align the labels first, and no substantive claim should ever be attached to a profile number.

**The cross-sectional model has no time in it.** `school_engagement` carries no
ordering at all — `sequences()` refused in section 5, and it was right to. Where
an ordering does exist, section 5 measured what ignoring it costs: on
`engagement_panel`, the same eleven parameters spent on transition
probabilities instead of a mixture of students bought a better fit on every
criterion. A mixture model describes an overall mix, not a trajectory, and
"students who start strong and fade" can land in the same class as "students who
are consistently middling".

**Twelve students is thin evidence about a school.** Section 7's entropy
ordering is not a quirk of this fit; it is a structural feature of a two-level
mixture with small groups. A school's class is inferred only from the mix of its
members, and small groups give overlapping mixes. Reporting a school's modal
class without its posterior overstates what is known.

**These data are simulated.** That is what allowed section 3 to check the
recovery and section 5 to check the persistence against 0.85. It is also the
reason none of the numbers here is a finding about schools. They are findings
about the estimator.

### When to use which

Use `n_group_classes = 1` when the grouping exists only as a nuisance — the
model then reduces to an ordinary latent profile analysis that still respects
the clustering. Raise it above one when the question is about the units rather
than the observations, as in sections 2 to 4. Use `fit_transitions()` when the
occasions are ordered and the question is about movement. Use
`fit_random_intercept()` when group differences are better described as a
continuum than as types, and compare the two on the information criteria rather
than assuming.

And run `descriptives()` before all of them. A model built to tell groups apart
will return a fit whether or not the groups differ.

## References

Vermunt, J. K. (2003). Multilevel latent class models. *Sociological
Methodology*, 33, 213–239.
