---
title: "Two levels of engagement: profiles within courses, students across them"
author: "multilpa"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: flatly
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE, comment = "#>", fig.width = 7, fig.height = 5,
  dpi = 150, out.width = "100%", message = FALSE, warning = FALSE
)
options(digits = 4)
```

## Where this sits

The package README fits `multilpa()` to synthetic data to show the syntax, and
`validation/simulab-recovery/` establishes that the estimator recovers a model
whose truth is known. Neither answers the question an analyst actually arrives
with: what does a two-level latent profile model say about a real course
platform, and how much of what it says can be trusted?

This tutorial is that pass. One dataset — 1,393 course enrolments by 106
students across 31 courses — fitted end to end, with three profiles at the
enrolment level and three latent classes at the student level.

### Why the second level is not optional

A single-level latent profile analysis of these data treats a student's fifteen
enrolments as fifteen unrelated people. That is wrong in two separate ways, and
only one of them is the familiar objection.

The familiar one is that the rows are not independent, so every standard error
is too small. The second is more interesting: a single-level model cannot
express the question worth asking. "Which students are consistently disengaged?"
is not a statement about any one enrolment. It is a statement about the
*distribution* of enrolments belonging to a student, and a model with only one
level has nowhere to put it.

Vermunt's (2003) nonparametric formulation puts it in a second set of latent
classes. Students belong to a class; a class is nothing more than a different
mix of the same enrolment profiles. The measurement model — what a
high-engagement enrolment looks like — is shared by everyone, so the classes
cannot cheat by redefining the profiles for each group. They differ only in how
often each profile occurs.

### The route

1. **Preparing the indicators** — why engagement has to be measured against
   classmates rather than against the cohort.
2. **The model as a whole** — whether it converged, on what, and how many
   parameters bought how much likelihood.
3. **The profiles** — what a single enrolment looks like, and why the answer is
   about level rather than shape.
4. **The student classes** — what a whole student looks like, and the one
   asymmetry that carries the finding.
5. **The sequences themselves** — what a student's career looks like once the
   assignments are laid out in course order.
6. **Testing the assumption the model makes** — whether the indicators are
   really independent inside a profile, and what it costs when they are not.
7. **How much to trust the assignments** — entropy and classification quality.
8. **What this model cannot tell you.**

### The verbs used

| verb | the question it answers |
|---|---|
| `multilpa()` | estimate the two-level model |
| `summary()` | the whole fit on one screen |
| `information_criteria()` | what the parameters cost, under both sample-size conventions |
| `entropy_table()` | how sharply units separate |
| `classification_table()` | whether the assignments are trustworthy |
| `as.data.frame()` | any part of the fit, as a tidy table |
| `plot()` | the measurement model and the class mixes |
| `sequences()` | the assignments in course order |
| `sequence_summary()` | how much data each student class contributes |
| `bivariate_residuals()` | whether the independence assumption holds |

Companion material: `validation/simulab-recovery/README.md` for how well these
estimates recover a known truth, and `?multilpa` for the estimation contract.

## 1. Preparing the indicators

The platform records nine activity counts per enrolment. One of them,
`Total_events`, is the arithmetic sum of four others — regressing it on those
four returns an R-squared of 0.99983 with every slope at 1.00, and including all
nine leaves the correlation matrix with a condition number above sixty thousand.
It is dropped, leaving eight.

The remaining eight cannot be used in their recorded units either, and the
reason is substantive rather than technical. A busy course inflates every count
for everyone enrolled in it, so raw activity confounds "this student worked
hard" with "this course demanded a lot". Standardising inside each course
removes the course from the comparison: a value of `+1` then means one standard
deviation above the classmates who sat the same course.

```{r prepare}
library(multilpa)

# Point this at your copy of the file.
data_path <- file.path("~", "My Drive (saqr@saqr.me)", "Git", "Qudent_26",
                       "lecture_data_with_LCA_labels_percentiles.csv")

#' Course-standardised engagement indicators
#'
#' @param path Path to the platform export.
#' @return A base `data.frame` with one row per enrolment and the columns
#'   `User`, `Sequence`, `course_id`, and one `*_z` column per indicator, each
#'   standardised within its course. Enrolments in a course with no second
#'   student have no reference group and are dropped.
prepare_engagement <- function(path) {
  stopifnot("`path` must name a readable file" = file.exists(path))
  sources <- c(browse = "Frequency_course_browse", lectures = "Frequency_lecture_viewed",
               forum_read = "Frequency_forum_consume", forum_post = "Frequency_forum_contribute",
               active_days = "Active_days", sessions = "Session_count",
               entropy = "Course_browse_entropy", duration = "Total_duration")
  lecture <- utils::read.csv(path, stringsAsFactors = FALSE)
  stopifnot(
    "the export must contain every indicator" = all(sources %in% names(lecture)),
    "`User` and `Sequence` must identify an enrolment uniquely" =
      !anyDuplicated(paste(lecture$User, lecture$Sequence, sep = "_"))
  )
  # A course whose indicator does not vary gives everyone a deviation of exactly
  # zero, which is the honest value; a course of one student has no reference
  # group at all and yields NA, which is dropped rather than filled.
  scale_within <- function(value, key) {
    centred <- value - stats::ave(value, key, FUN = mean)
    spread <- stats::ave(value, key, FUN = stats::sd)
    ifelse(spread > 0, centred / spread, ifelse(is.na(spread), NA_real_, 0))
  }
  scaled <- as.data.frame(lapply(lecture[sources], scale_within, key = lecture$course_id))
  names(scaled) <- paste0(names(sources), "_z")
  prepared <- data.frame(lecture[c("User", "Sequence", "course_id")], scaled,
                         row.names = NULL)
  prepared <- prepared[stats::complete.cases(prepared), , drop = FALSE]
  row.names(prepared) <- NULL
  prepared
}

engagement <- prepare_engagement(data_path)
str(engagement)
```

Eleven columns: the student, the position of the course in that student's
sequence, the course, and eight standardised indicators. One enrolment was
dropped — it belonged to the only student in its course, so there was no one to
compare against.

```{r indicators}
indicators <- c("browse_z", "lectures_z", "forum_read_z", "forum_post_z",
                "active_days_z", "sessions_z", "entropy_z", "duration_z")
```

Naming the indicators explicitly, rather than matching a pattern over the column
names, means a later column added to the export cannot silently join the model.

## 2. The model as a whole

`multilpa()` takes the indicators, the column identifying the second level,
and the two class counts. `n_starts` matters more than it looks: mixture
likelihoods are multimodal, and a single start reports whichever hill it
happened to climb.

`covariance_model = "full"` is not the default and is not cosmetic here. The
default assumes indicators are independent within a profile, and on these data
that assumption is false by a wide margin. Section 6 shows the evidence and what
ignoring it costs; the model is fitted correctly from the start so that every
number below comes from the model worth reporting.

```{r fit}
fit <- multilpa(engagement, indicators = indicators, group = "User",
                n_profiles = 3, n_group_classes = 3, time = "Sequence",
                covariance_model = "full", n_starts = 50, seed = 2026)
summary(fit)
```

All 50 starts reached the same maximum, at a log likelihood of -7900.55 with 140
free parameters. That unanimity is the single most reassuring number on the
page: where multiple optima exist, starts scatter across them, and a 50-of-50
replication says the solution is not an artefact of where the search began. No
variance sits on its lower bound, and no class has an effective membership below
one, so nothing has collapsed.

The parameter count is large because a full residual covariance costs 36 numbers
per profile rather than 8. It buys a great deal: the same three profiles fitted
under the independence assumption reach only -11083.49, so estimating the
associations is worth 3,183 log-likelihood points for 84 parameters, and the BIC
on the group count falls from 22428 to 16454.

```{r criteria}
information_criteria(fit)
```

The information criteria are reported under both sample-size conventions
because, for a multilevel mixture, "n" is genuinely ambiguous — 106 students or
1,393 enrolments. The recovery study in `validation/simulab-recovery/` shows the
two conventions selecting different structures at small sample sizes. They are
alternatives to be chosen between and declared, not a menu to pick the agreeable
one from.

Both conventions here prefer more than three profiles: BIC keeps falling at four
and at five. Neither is used, and the reason is in the restart diagnostics rather
than the criteria. Fitted from 30 starts, the four-profile solution is found
once, the five-profile solution is found once and sits on a variance boundary,
and the three-profile solution is found by every start. A criterion cannot
prefer a solution the optimiser cannot locate twice.

## 3. The profiles: what one enrolment looks like

The profiles are the measurement model, and they are shared across every student
class. A profile describes one enrolment: one student, in one course, for one
term.

```{r profiles-table}
as.data.frame(fit)
```

The table is the whole measurement model: one row per profile and indicator,
with the estimated mean, variance and standard deviation. Reading down the mean
column, the three profiles separate on every one of the eight indicators, and
they separate in the same direction each time.

```{r profiles-plot, fig.width = 9}
plot(fit)
```

The picture is a level ordering, not a set of shapes. Profile 1, holding 27% of
enrolments, sits between 0.49 and 1.03 standard deviations above the course mean
on all eight indicators at once. Profile 2, also 27%, sits between 0.58 and 1.01
below on all eight. Profile 3, the largest at 46%, runs along the course average
from -0.19 to +0.29.

That is worth stating plainly, because it is the kind of result that is easy to
over-read. These data contain no enrolment type that reads the forum heavily
while skipping lectures, or that browses widely in short bursts. Engagement on
this platform moves as a block: a student who is above their classmates on
session count is above them on duration, browsing, forum reading and lecture
viewing too. Three profiles is the right number here for describing *how much*,
not *in what way*.

The table reports each profile's variances, but this model also estimates the
covariances between indicators inside each profile, which is what section 6 is
about. Those are the numbers the default model fixes at zero, and on these data
they are large enough that fixing them at zero is not a simplification but a
misstatement.

## 4. The student classes: what a whole student looks like

Everything so far could have come from a single-level model. The second level is
where the enrolments belonging to one student are tied together, and a student
class is defined by nothing except the mix of profiles its students produce.

```{r probabilities-table}
as.data.frame(fit, what = "profile_probabilities")
```

```{r probabilities-plot}
plot(fit, what = "probabilities")
```

The three classes are uneven — 39%, 23% and 39%, or 41, 24 and 41 students —
and each is dominated by one profile. Class 1's students are typical 79% of the
time, class 2's take high-engagement courses 73% of the time, and class 3's take
low-engagement courses 71% of the time. The largest class is the unremarkable
one, and the high-engagement class is the smallest.

The interesting numbers are the small ones: **0.021 and 0.070**. A student in
the high-engagement class has a 2.1% chance of producing a low-engagement
enrolment, and a student in the low-engagement class has a 7.0% chance of
producing a high-engagement one. The two extremes are, for practical purposes,
mutually exclusive over a student's career.

Consistency therefore runs in both directions, and that is the finding. The
high-engagement students are not students who happen to have had a good term;
they almost never have a bad one. The low-engagement students are not students
waiting for the right course; across up to fifteen courses, the right course
does not arrive. What both groups do instead, roughly a quarter of the time, is
fall back to the typical profile — 0.247 for the high class and 0.220 for the
low one, which is to say the middle is where everybody's variation happens.

```{r group-posteriors}
head(as.data.frame(fit, what = "group_posteriors"))
```

One row per student, with the modal class, the student's own contribution to the
likelihood, and the posterior probabilities behind the assignment. The posteriors
are the honest object here: the modal class is a summary of them, not a fact
about the student.

## 5. The sequences themselves

A student class is a probability mix, and a mix is an average over careers that
may look nothing like each other. Class 3 says its students take a
low-engagement course 71% of the time; it does not say whether that means a
steady run of low-engagement terms or a good start followed by a collapse.

The model never looks at course order, but the fit can carry it. Passing
`time =` stores the column, which is what lets the assignments be read back in
sequence without handing the data over a second time.

```{r sequences}
head(sequences(fit), 12)
```

One row per enrolment, in course order, carrying both the enrolment's profile
and its student's class — 1,393 rows in all, the same enrolments the model was
fitted to, now arranged so that a student's career reads left to right. The
first student shown took a single course; the second took twelve.

```{r sequence-index, fig.width = 10, fig.height = 7}
plot(fit, what = "sequences")
```

Each row is one student, each column one course in their sequence, coloured by
the assigned profile, with students blocked by their class. This is section 4's
finding made visible: the first block is overwhelmingly one colour, the second
overwhelmingly another, and the two extreme colours almost never appear in each
other's block. The mutual exclusivity of the extremes is the dominant visual
fact here, not a subtle statistical effect.

The gaps are what the probability table could not show. They mark positions a
student never reached.

```{r attrition}
sequence_summary(fit)
```

Class 3 — the low-engagement class of section 4, the one whose students take a
low-engagement course 71% of the time — averaged 11.3 courses against 14.2 and
14.3 for the other two, and 25 of its 41 students reached the fifteenth against
37 of 41 and 22 of 24. That is a fact about who leaves, not about who engages,
and the prevalence matrix had no way to express it.

Read it carefully, because it is partly circular. A student who took one course
and was disengaged in it has nowhere to be assigned except the mostly-low
class, so short sequences push themselves into it. What is not circular is that
the students who left were disengaged rather than merely brief: had leaving
early been unrelated to engagement, those who did not finish would have spread
across the three classes instead of concentrating in one.


## 6. Testing the assumption the model makes

Everything above rests on an assumption that has not yet been examined. A
profile model says that, once you know which profile an enrolment belongs to,
its indicators are independent of one another: whatever made a student browse
heavily is the profile, not some separate thing that also made them log many
sessions. The default `covariance_model = "diagonal"` writes that down as a
constraint, fixing every within-profile covariance at zero.

Nothing in the fit reports whether that constraint is reasonable. A model can be
badly wrong in exactly this way and still converge, replicate across every
start, and return a respectable entropy. `bivariate_residuals()` is the check.

```{r residuals}
head(bivariate_residuals(fit, engagement), 5)
```

One row per pair of indicators per profile, worst first. `observed` is the
residual correlation left inside the profile, `expected` is what the fitted
covariance model implies for it, and `residual` is the gap. Here the largest gap
is a few parts in a hundred thousand, because this model estimates those
correlations rather than assuming them away.

The comparison that matters is with the model that does assume them away.

```{r residuals-diagonal}
independent <- multilpa(engagement, indicators = indicators, group = "User",
                        n_profiles = 3, n_group_classes = 3,
                        n_starts = 20, seed = 2026)
ignored <- bivariate_residuals(independent, engagement)
head(ignored, 5)
```

```{r residuals-count}
data.frame(pairs = nrow(ignored),
           above_half = nrow(subset(ignored, abs(residual) > 0.5)),
           significant_holm = nrow(subset(ignored,
                                          p.adjust(p_value, "holm") < 0.05)))
```

Under the default, `active_days_z` and `entropy_z` remain correlated at 0.87
*within a single profile*, against a model asserting that the correlation is
exactly zero. Of the 84 pairs, 19 exceed 0.5 and 60 stay significant after a
Holm correction. That is not a marginal violation; it is the assumption failing
across most of the instrument. (Which profile carries the worst pair is not
worth naming: profile labels are arbitrary and move between fits.)

This is worth dwelling on, because of how invisible it was. That diagonal fit
converged, replicated in 50 starts out of 50, and reported a relative entropy of
0.91 — every signal the package offers said it was sound. Only a diagnostic
aimed at the assumption itself could show otherwise.

What it cost is visible in the criteria: the same three profiles and three
classes, with the covariances estimated instead of fixed, gain 3,183
log-likelihood points and drop the group-count BIC from 22428 to 16454. The
profiles keep their shape, and the substance of sections 4 and 5 survives — the
extremes stay nearly disjoint and the low-engagement class still leaves early —
but the numbers move, and a model that misstates its own measurement structure
is not the one to report them from.

## 7. How much to trust the assignments

A profile solution can be well estimated and still be too blurred to act on. Two
diagnostics separate those cases, and they should be read before any assignment
is used for anything.

```{r entropy}
entropy_table(fit)
```

Relative entropy is 0.839 for enrolments and 0.914 for students. Both are high
enough that the modal assignments mean something. The ordering is worth noticing:
students separate more sharply than their individual enrolments do, because a
student is judged on fifteen observations while an enrolment is judged on one.
Under the independence assumption the ordering was the other way round, which is
one more thing that assumption was getting wrong.

```{r classification}
classification_table(fit, level = "both")
```

The average posterior probability is between 0.922 and 0.947 for the enrolment
profiles and between 0.961 and 0.979 for the student classes — every class is
assigned with high confidence on average. The odds of correct classification
range from 13.6 to 49.3 for enrolments and 38.6 to 92.0 for students, all far
above the value of 1 that a useless classifier would give.

Note also that `n_modal` and `estimated_n` agree closely — 365 against 373 for
profile 1, 370 against 373 for profile 2. Where a model separates poorly these
two diverge, because the modal count claims a certainty the posteriors do not
support. Here they do not diverge.

## 8. What this model cannot tell you

Three limits are worth naming, because each is invisible in the output above.

**The labels are arbitrary.** "Profile 1" is high-engagement in this fit with
this seed. Refitting can permute the numbering without changing the model at
all. Any comparison between two fits has to align the labels first.

**Sequence was not used.** The data carry each student's course order, and the
model ignores it: an enrolment's profile is estimated the same way whether it
was that student's first course or their fifteenth. So the classes describe a
student's overall mix, not a trajectory, and "students who start strong and
fade" could in principle land in the same class as "students who are
consistently middling". Section 5 lays the order out after the fact so it can
be read by eye, but reading it is not the same as modelling it.

**Standardising within course removes a real comparison.** Every value is
relative to classmates, so a student in a uniformly disengaged course can reach
profile 1 with activity that would be unremarkable elsewhere. That was the right
choice for asking who stands out among their peers. It is the wrong basis for
asking who did the most work.

### When to use which

Use `n_group_classes = 1` when the grouping exists only as a nuisance — the
model then reduces to an ordinary latent profile analysis that still respects
the clustering. Raise it above one when the question is about the units rather
than the observations, as here. Use `fit_random_intercept()` instead when
group differences are better described as a continuum than as types, and compare
the two on the information criteria rather than assuming.

## References

Vermunt, J. K. (2003). Multilevel latent class models. *Sociological
Methodology*, 33, 213–239.
