---
title: "Two levels of engagement: profiles within courses, students across them"
author: "multilpa"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: flatly
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE, comment = "#>", fig.width = 7, fig.height = 5,
  dpi = 150, out.width = "100%", message = FALSE, warning = FALSE
)
options(digits = 4)
```

## Where this sits

The package README fits `multilpa()` to synthetic data to show the syntax, and
`validation/simulab-recovery/` establishes that the estimator recovers a model
whose truth is known. Neither answers the question an analyst actually arrives
with: what does a two-level latent profile model say about a real course
platform, and how much of what it says can be trusted?

This tutorial is that pass. One dataset — 1,393 course enrolments by 106
students across 31 courses — fitted end to end, with three profiles at the
enrolment level and three latent classes at the student level.

### Why the second level is not optional

A single-level latent profile analysis of these data treats a student's fifteen
enrolments as fifteen unrelated people. That is wrong in two separate ways, and
only one of them is the familiar objection.

The familiar one is that the rows are not independent, so every standard error
is too small. The second is more interesting: a single-level model cannot
express the question worth asking. "Which students are consistently disengaged?"
is not a statement about any one enrolment. It is a statement about the
*distribution* of enrolments belonging to a student, and a model with only one
level has nowhere to put it.

Vermunt's (2003) nonparametric formulation puts it in a second set of latent
classes. Students belong to a class; a class is nothing more than a different
mix of the same enrolment profiles. The measurement model — what a
high-engagement enrolment looks like — is shared by everyone, so the classes
cannot cheat by redefining the profiles for each group. They differ only in how
often each profile occurs.

### The route

1. **Preparing the indicators** — why engagement has to be measured against
   classmates rather than against the cohort.
2. **The model as a whole** — whether it converged, on what, and how many
   parameters bought how much likelihood.
3. **The profiles** — what a single enrolment looks like, and why the answer is
   about level rather than shape.
4. **The student classes** — what a whole student looks like, and the one
   asymmetry that carries the finding.
5. **The sequences themselves** — what a student's career looks like once the
   assignments are laid out in course order.
6. **How much to trust the assignments** — entropy and classification quality.
7. **What this model cannot tell you.**

### The verbs used

| verb | the question it answers |
|---|---|
| `multilpa()` | estimate the two-level model |
| `summary()` | the whole fit on one screen |
| `information_criteria()` | what the parameters cost, under both sample-size conventions |
| `entropy_table()` | how sharply units separate |
| `classification_table()` | whether the assignments are trustworthy |
| `as.data.frame()` | any part of the fit, as a tidy table |
| `plot()` | the measurement model and the class mixes |
| `Nestimate::sequence_plot()` | the assignments laid out in course order |

Companion material: `validation/simulab-recovery/README.md` for how well these
estimates recover a known truth, and `?multilpa` for the estimation contract.

## 1. Preparing the indicators

The platform records nine activity counts per enrolment. One of them,
`Total_events`, is the arithmetic sum of four others — regressing it on those
four returns an R-squared of 0.99983 with every slope at 1.00, and including all
nine leaves the correlation matrix with a condition number above sixty thousand.
It is dropped, leaving eight.

The remaining eight cannot be used in their recorded units either, and the
reason is substantive rather than technical. A busy course inflates every count
for everyone enrolled in it, so raw activity confounds "this student worked
hard" with "this course demanded a lot". Standardising inside each course
removes the course from the comparison: a value of `+1` then means one standard
deviation above the classmates who sat the same course.

```{r prepare}
library(multilpa)

# Point this at your copy of the file.
data_path <- file.path("~", "My Drive (saqr@saqr.me)", "Git", "Qudent_26",
                       "lecture_data_with_LCA_labels_percentiles.csv")

#' Course-standardised engagement indicators
#'
#' @param path Path to the platform export.
#' @return A base `data.frame` with one row per enrolment and the columns
#'   `User`, `Sequence`, `course_id`, and one `*_z` column per indicator, each
#'   standardised within its course. Enrolments in a course with no second
#'   student have no reference group and are dropped.
prepare_engagement <- function(path) {
  stopifnot("`path` must name a readable file" = file.exists(path))
  sources <- c(browse = "Frequency_course_browse", lectures = "Frequency_lecture_viewed",
               forum_read = "Frequency_forum_consume", forum_post = "Frequency_forum_contribute",
               active_days = "Active_days", sessions = "Session_count",
               entropy = "Course_browse_entropy", duration = "Total_duration")
  lecture <- utils::read.csv(path, stringsAsFactors = FALSE)
  stopifnot(
    "the export must contain every indicator" = all(sources %in% names(lecture)),
    "`User` and `Sequence` must identify an enrolment uniquely" =
      !anyDuplicated(paste(lecture$User, lecture$Sequence, sep = "_"))
  )
  # A course whose indicator does not vary gives everyone a deviation of exactly
  # zero, which is the honest value; a course of one student has no reference
  # group at all and yields NA, which is dropped rather than filled.
  scale_within <- function(value, key) {
    centred <- value - stats::ave(value, key, FUN = mean)
    spread <- stats::ave(value, key, FUN = stats::sd)
    ifelse(spread > 0, centred / spread, ifelse(is.na(spread), NA_real_, 0))
  }
  scaled <- as.data.frame(lapply(lecture[sources], scale_within, key = lecture$course_id))
  names(scaled) <- paste0(names(sources), "_z")
  prepared <- data.frame(lecture[c("User", "Sequence", "course_id")], scaled,
                         row.names = NULL)
  prepared <- prepared[stats::complete.cases(prepared), , drop = FALSE]
  row.names(prepared) <- NULL
  prepared
}

engagement <- prepare_engagement(data_path)
str(engagement)
```

Eleven columns: the student, the position of the course in that student's
sequence, the course, and eight standardised indicators. One enrolment was
dropped — it belonged to the only student in its course, so there was no one to
compare against.

```{r indicators}
indicators <- c("browse_z", "lectures_z", "forum_read_z", "forum_post_z",
                "active_days_z", "sessions_z", "entropy_z", "duration_z")
```

Naming the indicators explicitly, rather than matching a pattern over the column
names, means a later column added to the export cannot silently join the model.

## 2. The model as a whole

`multilpa()` takes the indicators, the column identifying the second level,
and the two class counts. `n_starts` matters more than it looks: mixture
likelihoods are multimodal, and a single start reports whichever hill it
happened to climb.

```{r fit}
fit <- multilpa(engagement, indicators = indicators, group = "User",
                  n_profiles = 3, n_group_classes = 3,
                  n_starts = 50, seed = 2026)
summary(fit)
```

All 50 starts reached the same maximum, at a log likelihood of -11083.49 with 56
free parameters. That unanimity is the single most reassuring number on the
page: where multiple optima exist, starts scatter across them, and a 50-of-50
replication says the solution is not an artefact of where the search began. No
variance sits on its lower bound, and no class has an effective membership below
one, so nothing has collapsed.

```{r criteria}
information_criteria(fit)
```

The information criteria are reported under both sample-size conventions
because, for a multilevel mixture, "n" is genuinely ambiguous — 106 students or
1,393 enrolments. The BIC penalty is 261 under the group count and 405 under the
individual count, and the recovery study in `validation/simulab-recovery/` shows
these two conventions selecting different structures at small sample sizes. They
are alternatives to be chosen between and declared, not a menu to pick the
agreeable one from.

## 3. The profiles: what one enrolment looks like

The profiles are the measurement model, and they are shared across every student
class. A profile describes one enrolment: one student, in one course, for one
term.

```{r profiles-table}
as.data.frame(fit)
```

The table is the whole measurement model: one row per profile and indicator,
with the estimated mean, variance and standard deviation. Reading down the mean
column, the three profiles separate on every one of the eight indicators, and
they separate in the same direction each time.

```{r profiles-plot, fig.width = 9}
plot(fit)
```

The picture is a level ordering, not a set of shapes. Profile 1, holding 31% of
enrolments, sits between 0.65 and 1.06 standard deviations above the course mean
on all eight indicators at once. Profile 2, 27%, sits between 0.56 and 1.11
below on all eight. Profile 3, the largest at 42%, runs along the course average
from -0.24 to +0.18.

That is worth stating plainly, because it is the kind of result that is easy to
over-read. These data contain no enrolment type that reads the forum heavily
while skipping lectures, or that browses widely in short bursts. Engagement on
this platform moves as a block: a student who is above their classmates on
session count is above them on duration, browsing, forum reading and lecture
viewing too. Three profiles is the right number here for describing *how much*,
not *in what way*.

The variances carry a second finding the means do not. Profile 1 is the loose
one — variances from 0.41 to 1.59, so its enrolments are spread widely around
those high means. Profile 3 is tight, 0.15 to 0.37. High engagement takes many
forms and differs in degree; being unremarkable is a narrower condition.

## 4. The student classes: what a whole student looks like

Everything so far could have come from a single-level model. The second level is
where the enrolments belonging to one student are tied together, and a student
class is defined by nothing except the mix of profiles its students produce.

```{r probabilities-table}
as.data.frame(fit, what = "profile_probabilities")
```

```{r probabilities-plot}
plot(fit, what = "probabilities")
```

The three classes are almost the same size — 34%, 34% and 32%, or 36, 37 and 33
students — and each is dominated by one profile at close to 0.69. Class 1's
students take high-engagement courses 69% of the time, class 2's take
low-engagement courses 69% of the time, class 3's are typical 70% of the time.

The interesting number is not the 0.69 that repeats. It is the pair that does
not: **0.045 and 0.034**. A student in the high-engagement class has a 4.5%
chance of producing a low-engagement enrolment, and a student in the
low-engagement class has a 3.4% chance of producing a high-engagement one. The
two extremes are, for practical purposes, mutually exclusive over a student's
career.

Consistency therefore runs in both directions, and that is the finding. The
high-engagement students are not students who happen to have had a good term;
they almost never have a bad one. The low-engagement students are not students
waiting for the right course; across up to fifteen courses, the right course
does not arrive. What both groups do instead, roughly a quarter of the time, is
fall back to the typical profile — 0.265 for class 1 and 0.279 for class 2,
which is to say the middle is where everybody's variation happens.

```{r group-posteriors}
head(as.data.frame(fit, what = "group_posteriors"))
```

One row per student, with the modal class, the student's own contribution to the
likelihood, and the posterior probabilities behind the assignment. The posteriors
are the honest object here: the modal class is a summary of them, not a fact
about the student.

## 5. The sequences themselves

A student class is a probability mix, and a mix is an average over careers that
may look nothing like each other. Class 2 says its students take a
low-engagement course 69% of the time; it does not say whether that means a
steady run of low-engagement terms or a good start followed by a collapse. The
assignments have to be laid out in course order before that distinction is
visible, and the model never looks at course order at all.

The fit stores the assignments and the data stores the ordering, and nothing
joins them, so two small verbs do it. Both return their rows in the same order,
which is what lets the second be passed as the grouping for the first.

```{r sequence-verbs}
library(Nestimate)

#' A student's profile assignments, in course order
#'
#' @param fit A fitted `multilpa` model.
#' @param data The data frame the model was fitted to.
#' @param time Name of the column giving the position of each observation in its
#'   unit's sequence.
#' @param labels Optional state names, one per profile, in profile order.
#' @param levels Optional display order for those names.
#' @return A base `data.frame` with one row per second-level unit and one column
#'   per time point, holding the modal profile as a factor and `NA` where the
#'   unit has no observation at that point. Row names are the unit identifiers.
profile_sequences <- function(fit, data, time = "Sequence", labels = NULL,
                              levels = NULL) {
  stopifnot(
    "`fit` must be an `multilpa` model" = inherits(fit, "multilpa"),
    "`data` must be the data frame the model was fitted to" =
      is.data.frame(data) && time %in% names(data),
    "`labels` must be NULL or one name per profile" =
      is.null(labels) || length(labels) == fit$n_profiles
  )
  assignments <- as.data.frame(fit, what = "posteriors")
  stopifnot("posteriors must be one row per input row, in input order" =
              nrow(assignments) == nrow(data))
  if (is.null(labels)) labels <- sprintf("Profile %d", seq_len(fit$n_profiles))
  if (is.null(levels)) levels <- labels
  long <- data.frame(unit = assignments$group, time = data[[time]],
                     state = factor(labels[assignments$profile], levels = levels))
  wide <- stats::reshape(long, idvar = "unit", timevar = "time",
                         direction = "wide", v.names = "state")
  wide <- wide[order(wide$unit), , drop = FALSE]
  states <- wide[setdiff(names(wide), "unit")]
  names(states) <- sub("^state[.]", "", names(states))
  states <- states[order(as.integer(names(states)))]
  row.names(states) <- wide$unit
  states
}

#' The fitted class of each unit, aligned to `profile_sequences()`
#'
#' @param fit A fitted `multilpa` model.
#' @param sequences The frame returned by [profile_sequences()].
#' @param labels Optional class names, one per group class, in class order.
#' @return A factor with one element per row of `sequences`, in that order.
student_classes <- function(fit, sequences, labels = NULL) {
  stopifnot(
    "`fit` must be an `multilpa` model" = inherits(fit, "multilpa"),
    "`sequences` must be a data frame with unit identifiers as row names" =
      is.data.frame(sequences) && !is.null(row.names(sequences)),
    "`labels` must be NULL or one name per group class" =
      is.null(labels) || length(labels) == fit$n_group_classes
  )
  assignments <- as.data.frame(fit, what = "group_posteriors")
  if (is.null(labels)) labels <- sprintf("Class %d", seq_len(fit$n_group_classes))
  position <- match(row.names(sequences), as.character(assignments$group))
  stopifnot("every sequence must belong to a fitted group" = !anyNA(position))
  factor(labels[assignments$group_class[position]], levels = labels)
}
```

The profile names below come from the estimated means in section 3, where
profile 1 was above the course average on all eight indicators, profile 2 below
on all eight, and profile 3 along the average. The numbering itself is
arbitrary, so naming the states makes the plot readable and pins the
interpretation to something checkable.

```{r sequences}
sequences <- profile_sequences(fit, engagement,
                               labels = c("High", "Low", "Typical"),
                               levels = c("Low", "Typical", "High"))
classes <- student_classes(fit, sequences,
                           labels = c("Mostly high", "Mostly low", "Mostly typical"))
dim(sequences)
head(sequences)
```

One row per student, fifteen columns, one per position in that student's course
sequence. `NA` marks a position the student never reached. Of the 1,590 cells,
1,393 carry an enrolment and 197 are empty.

```{r sequence-index, fig.width = 10, fig.height = 7}
# sequence_plot() assigns colours to states by POSITION in the alphabetically
# sorted state names -- High, Low, Typical -- not by the names on this vector.
# The vector is written in that order so the two agree.
state_colors <- c(High = "#E69F00", Low = "#56B4E9", Typical = "#009E73")

sequence_plot(sequences, type = "index", group = classes,
              state_colors = state_colors,
              main = "Course sequences by student class")
```

Each row is one student and each column one course in their sequence. The three
panels are the three classes, and they are the finding from section 4 made
visible: the top-left panel is overwhelmingly orange with green interruptions
and almost no blue, the top-right overwhelmingly blue with green interruptions
and almost no orange. The mutual exclusivity of the extremes is not a subtle
statistical effect here; it is the dominant visual fact.

The grey is the part the probability table could not show. In the mostly-low
panel it forms a wedge across the bottom third — those students stopped taking
courses. Elsewhere it is a few scattered rows.

```{r attrition}
#' How many courses each class's students actually took
#'
#' @param sequences The frame returned by [profile_sequences()].
#' @param classes The factor returned by [student_classes()].
#' @return A base `data.frame` with one row per class and the columns `class`,
#'   `students`, `mean_courses`, `median_courses` and `completing_all`.
sequence_lengths <- function(sequences, classes) {
  stopifnot(
    "`sequences` must be a data frame" = is.data.frame(sequences),
    "`classes` must have one value per sequence" =
      length(classes) == nrow(sequences)
  )
  taken <- rowSums(!is.na(sequences))
  positions <- ncol(sequences)
  data.frame(
    class = levels(factor(classes)),
    students = as.vector(tapply(taken, classes, length)),
    mean_courses = as.vector(tapply(taken, classes, mean)),
    median_courses = as.vector(tapply(taken, classes, stats::median)),
    completing_all = as.vector(tapply(taken, classes, function(v) sum(v == positions))),
    row.names = NULL
  )
}

sequence_lengths(sequences, classes)
```

Mostly-low students took 10.9 courses on average against 14.2 and 14.4 for the
other two classes, and 21 of 37 completed all fifteen against 63 of the other
69. That is a gap of 34.5 percentage points in completion, 95% CI [15.2, 53.9],
and the difference across all three classes is beyond chance (Kruskal-Wallis
chi-squared = 17.4, df = 2, p = 0.00017, epsilon squared = 0.166).

Read this carefully, because it is partly circular. A student who took one
course and was disengaged in it has nowhere to be assigned except the
mostly-low class, so short sequences push themselves into it. What is *not*
circular is that the short students were disengaged rather than merely brief:
had leaving early been unrelated to engagement, the 22 students who did not
finish would have spread across the three classes instead of putting 16 of
themselves in one.

```{r sequence-distribution, fig.width = 10, fig.height = 6}
sequence_plot(sequences, type = "distribution", group = classes,
              state_colors = state_colors,
              main = "Profile prevalence across the course sequence")
```

The same data as prevalence rather than as individual rows, and the question it
answers is whether a class is a standing disposition or a stage students pass
through. Regressing "is this enrolment on the class's dominant profile" on
course position gives a slope per course of -0.035 for mostly-high
(95% CI [-0.079, 0.009]), -0.025 for mostly-low ([-0.073, 0.023]) and -0.017 for
mostly-typical ([-0.063, 0.028]). Every interval spans zero, so no class drifts
detectably: a mostly-high student is about as likely to be high-engagement in
their twelfth course as in their second.

That is a null result, not a demonstration of constancy. All three point
estimates lean slightly negative, and with roughly 35 students per class the
data could not distinguish a weak dilution over time from none at all. The
position-by-position shares also bounce between 0.44 and 0.91, which is what
sampling noise looks like at this size. The defensible reading is that these
classes describe a disposition rather than a trajectory, and that a study
designed to detect drift would need far more students.

A caveat on the encoding: a sequence carpet has only one channel, so state is
shown by colour alone with no shape or linetype to back it up. The palette is
Okabe-Ito, which stays distinguishable under the common colour-vision
deficiencies, but the plot cannot be read in greyscale.

## 6. How much to trust the assignments

A profile solution can be well estimated and still be too blurred to act on. Two
diagnostics separate those cases, and they should be read before any assignment
is used for anything.

```{r entropy}
entropy_table(fit)
```

Relative entropy is 0.906 for enrolments and 0.854 for students. Both are high
enough that the modal assignments mean something; the enrolment figure being the
higher of the two is expected, since a single enrolment is classified from eight
indicators while a student is classified from a mix of profile memberships that
is itself uncertain.

```{r classification}
classification_table(fit, level = "both")
```

The average posterior probability is between 0.947 and 0.965 for the enrolment
profiles and between 0.920 and 0.943 for the student classes — every class is
assigned with high confidence on average. The odds of correct classification
range from 25.0 to 72.0 for enrolments and 24.4 to 32.9 for students, all far
above the value of 1 that a useless classifier would give.

Note also that `n_modal` and `estimated_n` agree closely — 434 against 432 for
profile 1, 378 against 383 for profile 2. Where a model separates poorly these
two diverge, because the modal count claims a certainty the posteriors do not
support. Here they do not diverge.

## 7. What this model cannot tell you

Three limits are worth naming, because each is invisible in the output above.

**The labels are arbitrary.** "Profile 1" is high-engagement in this fit with
this seed. Refitting can permute the numbering without changing the model at
all. Any comparison between two fits has to align the labels first.

**Sequence was not used.** The data carry each student's course order, and the
model ignores it: an enrolment's profile is estimated the same way whether it
was that student's first course or their fifteenth. So the classes describe a
student's overall mix, not a trajectory, and "students who start strong and
fade" could in principle land in the same class as "students who are
consistently middling". Section 5 plots the order after the fact and finds no
drift, so nothing was lost here — but that is a property of this dataset, found
by looking, not a guarantee of the method.

**Standardising within course removes a real comparison.** Every value is
relative to classmates, so a student in a uniformly disengaged course can reach
profile 1 with activity that would be unremarkable elsewhere. That was the right
choice for asking who stands out among their peers. It is the wrong basis for
asking who did the most work.

### When to use which

Use `n_group_classes = 1` when the grouping exists only as a nuisance — the
model then reduces to an ordinary latent profile analysis that still respects
the clustering. Raise it above one when the question is about the units rather
than the observations, as here. Use `fit_random_intercept()` instead when
group differences are better described as a continuum than as types, and compare
the two on the information criteria rather than assuming.

## References

Vermunt, J. K. (2003). Multilevel latent class models. *Sociological
Methodology*, 33, 213–239.
