---
title: "Case study: Two-level latent class analysis of leisure activities"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Case study: Two-level latent class analysis of leisure activities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", dpi = 96)
options(digits = 3, width = 100)
```

Two-level latent class analysis estimates latent classes of categorical
responses at the level of single observations and, above them, latent group
classes that differ in how probable each class is for their observations. In
the output the observation-level classes are called profiles. Each profile is
described by the probability of every answer to every item; the group classes
differ only in how often their observations fall in each profile.

The question here is whether the leisure activities that university students
report in daily life form a small number of recurring patterns, and whether
students differ in how often they are in each pattern.

## Data

`student_esm` holds experience-sampling data from a study of university
students (Neubauer & Schmiedek, 2024). The students answered six prompts a day
for fourteen days. At each prompt when they had not studied since the previous
one, they reported which of eight leisure activities they had done. The
dataset holds these prompts for 100 students: 2,582 prompts, between 15 and 51
per student. Prompts are the observations and students the groups, so a
student contributes many rows.

The eight activities are yes/no items, and they are named once so that each
model can refer to them.

```{r setup}
library(multilpa)
activities <- c("time_with_friends", "on_social_media", "tv_video_games",
                "listened_music", "sports", "walking", "reading",
                "part_time_job")
```

`summary()` counts the answers to each activity item and summarizes the other
variables.

```{r counts}
summary(student_esm)
```

Social media is the most frequent activity, reported at 745 of the 2,582
prompts, followed by TV or video games at 549. A part-time job is the rarest,
at 103 prompts. Every item is answered "no" at most prompts, so a profile is
described by how much more often than average its prompts say "yes".

## The model

For prompt $t$ of student $i$, let $C_{it}$ be the profile and $G_i$ the
student's class. The model estimates three sets of probabilities:

- the response probabilities $\theta_{kjc}=P(Y_{itj}=c\mid C_{it}=k)$ of each
  answer $c$ to item $j$ in profile $k$, the same in every student class;
- the profile probabilities $\pi_{k\mid m}=P(C_{it}=k\mid G_i=m)$ of each
  student class $m$, which sum to one across profiles;
- the class probabilities $\omega_m=P(G_i=m)$.

The items are assumed independent given the profile, the prompts of a student
independent given the student's class, and the students independent of one
another. The likelihood can have several local maxima, so each model is
started from several random values. A rare answer can drive a response
probability to its lower bound, and too many profiles or classes for the data
show up as starts that fail to reach the same maximum.

## How many profiles and classes

`enumerate_classes()` fits every combination of two to four profiles and one
to three student classes, each from five random starts, and reports the
information criteria and the diagnostics of each fit. It does not choose a
model.

```{r enumerate}
candidates <- enumerate_classes(student_esm, vars = activities,
                                categorical = activities, id = "student",
                                n_profiles = 2:4, n_group_classes = 1:3,
                                n_starts = 5, seed = 1)
as.data.frame(candidates)
get_results(candidates, "criteria")
```

The criteria disagree, and the diagnostics explain why. BIC with the number of
students keeps falling up to the largest model in the grid, four profiles and
three classes. Most of the larger models reach their best solution from only
one of the five starts, and two of the four-profile models do not converge,
which the two warnings above report, so their low BIC rests on unstable
solutions. The criteria that also penalize
classification uncertainty (ICL, AWE and CLC counted over prompts) choose two
profiles and two classes, the model whose best solution recurs in all five
starts.

The ICL plot shows the same choice across the grid.

```{r icl-plot}
plot(candidates, criterion = "icl_individual")
```

The plot marks the minimum at two profiles and two classes, and the analysis
continues with that model.

## The chosen model

`multilpa()` fits it with `categorical` naming the items, ten random starts,
and a tight convergence tolerance, which the standard errors below need.

```{r fit}
lca <- multilpa(student_esm, vars = activities, categorical = activities,
                id = "student", n_profiles = 2, n_group_classes = 2,
                n_starts = 10, tol = 1e-10, seed = 1)
get_results(lca, "model")
```

The fit converges, and six of the ten starts reach the best log likelihood of
-7740.

## Two activity profiles

The `"responses"` table gives the probability of each answer in each profile,
with a "no" and a "yes" row for every item. The "yes" rows carry the
information, since the two probabilities of an item sum to one.

```{r responses}
get_results(lca, "responses")
```

The two profiles differ mainly in media use. At prompts in profile 2, students
used social media with probability 0.765, watched TV or played video games with
probability 0.396, and listened to music with probability 0.364. In profile 1
the same probabilities are 0.119, 0.147 and 0.075. Profile 1 is slightly more
likely to include time with friends (0.157 against 0.106), walking (0.138
against 0.096) and a part-time job (0.051 against 0.010), and both profiles
report sports at 0.07. Profile 2 is therefore a media-heavy pattern, and
profile 1 a pattern of little media use with the remaining activities at
their usual low rates.

The response plot draws the same probabilities, one line per profile across
the items.

```{r responses-plot}
plot(lca, what = "responses")
```

The lines separate on the three media items and nearly coincide on sports, so
the media items carry almost all of the difference between the profiles.

## Two student classes

The `"counts"` table gives the size of each profile and class, and the
`"profile_probabilities"` table the profile mix of each student class.

```{r mix}
get_results(lca, "counts")
get_results(lca, "profile_probabilities")
```

The low-media profile holds 73.7% of prompts and the media-heavy profile
26.3%. The student classes differ sharply in their mix. In class 2, which holds
58.4% of students, 0.974 of prompts fall in the low-media profile. In class 1,
with 41.6% of students, the media-heavy profile is the more common, at 0.565.
Media-heavy leisure is therefore concentrated in a minority of students, and
for them it is the usual pattern, while most students almost never report it.

The probability plot compares the two mixes.

```{r probabilities-plot}
plot(lca, what = "probabilities")
```

Class 2 runs close to one on the low-media profile and close to zero on the
media-heavy one; class 1 divides its prompts more evenly, with the larger
share on the media-heavy profile.

## Classification certainty

Profiles and classes are estimated as posterior probabilities. Relative
entropy summarizes how concentrated those probabilities are, from 0
for no separation to 1 for certain assignment, and the `"classification"`
table reports the average posterior probability of each assigned class.

```{r classification}
get_results(lca, "entropy")
get_results(lca, "classification")
```

Relative entropy is 0.668 for prompts and 0.877 for students. A single prompt
answers eight yes/no questions, which leaves its profile uncertain; a student
contributes many prompts, which pins down the class. The average posterior
probability is 0.926 for prompts assigned to the low-media profile and 0.859
for the media-heavy one, and 0.980 and 0.960 for the two student classes.
Conclusions about students are therefore well supported, while an analysis of
individual prompts should carry the posterior probabilities forward.

## Local independence

The model assumes the items are independent within a profile. The
`"residuals"` table measures the association that each pair of items keeps
within a profile, with a chi-square test for each pair; the ten largest are
shown.

```{r residuals}
get_results(lca, "residuals") |> head(10)
```

The largest residual is 0.171, between sports and walking in the media-heavy
profile, followed by friends and walking (0.140) and music and walking (0.133)
in the same profile. Physical and social activities occur together more often
than two profiles allow, which is a limit of a two-profile model and one reason
BIC rewards larger models: extra profiles can absorb this association.

## Uncertainty in the response probabilities

`parameter_inference()` gives standard errors and 95% Wald intervals for every
parameter. The intervals of the "yes" probabilities, in the rows whose `term`
ends in `:yes`, show how firmly the profiles are distinguished.

```{r inference}
parameter_inference(lca)
```

The profiles are clearly separated on the media items: the interval for
social media is 0.696 to 0.834 in the media-heavy profile and 0.082 to 0.156 in
the low-media profile. On sports the two intervals overlap almost entirely.
The part-time-job probability in the media-heavy profile has an interval
starting at 0.0001, close to the bound, because very few of its prompts report
a job.

## Adding affect

A mixed model adds three affect ratings (happy, relaxed, exhausted, each from
1 to 7) as continuous indicators. `worried` is left out because about half of
its ratings are 1, which can push a profile's variance to its lower bound.

```{r mixed}
mixed <- multilpa(student_esm,
                  vars = c("happy", "relaxed", "exhausted", activities),
                  categorical = activities, id = "student", n_profiles = 2,
                  n_group_classes = 2, n_starts = 10, seed = 1)
get_results(mixed)
get_results(mixed, "responses")
get_results(mixed, "profile_probabilities")
```

With affect included, the profiles become profiles of mood. Profile 1 has mean
ratings of 5.85 for happy, 5.59 for relaxed and 2.55 for exhausted; profile 2
has 3.58, 3.33 and 4.33. The activities now differ less and in a different
way: time with friends is reported at 0.188 of profile-1 prompts and 0.084 of
profile-2 prompts, social media at 0.269 and 0.315. Good mood goes with social
time more than with media use. The student classes remain distinct: 0.833 of
the prompts of class 1 fall in the good-mood profile, and 0.873 of those of
class 2 in the low-mood profile. The profiles a model finds are set by the
indicators it is given, so the choice of indicators is part of the question.

## Interpretation

Students' leisure at non-study prompts falls into two patterns, one defined by
social media, TV and music and one with little media use. The distinction that
matters most is between students: a majority almost never reports the
media-heavy pattern, and a minority reports it at more than half of their
prompts. That student-level result is well supported by the classification
diagnostics, while the assignment of any single prompt is uncertain.

## Limitations

The data contain only prompts at which the student had not studied, so the
patterns describe leisure time, not the whole day. The two-profile model leaves
association between physical and social activities within its profiles, and
larger models fit better by BIC but are not stable across starts on these
data. Prompts are treated as independent given the student's class, so the
model ignores the order of prompts within a day. The associations are
observational and say nothing about causes.

## Reference

Neubauer, A. B., & Schmiedek, F. (2024). Approaching academic adjustment on
multiple time scales. *Zeitschrift für Erziehungswissenschaft, 27*(1),
147-168. <https://doi.org/10.1007/s11618-023-01182-8>
