---
title: "Case study: Covariates and staged estimation in multilevel LPA"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Case study: Covariates and staged estimation in multilevel LPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4, width = 100)
```

Three-step covariate analysis is a method for examining associations between latent profile membership and variables that are external to the measurement model. An important feature of this approach is that it adjusts for uncertainty in profile classification. The adjustment reduces the bias that arises when individuals are assigned to latent classes with imperfect certainty.

In the course-engagement example, the primary research question is whether students' previous grades are associated with their engagement patterns in a subsequent course. This relationship can be investigated in several ways. One approach is to regress latent profile membership on previous grades in order to assess whether prior academic performance predicts the likelihood of belonging to a particular engagement profile. A second approach is to compare mean previous grades across the identified profiles. A third approach is to estimate the model jointly by including previous grades directly in the latent profile model.

These approaches differ in the extent to which the external variable is allowed to influence the latent profile solution. In the three-step approaches, the measurement model is estimated first, and the resulting profile definitions are held fixed while associations with previous grades are examined. This preserves the original interpretation of the engagement profiles. In one-step estimation, previous grade enters the latent profile model and can change the profiles. The final section turns to staged estimation, which holds the profile definitions fixed while the student classes are estimated.

```{r setup}
library(latents)
```

## Data and engagement profiles
The bundled synthetic `course_engagement` dataset contains enrolment records nested within students. Engagement is represented by five activity indicators: page views (`browse`), lecture videos viewed (`lectures`), forum posts read (`forum_read`), forum posts written (`forum_post`), and active days (`attendance`). Each indicator is simulated on the `log1p` scale of an activity count and standardized within course to make activity levels comparable across courses.

The variable `previous_grade` represents the standardized grade obtained in the preceding course. For a student's first observed course, an entry grade is used instead. Because this grade is measured before engagement in the subsequent course, it can be treated as a predictor of later engagement patterns. The [introductory vignette](case-engagement-profiles.html) provides additional explanation of the longitudinal structure of the data and the distinction between the student and enrolment levels.

The indicator vector `vars` specifies the activity measures used to estimate the latent engagement profiles. In keeping `previous_grade` separate from these indicators, the engagement profiles can first be established using activity data alone. The relationship between previous academic performance and profile membership can then be examined in a subsequent stage without allowing grades to influence the initial definition of the profiles.

```{r variables}
vars <- c("browse", "lectures", "forum_read", "forum_post", "attendance")
```

The `multilpa()` function estimates the means and variances that define each latent profile, together with the probability of observing each profile within each student class. Let this conditional probability be written as

$$
\pi_{k\mid m}=P(C_{it}=k\mid G_i=m),
$$

where $C_{it}$ denotes the latent engagement profile of student $i$ in course $t$, and $G_i$ denotes the student's latent class. For each student class, the profile probabilities sum to one across all profiles. These probabilities therefore describe how the distribution of course-level engagement profiles differs across student-level classes.

The model considered here includes three engagement profiles and three student classes. Within each profile, the five engagement indicators are modeled as independent Gaussian variables. The profile definitions are held common across student classes, meaning that the same profile-specific means and variances apply to all students. Student classes differ instead in the probabilities with which the engagement profiles occur.

Model estimation relies on several assumptions. Students are assumed to be independent of one another, while enrolments are assumed to be conditionally independent given student class. The model also assumes conditional independence among the engagement indicators within each profile. Violations of this latter assumption can affect the profiles that are recovered from the data. In addition, because latent profile models use likelihood-based estimation, the optimization procedure may converge to different local maxima depending on its starting values.

The [evaluation vignette](case-engagement-evaluation.html) discusses indicator dependence, likelihood maxima, and related model-evaluation issues in greater detail. In the present tutorial, the same candidate model is retained throughout so that the different covariate methods can be compared on a common basis. Ten random starts are used to reduce sensitivity to local solutions, and a fixed random seed ensures that the numerical results are reproducible.

```{r fit}
fit <- multilpa(course_engagement, vars, id = "student", time = "sequence",
                n_profiles = 3, n_group_classes = 3, n_starts = 10,
                max_iter = 2000, tol = 1e-10, seed = 1)
fit
```
The three profiles can be characterized as representing higher, intermediate, and lower levels of course activity, respectively. This ordering is particularly evident in the estimated means for `attendance`, which are 1.133 for Profile 1, 0.256 for Profile 2, and -0.940 for Profile 3. On this basis, Profile 3 is treated as the lower-activity reference category when interpreting the profile-membership regressions presented below.

These labels describe the current fitted solution; the profile numbers themselves carry no fixed meaning. Because latent profile labels are determined by the estimated parameters in each model, the profile-specific means should be examined again whenever a new model is fitted to confirm that the same substantive interpretation still applies.

## Accounting for classification error

The three-step approach accounts for uncertainty in profile assignment by using a classification-error matrix. This matrix describes the probability that an individual is assigned to a particular profile given their underlying latent profile. Let

$$
D_{ks}=P(\widehat{C}=s\mid C=k),
$$

where $C$ denotes the underlying latent profile and $\widehat{C}$ denotes the assigned profile, defined as the profile with the largest posterior probability. Each row of $D$ sums to one because it represents the distribution of possible assignments for a given underlying profile. The diagonal elements therefore give the estimated probabilities of being assigned to the correct profile, whereas the off-diagonal elements represent misclassification between profiles. In `multilpa`, this matrix can be extracted with `get_results()` using `"classification_errors"`.

The correction relies on the assumption that, conditional on the underlying latent profile, classification errors are independent of the auxiliary variable being studied. In other words, once true profile membership is taken into account, the probability of being misclassified is assumed not to depend systematically on the covariate or distal outcome.

The classification-error matrix is estimated from the initial measurement model and is subsequently treated as fixed during the covariate analysis. As a result, confidence intervals and standard errors from the later stages do not incorporate uncertainty associated with estimating this matrix. This distinction is important when interpreting inferential results, particularly when profile separation is limited. Poorly separated profiles produce greater classification uncertainty and can make the error correction less stable.

```{r classification-errors}
get_results(fit, "classification_errors", level = "individuals")
```

The estimated probabilities of correct profile assignment are 0.842 for Profile 1, 0.854 for Profile 2, and 0.980 for Profile 3. Classification is therefore most precise for Profile 3, whereas greater uncertainty remains between Profiles 1 and 2.

The main source of ambiguity is the overlap between these first two profiles. The estimated probability of assigning an observation from Profile 1 to Profile 2 is 0.158, while the probability of assigning an observation from Profile 2 to Profile 1 is 0.120. These off-diagonal probabilities indicate that a non-negligible proportion of observations may be classified into the neighboring profile rather than their underlying profile.

Treating the assigned profiles as if they were observed without error would ignore this uncertainty. The three-step correction instead carries the estimated misclassification structure forward into the subsequent covariate analysis, so associations with external variables are estimated with classification error taken into account.

## Previous grade as a membership predictor

The `r3step()` function estimates the association between an external predictor and latent profile membership while keeping the profile definitions and classification-error matrix fixed. In this example, `previous_grade` is used to predict membership in the engagement profiles identified in the measurement model. Because Profile 3 represents the lower-activity profile, it is used as the reference category.

Let $x$ denote standardized previous grade. For Profiles 1 and 2, the multinomial regression can be written as

$$
\log\frac{P(C=k\mid x)}{P(C=3\mid x)}
=
\alpha_k+\beta_k x,
\qquad k=1,2.
$$

The intercept $\alpha_k$ represents the log odds of belonging to Profile $k$, rather than the lower-activity reference profile, when $x=0$. Because `previous_grade` is standardized, $x=0$ corresponds to the average previous grade. The slope $\beta_k$ represents the change in these log odds associated with a one-standard-deviation increase in previous grade. A positive slope therefore indicates that higher previous grades are associated with greater relative odds of belonging to Profile $k$ rather than the lower-activity profile.

The model assumes that previous grade has a linear effect on the log-odds scale. For analyses conducted at the enrolment level, `vcov_type = "robust"` can be used to account for the clustering of multiple enrolments within the same student. Under this option, score contributions are clustered by student when estimating the covariance matrix. Reliable estimation therefore requires a sufficient number of independent students.

The `r3step()` regression describes the marginal association between previous grade and profile membership. This differs from the one-step model considered later, in which profile membership is modeled conditional on student class. The two approaches therefore answer related but distinct questions about how previous academic performance is associated with subsequent engagement.

The reported 95% confidence intervals are based on the normal approximation. When several slope coefficients are tested simultaneously, the reported adjusted $p$-values use the Benjamini–Hochberg procedure to control the false discovery rate across those tests. The confidence intervals remain pointwise intervals and are not adjusted for multiple comparisons.

```{r membership-predictor}
r3step(fit, data = course_engagement, covariates = "previous_grade", vcov_type = "robust")
```

In the fitted model, a one-standard-deviation increase in previous grade is associated with a 0.677 increase in the log odds of belonging to the higher-activity profile rather than the lower-activity profile. The corresponding 95% confidence interval ranges from 0.508 to 0.846. For the intermediate-activity profile, the estimated increase in log odds is 0.475, with a 95% confidence interval from 0.340 to 0.611.

Both coefficients are positive: higher previous grades are associated with greater relative odds of membership in the higher- or intermediate-activity profiles compared with the lower-activity profile. The results therefore show a consistent association between stronger previous academic performance and more active engagement in the subsequent course.

These findings are observational associations. Although `previous_grade` is measured before engagement in the subsequent course, temporal ordering alone does not establish that higher grades cause students to adopt a more active engagement pattern.

## Comparing previous grades across profiles

The `three_step()` function estimates profile-specific means of a numeric auxiliary variable and tests differences between those means while accounting for uncertainty in profile classification. In this example, `previous_grade` is treated as the auxiliary variable, so prior academic performance can be compared across the engagement profiles identified in the measurement model.

By default, `three_step()` uses the Bolck–Croon–Hagenaars (BCH) method. This approach constructs signed weights from the inverse of the classification-error matrix and uses those weights to correct for imperfect profile assignment. Let $x_u$ denote the previous grade for enrolment $u$, and let $w_{uk}$ denote its BCH weight for Profile $k$. The estimated profile-specific mean is

$$
\widehat{\mu}_k
=
\frac{\sum_u w_{uk}x_u}
{\sum_u w_{uk}}.
$$

Differences between profiles are then expressed as pairwise contrasts. For Profiles $k$ and $l$, the contrast is

$$
\widehat{\mu}_k-\widehat{\mu}_l.
$$

The BCH correction requires the classification-error matrix to be invertible and the resulting weighted profile sizes to be sufficiently well behaved for estimation. The default standard errors account for the clustering of multiple enrolments within students, while the previously estimated classification-error matrix is treated as fixed. Consequently, uncertainty from estimation of that matrix is not propagated into the reported standard errors.

Although `three_step()` can also be used with distal outcomes measured after profile membership, `previous_grade` occurs before the engagement measures in this example. The estimated contrasts should therefore be interpreted as differences in prior academic performance across subsequently observed engagement profiles rather than as effects of profile membership on grades.

The reported 95% confidence intervals are pointwise normal intervals. Adjusted $p$-values use the Benjamini–Hochberg procedure across the set of pairwise profile contrasts.

```{r grade-contrasts}
three_step(
  fit,
  data = course_engagement,
  outcome = "previous_grade",
  contrast = "pairs"
)
```

The BCH-adjusted estimates show that previous grades in the lower-activity profile are 0.629 standard deviations below those in the higher-activity profile and 0.446 standard deviations below those in the intermediate-activity profile. These contrasts indicate that students who later exhibit lower engagement tend to have had lower grades in the preceding course.

The difference between the intermediate- and higher-activity profiles is smaller. The estimated Profile 2 versus Profile 1 contrast is $-0.183$, with a 95% confidence interval from $-0.368$ to $0.002$. Because this interval extends slightly across zero, the difference in previous grades between the two more active profiles is less precisely estimated than their respective differences from the lower-activity profile.

### Comparison with modal assignment

A useful comparison is provided by modal analysis, in which each enrolment is assigned with certainty to the profile having the largest posterior probability. The same profile means and pairwise contrasts are then calculated using these hard assignments rather than classification-error-adjusted weights.

```{r modal-contrasts}
three_step(
  fit,
  data = course_engagement,
  outcome = "previous_grade",
  contrast = "pairs",
  method = "modal"
)
```

Modal analysis is simple to interpret, but it treats estimated profile assignments as if they were observed without error. Its standard errors therefore do not incorporate uncertainty associated with imperfect classification.

The effect of this simplification is particularly visible for Profiles 1 and 2, which showed the greatest classification overlap. Under modal assignment, the estimated Profile 2 versus Profile 1 difference is $-0.143$, compared with $-0.183$ under the BCH correction. The corresponding standard error is 0.067 under modal assignment and 0.094 under BCH.

Thus, accounting for classification error affects both the estimated contrast and its estimated uncertainty. The difference is most noticeable for the two profiles that are hardest to distinguish. Classification adjustment matters most when external variables are compared across partially overlapping latent profiles.

## Estimating profiles and grade associations together

One-step estimation differs from the staged approaches considered above because the latent profile measurement model and the regression of profile membership on the covariate are estimated simultaneously. In this example, specifying `profile_covariates = "previous_grade"` in `multilpa()` allows information about previous grades to contribute directly to estimation of the engagement profiles as well as to the association between grades and profile membership.

Let $x$ denote standardized previous grade and $G=m$ denote membership in student class $m$. With Profile 3 as the reference category, the profile-membership model is

$$
\log\frac{P(C=k\mid G=m,x)}
{P(C=3\mid G=m,x)}
=
\alpha_{km}+\beta_k x,
\qquad k=1,2.
$$

The intercepts $\alpha_{km}$ are allowed to vary across student classes. They therefore describe differences in the baseline profile probabilities between student classes when previous grade is at its standardized mean, $x=0$. In contrast, the grade slopes $\beta_k$ are shared across student classes. Each slope represents the change in the log odds of belonging to Profile $k$, rather than Profile 3, associated with a one-standard-deviation increase in previous grade, conditional on student class.

The one-step model retains the same Gaussian measurement assumptions used in the original latent profile model, including conditional independence of the engagement indicators within profiles. The important difference is that previous grade is now part of the model during estimation rather than being introduced only after the profiles have been established.

As a consequence, the profiles obtained from the one-step model need not be identical to those from the original measurement model. Allowing previous grade to contribute to estimation can alter profile-specific means, variances, or membership probabilities. The estimated regression coefficients also have a different interpretation from the three-step coefficients because they describe the association between previous grade and profile membership conditional on student class, whereas the earlier `r3step()` analysis describes marginal profile membership.

For these reasons, the magnitude of the one-step coefficients should not be expected to match the corresponding three-step estimates exactly. Before interpreting the coefficients, the fitted profile means should be inspected to determine whether Profiles 1, 2, and 3 still correspond to higher-, intermediate-, and lower-activity engagement patterns. Profile numbers are labels generated by the fitted model and do not carry fixed substantive meanings across separate model fits.

```{r one-step}
covariate_fit <- multilpa(
  course_engagement,
  vars,
  id = "student",
  n_profiles = 3,
  n_group_classes = 3,
  profile_covariates = "previous_grade",
  n_starts = 6,
  max_iter = 2000,
  tol = 1e-10,
  seed = 1
)

covariate_fit
```

This fit reports `converged FALSE`, and `multilpa()` warns that the best start had not converged when `max_iter` was reached, so the returned estimate is not a maximum. The profile means and membership coefficients of a one-step model are meaningful only when the fit has converged. They are therefore not interpreted here; a substantive analysis would first resolve the convergence problem, for example with more starts or iterations.

The `"coefficients"` table shows the form of the result: one intercept for each profile and student class, one `previous_grade` slope for each profile relative to the reference profile, and the intercepts of the student-class model.

```{r one-step-coefficients}
get_results(covariate_fit, "coefficients")
```

In a converged fit, a positive slope $\beta_k$ indicates higher odds of profile $k$ relative to profile 3 for each standard deviation of previous grade, conditional on student class. Because these slopes are conditional on student class and the profiles are estimated together with the covariate, they need not equal the three-step slopes even when both fits have converged.

## Holding profile definitions fixed across student classes

Staged estimation separates the estimation of the course-level engagement profiles from the estimation of the higher-level student-class structure. In the first stage, a common set of engagement profiles is estimated without allowing student classes to alter the profile measurement parameters. In the second stage, the composition of student classes is estimated while the profile means and variances from the first stage are held fixed.

The `fit_staged()` function implements this procedure by first fitting a model with a single student class and then estimating the requested three-class student-level structure. During the second stage, the model estimates the profile probabilities within each student class,

$$
\pi_{k\mid m}=P(C_{it}=k\mid G_i=m),
$$

together with the marginal student-class probabilities,

$$
\omega_m=P(G_i=m).
$$

The profile-specific means and variances remain fixed at their first-stage estimates. Staged estimation therefore isolates the estimation of the higher-level class structure from the estimation of the profile measurement model.

This separation has an important inferential consequence. Uncertainty in the first-stage measurement parameters is not propagated into the second-stage estimates. The resulting model is therefore conditional on the fixed first-stage profile solution and should not be interpreted as the joint maximum-likelihood solution that would be obtained by estimating all parameters simultaneously.

Profile numbering also requires careful attention when comparing staged and joint models. Latent profile and student-class labels are arbitrary numerical identifiers, and separate estimation procedures may assign different numbers to substantively similar groups. Comparisons should therefore be based on the estimated activity patterns and class compositions rather than on the numerical labels alone.

```{r staged}
staged <- fit_staged(
  course_engagement,
  vars,
  id = "student",
  n_profiles = 3,
  n_group_classes = 3,
  n_starts = 6,
  max_iter = 2000,
  tol = 1e-10,
  seed = 1
)

staged
```

In the staged solution, Profile 1 represents intermediate activity, Profile 2 represents lower activity, and Profile 3 represents higher activity. Their estimated `attendance` means are 0.197, -0.978, and 1.075, respectively. Although the numerical labels differ from those in the earlier joint model, the same broad engagement patterns remain recognizable. The estimated means also differ slightly because the profiles were obtained under a different estimation procedure.

The parameters estimated at each stage can be examined through the `"stages"` results table:

```{r stages}
get_results(staged, "stages")
```

The measurement stage estimates 32 parameters. The subsequent membership stage estimates 8 additional parameters governing the student-class composition. When the profile parameters retained from the measurement stage are also counted, the membership-stage model contains 38 `parameters_with_measurement`. This latter quantity is useful when making descriptive comparisons with a jointly estimated model because it reflects both the newly estimated higher-level parameters and the measurement parameters carried forward from the first stage.

The second-stage log likelihood is -8351.808, compared with -8348.429 for the joint model reported earlier. The lower log likelihood is consistent with the additional restriction imposed by staged estimation: the second stage cannot adjust the profile means and variances to improve the overall fit because those parameters have already been fixed. The two likelihoods should therefore be interpreted in light of the different estimation constraints.

Overall, staged estimation provides a way to preserve a previously established set of engagement profiles while estimating how those profiles are distributed across student classes. Its main advantage in this context is interpretive stability: the higher-level class structure is estimated without allowing the course-level profile definitions to change during the second stage.

## Interpretation and limitations

The membership regressions and grade contrasts indicate a consistent relationship between prior academic performance and subsequent engagement. Students in the lower-activity profile tend to have lower previous grades than students in the more active profiles. The distinction between the higher- and intermediate-activity profiles is less clear, however, because these two profiles show greater classification overlap and a correspondingly less certain difference in previous grades.

The different estimation strategies should not be interpreted as interchangeable versions of the same analysis. Three-step methods retain the profile definitions established in the measurement model and then examine their associations with external variables. In contrast, one-step estimation allows the covariate to contribute directly to estimation of the profiles and estimates the membership regression conditional on student class. The choice between these approaches therefore depends on the analytic objective: whether the priority is to preserve an existing profile solution or to estimate the measurement and covariate components jointly.

All of these results remain conditional on the adequacy of the underlying measurement model. As discussed in the [evaluation vignette](case-engagement-evaluation.html), assumptions such as diagonal within-profile covariance structures, the selected numbers of profiles and student classes, and the stability of the likelihood solution require careful evaluation. Classification-error corrections can account for uncertainty in assigning observations to latent profiles, but they cannot compensate for a poorly specified profile model.

Several inferential limitations also remain. Robust standard errors that cluster enrolments within students account for dependence in the observed data structure, but they do not convert an observational association into a causal effect. In addition, both staged and three-step procedures treat quantities estimated in earlier stages as fixed. Their reported standard errors and confidence intervals therefore do not propagate all of the uncertainty associated with estimating the initial measurement model or classification-error matrix.

