---
title: "Case study: Latent transition analysis of course engagement"
author: "Written by [Mohammed Saqr](https://saqr.me) and [Sonsoles López-Pernas](https://sonsoles.me)"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
---

```{r options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, fig.align = "center", dpi = 96)
options(digits = 4, width = 100)
```

Latent transition analysis (LTA) estimates the probability of moving from one
latent profile to another between consecutive occasions (Collins & Lanza,
2010). With group classes, each class has its own starting distribution and its
own transition matrix, so the classes separate students who differ in how their
engagement changes from students who differ only in where it starts.

## Data and model

`course_engagement` holds 1,422 course enrolments of 106 students, ordered
within each student by `sequence`. The five indicators are simulated on the
`log1p` scale of activity counts and standardized within course. The data are
synthetic and revisit the design of Saqr and López-Pernas (2021), who asked how
students' online engagement develops across a program. That study
identified course-level engagement states and longer-term trajectories with
latent class analysis and hidden Markov models. Here, three profiles and two
student classes are working choices for illustrating the transition model;
they do not establish the number of profiles or reproduce the published
estimates.

For student $i$ at course position $t$, $C_{it}$ is the latent profile and
$G_i$ the student's class. The model estimates

- the initial probabilities $\delta_{k\mid m}=P(C_{i1}=k\mid G_i=m)$,
- the transition probabilities
  $a_{kl\mid m}=P(C_{i,t+1}=l\mid C_{it}=k,G_i=m)$, which sum to one across
  destinations $l$,
- and a Gaussian mean and variance for each profile and indicator.

The transitions are first order and homogeneous: the next profile depends only
on the current profile and the student's class, and the probabilities are the
same at every position. Profile means and variances are shared across positions
and classes, so a change of profile is a change of state. Indicators are
independent within a profile under the default diagonal covariance. The fit can
fail to reach the global maximum from a single start, and a transition that is
rare within a class is estimated from little information; the expected count
of each transition shows how much.

```{r fit}
library(latents)
vars <- c("browse", "lectures", "forum_read", "forum_post", "attendance")

moves <- lta(course_engagement, vars, "student", n_profiles = 3,
             n_group_classes = 2, time = "sequence", n_starts = 10, seed = 1)
get_results(moves, "starts")
```

All ten starts converge to the same log likelihood, so the solution is
replicated within this seed's starts.

## The profiles

`plot()` draws the profile means across the indicators, with point size showing
profile prevalence.

```{r profiles}
plot(moves)
```

```{r profile-means}
get_results(moves)
```

Profile 1 is the lowest on every indicator, profile 2 the highest, and profile
3 lies between them; their attendance means are -0.929, 1.122 and 0.252. The
profiles are therefore lower, higher and intermediate activity. The heatmap
shows the same means as standard deviations from each indicator's grand mean.

```{r heatmap}
plot(moves, what = "heatmap")
```

The contrast between profiles is present on all five indicators, so the
profiles differ in overall activity more than in its composition.

## Where students start

In the `"initial"` table, `probability` is $\delta_{k\mid m}$, the chance of
starting in each profile. `prevalence` is the expected share of all of a
class's enrolments in that profile, across the whole history, and
`group_class_probability` is the share of students in the class.

```{r initial}
get_results(moves, "initial")
```

Student class 1, with 30.1% of students, starts in the lower-activity profile
with probability 0.762 and in the higher-activity profile with probability
0.238; its probability of starting in the intermediate profile is effectively
zero. Class 2, with 69.9% of students, starts across all three profiles, with
probabilities 0.204, 0.445 and 0.351. Starting and overall shares differ: class
1 starts in the lower profile with probability 0.762, but 0.819 of its
enrolments are expected there over the whole history, while class 2 starts in
the higher profile with probability 0.445 and spends 0.334 of its enrolments
there.

## How engagement changes

The `"transitions"` table gives $a_{kl\mid m}$ for every class and pair of
profiles. `stable` marks staying in the same profile. `expected_count` is the
posterior expected number of such transitions in the data, so it can be
fractional, and it shows how much information supports each estimate.
`estimated = FALSE` would mark a row the data cannot inform, because the
profile is never occupied before a student's last course; such a row holds
uniform placeholder values. Every row is estimated in this fit.

```{r transitions}
get_results(moves, "transitions")
```

The transition plot draws each class's matrix as a heatmap, with the origin
profile in rows and the destination in columns.

```{r transition-plot}
plot(moves, what = "transitions")
```

In class 1, lower activity persists with probability 0.938, estimated from 303
expected transitions. Each of its other transitions rests on fewer than 20
expected transitions, because class-1 students rarely leave the lower profile.
Class 2 moves between profiles. Its students remain in the lower profile with
probability 0.614, move from higher to intermediate activity with probability
0.494, and from intermediate to higher with probability 0.362. Class 1 is
therefore a persistently low-engagement class, and class 2 a class whose
engagement changes from course to course.

## Sparse and boundary estimates

Some estimates rest on very little information. Class 1's probability of
starting in the intermediate profile is at its lower bound, because no class-1
student is expected to start there. Class 1's transitions out of the higher
and intermediate profiles rest on between 4.8 and 19.0 expected transitions
each, because its students rarely leave the lower profile. A probability near
zero shows that the data carry little information about that movement, not
that the movement is impossible. Estimates like these can also mean that the
chosen number of classes divides the available transitions too finely.

## Assigned sequences

The sequence plot shows each student's modal profile at each course position,
one row per student. Modal assignment discards classification uncertainty.

```{r sequences, fig.height = 6}
plot(moves, what = "sequences")
```

```{r sequence-summary}
get_results(moves, "sequence_summary")
```

The summary gives the number of students and courses in each assigned class.
Sequence lengths differ between students, so the rows of the plot end at
different positions.

## Classification uncertainty

```{r entropy-table}
get_results(moves, "entropy")
```

Relative entropy is 0.802 for enrolments and 0.777 for students. The entropy
plot shows, within each profile, how the per-case entropy contributions are
distributed, and the posterior plot shows the posterior probability of each
case's assigned profile.

```{r entropy-plot}
plot(moves, what = "entropy")
```

```{r posteriors-plot}
plot(moves, what = "posteriors")
```

Both levels retain uncertainty, so an analysis that replaces the posteriors by
modal assignments treats some enrolments and students as more certain than
the model estimates them to be.

## Transition networks

`get_tna()` returns the marginal transition matrix as a model of the tna
package (Tikka et al., 2025). It pools the posterior expected transition counts
across classes and normalizes them, so it uses the estimated transitions, not
transitions counted from modal assignments. `get_group_tna()` returns one
network per class.

```{r network, message = FALSE}
network <- get_tna(moves)
plot(network)
```

```{r group-network, message = FALSE, fig.height = 4}
per_class <- get_group_tna(moves)
plot(per_class)
```

The class networks show the two patterns of the transition table: a dominant
self-loop on the lower profile in class 1, and movement among all three
profiles in class 2.

## Limitations

The model assumes constant transition probabilities across course positions,
so change specific to one stage of the program is averaged over positions.
Transitions are between successive observed courses, which can differ in
duration, content and assessment. Class differences describe patterns in
engagement histories and do not identify their causes.
Standard errors, likelihood-ratio tests and class enumeration are not available
for transition models, so the number of profiles and classes here is a working
choice. The diagonal covariance ignores the association between `attendance`
and the click measures, which the evaluation vignette shows is present in
these data; unmodelled dependence can be absorbed by extra profiles and change
the apparent transitions. Transitions supported by few expected counts, such as those out of
the higher and intermediate profiles in class 1, should not be interpreted.

## References

Collins, L. M., & Lanza, S. T. (2010). *Latent class and latent transition
analysis*. Wiley.

Saqr, M., & López-Pernas, S. (2021). The longitudinal trajectories of online
engagement over a full program. *Computers & Education, 175*, 104325.
<https://doi.org/10.1016/j.compedu.2021.104325>

Tikka, S., López-Pernas, S., & Saqr, M. (2025). tna: An R package for
transition network analysis. *Applied Psychological Measurement*.
<https://doi.org/10.1177/01466216251348840>
